---
title: "Combining mizerReef with mizerMR: multiple background resources"
output:
  html_document:
    toc: yes
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Combining mizerReef with mizerMR: multiple background resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 300,
  message = FALSE,
  warning = FALSE
)
```

## Overview

A standard mizer model has a single, size-structured background resource
spectrum that represents everything the fish community feeds on but which is
not itself resolved as a dynamic species. mizerReef keeps that single resource
and adds two *unstructured* benthic components on top of it — algae and
detritus — so that herbivores and detritivores have somewhere to feed.

Sometimes one lumped planktonic resource is too coarse. Different parts of the
community may rely on ecologically distinct pools of prey that occupy different
size ranges and replenish at different rates. The
[mizerMR](https://sizespectrum.org/mizerMR/) extension package solves exactly
this problem: it replaces the single background resource with any number of
independent, size-structured resource spectra, and lets each predator species
have its own preference for each resource.

Because both packages are written as mizer *extensions*, they compose. This
vignette shows how to:

1. [Load both extensions](#loading-both-extensions) and understand what the
   combined model class means.
2. [Split the reef's background resource](#splitting-the-background-resource)
   into several size-structured resources.
3. [Give species different resource preferences](#giving-species-different-preferences)
   with an interaction matrix.
4. [Retune the steady state](#retuning-the-steady-state) and
   [explore the results](#exploring-the-results).

We assume you have read `vignette("mizerReef")` and are comfortable with the
basic reef workflow. Familiarity with
`vignette("using-extension-packages", package = "mizer")` is helpful but not
required.

## Loading both extensions

Install mizerMR alongside mizerReef if you do not have it yet:

```{r eval = FALSE}
# install.packages("remotes")
remotes::install_github("sizespectrum/mizerMR")
```

Then load both packages. Each mizer extension announces itself to mizer when it
is loaded, building up an *extension chain*. The package you load **last** sits
at the front of the chain.

```{r}
library(mizerReef)
library(mizerMR)
```

```{r}
getRegisteredExtensions()
```

Both extensions are now active. mizerReef ships an example model for a generic
three-group Caribbean reef, `caribbean_3_model`, with functional groups for
predators, herbivores and invertebrates:

```{r}
params <- caribbean_3_model
class(params)
```

The object carries the `mizerReef` class, so all of mizer's generics
(`project()`, `plotSpectra()`, `getBiomass()`, …) dispatch to mizerReef's
methods. It has a single background resource plus the reef's algae and detritus
components:

```{r}
names(params@other_params)
```

## Splitting the background resource

In the example model, predators feed mostly on the size-structured background
resource (think planktonic and small pelagic prey), while herbivores and
invertebrates take a smaller share of it in addition to algae and detritus:

```{r}
sp <- species_params(params)
data.frame(species = sp$species,
           interaction_resource = sp$interaction_resource,
           interaction_algae = sp$interaction_algae,
           interaction_detritus = sp$interaction_detritus)
```

Suppose we want to distinguish two planktonic pools: a **small plankton**
resource that dominates the smallest sizes and a **large plankton** resource
that carries the larger prey. mizerMR describes each resource with one row of a
`resource_params` data frame. The important columns are the coefficient
(`kappa`) and exponent (`lambda`) of the carrying-capacity power law, the size
range (`w_min`, `w_max`), and the replenishment rate (`r_pp`, `n`). Any column
you leave out is filled with a sensible default (see
`mizerMR::validResourceParams()`).

Here we give both resources the same power law that the original single
resource used, but restrict each to half of the size axis, so that together
they reproduce a continuous background spectrum:

```{r}
wf <- w_full(params)
resource_params <- data.frame(
    resource = c("small plankton", "large plankton"),
    kappa    = 1e11,
    lambda   = 2.05,
    w_min    = c(min(wf), 1),
    w_max    = c(1, max(params@w)),
    r_pp     = 4,
    n        = 2 / 3
)
resource_params
```

`setMultipleResources()` installs these resources into the model. It silences
the built-in single resource and takes over the resource dynamics, while
leaving mizerReef's algae, detritus and refuge machinery untouched:

```{r}
params <- setMultipleResources(params, resource_params = resource_params)
```

The returned object now belongs to a class that chains **both** extensions:

```{r}
class(params)
is(params, "mizerReef")
is(params, "mizerMR")
```

`class()` reports the most-derived class, `mizerMR`, but the object still
inherits from `mizerReef`, so reef generics and multiple-resource generics both
dispatch correctly. The multiple resources live in a new `"MR"` component:

```{r}
names(params@other_params)
dimnames(initialNResource(params))$resource
```

## Giving species different preferences

So far both resources are eaten in the same proportions as the original single
resource. The real payoff of multiple resources is that each species can prefer
different ones. This is controlled by a *resource interaction matrix* with one
row per species and one column per resource.

Let us say predators chase the larger planktonic prey, while herbivores and
invertebrates graze mostly on the small pool:

```{r}
resource_interaction <- matrix(
    c(0.2, 1.0,    # predators:  small, large
      0.2, 0.05,   # herbivores
      0.3, 0.05),  # inverts
    nrow = 3, byrow = TRUE,
    dimnames = list(sp = sp$species,
                    resource = resource_params$resource)
)
resource_interaction
```

Note the names on the two dimensions (`sp` and `resource`): mizerMR checks that
the matrix matches the species and resources of the model, so it is worth
labelling both axes. We pass the matrix straight to `setMultipleResources()`:

```{r}
params <- setMultipleResources(
    params,
    resource_params      = resource_params,
    resource_interaction = resource_interaction
)
resource_interaction(params)
```

## Retuning the steady state

Changing the resources moves the model away from its previous steady state, so
we retune it. mizerReef's `reefSteady()` works exactly as it does for an
ordinary reef model — it is unaware that the background is now several
resources, because the multiple-resource dynamics are handled transparently by
mizerMR's methods further down the chain. A couple of iterations are usually
enough:

```{r}
params <- reefSteady(params)
params <- reefSteady(params)
```

## Exploring the results

Every reef and multiple-resource plotting function now works on the combined
model. `plotSpectra()` shows the fish community together with the two
background resources, which tile the size axis:

```{r}
plotSpectra(params, power = 2)
```

The diet composition resolves the two resources separately, so you can see the
ontogenetic shift between them. Predators lean on large plankton, while the
smaller-bodied groups take the small-plankton pool (alongside algae and
detritus):

```{r}
plotDiet(params)
```

Biomasses of the fish groups and of the unstructured algae and detritus
components are reported by `getBiomass()` as before. The size-structured
multiple resources are not folded into the species biomass table; they are
reported through the resource accessors instead:

```{r}
sim <- project(params, t_max = 20)
getBiomass(sim)[c(1, 21), ]
```

```{r}
plotBiomass(sim)
```

## How the chaining works

Under the hood, both packages register themselves as *dispatching extensions*.
mizer builds an S4 class hierarchy in which the combined object extends
`mizerMR`, which extends `mizerReef`, which extends `MizerParams`:

```{r}
extends("mizerMR")
```

When you call a generic such as `getEncounter()` or `getBiomass()`, R walks this
hierarchy from the outermost class inward. Each extension's method does its own
work and then calls `NextMethod()` to let the next extension in the chain
contribute. That is why the reef's refuge-modified encounter and the multiple
resources' encounter both end up in the final rate, and why a single
`plotSpectra()` call draws the reef community and every background resource
together. Neither package needs to know about the other; they only need to be
polite about calling `NextMethod()`.

For more on the extension chain — including how the load order is recorded in
the object and restored when you save and reload it — see
`vignette("using-extension-packages", package = "mizer")`.
