---
title: "Simple 3-Group Example"
subtitle: "Caribbean 3-Group Steady State Model Description"
author: ""
bibliography: references.bib
csl: ecology.csl
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simple 3-Group Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(dpi=400)
```

```{r, include = FALSE}
library(ggplot2)
library(dplyr)
library(knitr)
library(mizer)
library(mizerExperimental)
library(mizerReef)

data(caribbean_3_model)
params <- caribbean_3_model
gear_params(params)["inverts, knife_edge_gear", "catchability"] <- 0
params@other_params$include_sen_mort <- TRUE
```

\newpage

# Model Description

::: {.content-visible when-format="html"}
[Skip to Resources](#resources)
:::

`caribbean_3_model` reproduces the three-group trait-based reef model from
@Rogers_2018 -- predators, herbivores, and a general group of benthic
invertebrates -- inside mizerReef. It is deliberately coarse: useful when a
species-group resolution is enough, or as a minimal starting point for
setting up a new reef model, before working up to something like the
[10-group Karpata Reef model](karpata_model-description.html), which
parameterises the same three broad functional groups against finer field
data. This page documents this model's specific parameters and how they were
tuned; for the general mizerReef formulation these parameters plug into, see
`vignette("model-description")`.

Biomass estimates were based on data collected from relatively un-fished
reefs within a marine reserve in Bonaire. This data was part of the FORCE
dataset that categorised fish community structure and habitat complexity
across Caribbean coral reefs [@Dryden2016; @Rogers_2014; @Newman_2015;
@Williams_2015; @Williams_2016]. A marine-reserve site was chosen
deliberately, for this model and for the 10-group Karpata Reef model alike,
so that observed biomasses would not need to be corrected for fishing
pressure before being used to calibrate the models. Visual surveys recorded
fish up to 5 cm in length. The cutoff weight for observed biomass estimates
was calculated using published length-weight relationships [@FishBase]. The
cutoff size is given in the table below along with the observed biomass per
square meter in grams.

```{r, echo = FALSE}
params@species_params %>%
    filter(!is.na(biomass_observed)) %>%
    select(species,biomass_observed, biomass_cutoff) %>%
    kable(row.names = FALSE,
          col.names = c("Species Group",
                        "Observed Biomass [g/m^2]", "Cutoff Size [g]"))
```

## Interaction matrix

The $\theta_{ij}$ matrix sets the interaction strength between predator group
$i$ and prey group $j$, with entries between 0 (groups cannot interact) and 1.

```{r, echo = FALSE, fig.height = 3.5, fig.width = 5, fig.cap = "Interaction strength between each predator group (rows) and prey group (columns). Predators consume all three groups, at different strengths; herbivores and invertebrates do not prey on fish."}
ggplot(melt(params@interaction),
       aes(x = prey, y = reorder(predator, desc(predator)), fill = value)) +
    geom_tile(colour = "black", linewidth = 0.25) +
    scale_fill_gradient(low = "white", high = "blue", limit = c(0, 1),
                        guide = "legend") +
    labs(y = "Predator", x = "Prey", fill = "Interaction") +
    scale_x_discrete(labels = c("Predators","Herbivores", "Inverts")) +
    scale_y_discrete(labels = c("Inverts","Herbivores", "Predators")) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Predation kernel

The parameters for the predation kernels were based on the global average
predator-prey mass ratio for marine organisms and home range size estimates
from @Nash_2014. All estimates fall within observed ranges. All groups use a
lognormal predation kernel with the same parameters as in @Rogers_2014.

```{r, echo = FALSE, results='asis'}
params@species_params %>%
    filter(pred_kernel_type == "lognormal") %>%
    select(beta, sigma, gamma) %>%
    kable()
```

## Vulnerability to predation

```{r, include = FALSE}
refuge_params <- params@other_params$refuge_params
```

This model uses the `r print(params@other_params$refuge_params$method)`
refuge method (see `vignette("model-description")`'s
[refuge profiles](model-description.html#refuge-profiles) section for the
general formulation and @Rogers_2018 for its derivation). All fish are
assumed to utilize refuge ($\tau = 1$). Fish smaller than
`r paste(refuge_params$w_settle,"g")` are assumed to be larval reef fish that
have not yet settled to the reef. A maximum proportion of
`r paste(100*refuge_params$max_protect,"%")` of fish are protected at any
given time.

<details>
<summary>Refuge length bins and densities used in the steady state (same Bonaire FORCE data as the 10-group model)</summary>

The following table gives the fish length bins and refuge densities that
define the refuge profile used in the steady state. Length bins and refuge
densities were garnered from the FORCE data in the same location as biomass
estimates -- the same [`karpata_refuge`](../reference/karpata_refuge.html)
profile used by the 10-group model.

```{r, echo = FALSE}
params@other_params$refuge_params$method_params %>%
    mutate(refuge_density = round(refuge_density,3)) %>%
    kable(row.names = FALSE,
          col.names = c("Start of Bin (cm)",
                        "End of Bin (cm)",
                        "Refuge Density (no./m^2)"))
```

</details>

<details>
<summary>How each species group interacts with predation refuge</summary>

Each species group can utilise benthic structures differently depending on
their specific traits.

```{r, echo = FALSE, results='asis'}
params@species_params %>%
    select(refuge_user, blocked_pred) %>%
    mutate(refuge_user  = case_when(refuge_user == TRUE ~ 'Yes',
                                    refuge_user == FALSE ~ '×')) %>%
    mutate(blocked_pred = case_when(blocked_pred == TRUE ~ '×',
                                    blocked_pred == FALSE ~ 'Yes')) %>%
    kable(col.names = c("Group", "Uses refuge?", "Accesses prey in refuge?"))
```

</details>

The figure below shows the density-dependent refuge profile produced by the
competitive method at steady state. Invertebrates are not shown as they do
not use refuge (`refuge_user = FALSE`, see the table above).

```{r, echo = FALSE, fig.height = 3, fig.width = 7, fig.cap = "Proportion of each refuge-using group protected from predation, by body length, at steady state."}
plotRefugeProfile(params)
```

\newpage

## Consumption

Consumption of the detrital and algal resources is subject to a Holling type
II functional response (see `vignette("model-description")`'s
[Consumption and satiation](model-description.html#consumption-and-satiation)
section for the general formula). The parameter $h$ is the max consumption
rate for a consumer of size 1 gram, chosen so that consumers subject to
satiation are neither too starved nor totally satiated; $\alpha_i$ is the
proportion of consumed biomass retained, with $1-\alpha_i$ expelled as
faeces, which contribute to detritus.

No maximum consumption rate is imposed for predators. Herbivores are subject
to satiation here, unlike the package default for herbivorous groups:
recalibrating this model against the corrected senescence-mortality formula
showed herbivore biomass has no density-dependent brake at all without some
cap on individual intake once mortality is realistically low. The realised
feeding level for herbivores is consistently close to 1, consistent with
Caribbean herbivores' guts being observed to be full essentially continuously
[@Ferreira1998; @Kopp2010] -- the citations behind the package's
herbivore-satiation default describe grazing *pressure on the shared algae
resource* not self-regulating when food is abundant, which is unaffected by
this change (see [the main model description's Algae
section](model-description.html#algae)).

```{r, echo = FALSE}
params@species_params %>%
    filter(satiation == TRUE) %>%
    select(h, alpha, n) %>%
    kable()
```

### Metabolic losses

Standard metabolism occurs at rate $k_{s.i}$; losses due to activity and
movement occur at rate $k_i$ (see `vignette("model-description")` for the
general growth/reproduction formulation).

```{r, results='asis', echo = FALSE}
params@species_params %>%
    select(ks, p, k) %>%
    kable()
```

### Maturation and growth

Maturation length and age data were based on the most commonly observed
species from each functional group in the FORCE data set [@Dryden2016;
@Rogers_2014; @Newman_2015; @Williams_2015; @Williams_2016]. For predators,
this was the graysby grouper, *Cephalopholis cruentata*. The most commonly
observed herbivore was the stoplight parrotfish, *Sparisoma viride*. Relevant
maturity parameters were pulled from FishBase [@FishBase].

The herbivore `age_mat` value is set to 1.6 years, the age at median sexual
maturity (AM50) reported for *S. viride* by @RiveraHernandez2025 -- a
2013-2023 otolith/gonad-histology study of 1801 U.S. Caribbean stoplight
parrotfish. An earlier version of this model used `age_mat = 4`, which
appears to have conflated AM50 with that same study's age at median sexual
transition (AT50 = 4.5 years, the age at which female stoplight parrotfish --
a protogynous hermaphrodite species -- become male), a distinct milestone
from first reproductive maturity.

```{r, echo = FALSE}
params@species_params %>%
    select(w_mat, w_max, age_mat) %>%
    kable()
```

The values for the growth parameters below were chosen so that the resulting
growth curves would be close to von Bertalanffy growth curves; $a$ and $b$
are the allometric weight-length parameters $w = a l^b$ ($w$ in grams, $l$ in
centimetres), taken from the literature.

```{r, echo = FALSE}
params@species_params %>%
    select(k_vb, w_max, a, b) %>%
    kable()
```

## Mortality {#mortality}

See `vignette("model-description")`'s [Mortality](model-description.html#mortality)
section for the general formulation. This model's residual-natural- and
senescence-mortality parameters:

```{r, results = 'asis', echo = FALSE}
ext_mort <- params@other_params[['ext_mort_params']]
```

We use a residual natural mortality rate of $\mu_{nat} =$
`r ext_mort$nat_mort` per year at size 1 gram, and senescence mortality
parameters $k_{sen} =$ `r ext_mort$sen_prop` (`sen_prop`) and $p_{sen} =$
`r ext_mort$sen_curve` (`sen_curve`, the exponent governing how steeply
mortality climbs as individuals approach their maximum size), both based on
estimates from @Hatcher_1988.

### Fishing mortality

Fishing parameters for mizerReef models can be set up with `setFishing()`.
The vignettes were run with the following fishing parameters:

```{r, echo = FALSE}
params@gear_params %>%
    filter(catchability > 0) %>%
    select(species, knife_edge_size, catchability) %>%
    kable(row.names = FALSE,
          col.names = c("Group", "Minimum fishing size [g]",
                        "Catchability [1/year]"))
```

## Reproduction {#reproduction}

The reproduction parameters $\epsilon_i$ and $R_{max.i}$ are not directly
observable. Their values were instead chosen so as to produce steady-state
abundances of the groups that are in line with observations and to give
reasonable values for the reproduction level -- the ratio between the actual
reproduction rate $R_i$ and the maximal possible reproduction rate
$R_{\max.i}$.

```{r, echo = FALSE}
df <- params@species_params
df %>%
    select(w_min, erepro, R_max) %>%
    kable()
```

# Resources {#resources}

::: {.content-visible when-format="html"}
[Skip to Tuning the Steady State](#tuning-the-steady-state)
:::

The fish spectrum is fed by three background resources: plankton, algae, and
detritus. Small individuals of all species feed on plankton, while only
herbivorous groups and invertebrates feed on algae or detritus. See
`vignette("model-description")`'s [unstructured resource
dynamics](model-description.html#unstructured-resource-dynamics) section for
the general formulation and its relationship to mizerShelf.

### Plankton {#plankton}
```{r, echo = FALSE}
wcutoff <- params@resource_params$w_pp_cutoff
lambda <- params@resource_params$lambda
kappa <- params@resource_params$kappa
```

The plankton spectrum ranges from $w_0=`r signif(min(params@w_full), digits = 1)`$
to $w_{cutoff}=`r wcutoff`$ grams. The steady state abundance of plankton at
size 1 gram is $\kappa=`r signif(kappa, digits = 3)` (g/m^{2})$, with slope
$\lambda=`r signif(lambda, digits = 3)`$.

### Algae {#algae}
The algal resource is described only by its total biomass $B_A$, and feeding
on it is not size-based. In the steady state the total algal biomass per
square meter is $B_A = `r signif(algae_biomass(params), digits = 3)`$ grams.

```{r, echo = FALSE}
params@species_params %>%
    select(rho_algae, interaction_algae) %>%
    rename(rho = rho_algae) %>%
    filter(rho > 0) %>%
    kable()
```

```{r, echo = FALSE, include=FALSE}
pA <- params@other_params$algae$growth
```

Algal *production* -- a fixed, literature-informed constant, `algae_growth_initial`
in `setAlgaeParams()` -- is `r signif(pA,digits = 4)` grams per square meter
per year at steady state. Note this value is not retuned to match
consumption the way detritus production is (see below); see
`vignette("model-description")`'s Algae section for why.

### Detritus
Detritus is consumed by herbivores and benthic invertebrates, also not
size-based. In the steady state the total detrital biomass per square meter
is $B_D = `r signif(detritus_biomass(params), digits = 4)`$ grams.

```{r, echo = FALSE}
params@species_params %>%
    select(rho_detritus,interaction_detritus) %>%
    rename(rho = rho_detritus) %>%
    filter(rho > 0) %>%
    kable()
```

```{r, echo = FALSE}
#| include: false
e_d <- params@other_params$detritus$ext_decomp
s_d <- params@other_params$detritus$sen_decomp
r <- getDetritusProduction(params)
pf <- r[1]
pd <- r[2]
dext <- params@other_params$detritus$external
```
Detritus production comes from three sources: defecation, decomposing dead
organisms, and external input. At steady state, decomposing dead organisms
contribute `r paste(round(pd,2))` g/year and defecation contributes
`r paste(round(pf,2))` g/year -- `r paste(100*e_d,"%")` of external mortality
and `r paste(100*s_d,"%")` of senescence mortality is assumed to decompose to
detritus (estimates from @Hatcher_1988). The remaining external-input term is
solved for so that, unlike algae, detritus *production* matches current
consumption: at steady state it is `r paste(round(dext,2))` g/year. Where
this is negative, faeces and mortality alone already produce more detritus
than is consumed, and the excess is assumed to be washed away.

# Tuning the Steady State

The following R script was used to tune the steady state parameters for this
model (see `inst/scripts/Caribbean_3_model-calibration.R` in the package
source for the exact, currently-runnable version, and
`vignette("steady-state-recipe")` for the general recipe this follows,
including the final algae/detritus absolute-scale rescaling step).

<details>
<summary>Show the full steady-state calibration script</summary>

```{r, echo = TRUE, eval = FALSE}
# Setting up a generic Caribbean coral reef model with multiple resources
# Three groups: Predators, Herbivores, Invertebrates
# Model steady state calibration

## Setup - load packages --------------------------------------------------
library(mizer)
library(mizerExperimental)
library(mizerReef)
library(here)

## Load parameters ----------------------------------------------------------
caribbean_3_species <- read.csv(here("inst/data-csv/caribbean_3_species.csv"))
caribbean_3_interaction <- read.csv(here("inst/data-csv/caribbean_3_interaction.csv"),
                                    row.names = 1)
# Refuge densities from Karpata reef in Bonaire, FORCE dataset
karpata_refuge <- read.csv(here("inst/data-csv/karpata_refuge.csv"))
tuning_profile <- read.csv(here("inst/data-csv/tuning_profile.csv"))

# With these parameters, herbivores consume plankton at small sizes and
#   transition fully to algae by maturity
# With these parameters, invertebrates consume plankton and detritus,
#   with the proportion of detritus increasing with size

## Set model, with info_level = 1 to keep only the reports that say
## something went differently from what was asked, not every default fill ---
params <- newReefParams(
    species_params = caribbean_3_species,
    interaction = caribbean_3_interaction,
    method = "binned",
    method_params = tuning_profile,
    info_level = 1
)

## Project to first steady state ----------------------------------------------
params <- reefSteady(params)

## Calibrate biomasses and growth ----------------------------------------------

# Match observed species group biomasses
params <- calibrateReefBiomass(params)
params <- matchBiomasses(params)
params <- reefSteady(params)

# Match observed growth rates
params <- matchReefGrowth(params)
params <- reefSteady(params)

# Iterate to refine biomass - repeat many times; piping keeps it readable.
params <- params |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |>
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |>
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |>
    reefSteady()
# (repeated several more times in the full calibration script until
#  plotBiomassVsSpecies(params) matches observations closely)

plotBiomassVsSpecies(params) # spot on
plotTotalAbundance(params)
plotTotalBiomass(params)

## Now switch to competitive method ---------------------------------------------
params <- newRefuge(params,
    new_method = "competitive",
    new_method_params = karpata_refuge
)

# Match biomasses again with the new refuge profile
params <- params |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |>
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |>
    reefSteady()
# (repeated several more times in the full calibration script)

# Make sure new refuge is in place
plotVulnerable(params)
plotRefugeProfile(params)
plotBiomassVsSpecies(params) # spot on

## Check resulting spectra and tune resources ------------------------------------

# Spectra should be reasonably straight to match predictions of Sheldon's
# spectrum but also have nonlinearities at refuge sizes
plotSpectra(params, total = TRUE, biomass = TRUE)
plotSpectra(params, total = TRUE, biomass = TRUE, per_log_size = TRUE)
plotFeedingLevel(params, species = "inverts")

# Tune reproduction -----------------------------------------------------------
# We do not have yield or catch data - can't tune size distribution
params <- setBevertonHolt(params, erepro = 0.0134)
params <- reefSteady(params)

# Increase reproduction level to reduce excessive density dependence
rep_level <- c(0.5, 0.5, getReproductionLevel(params)["inverts"])
names(rep_level) <- c("predators", "herbivores", "inverts")
params <- setBevertonHolt(params, reproduction_level = rep_level)

# Iterate to get back to steady state
params <- params |> reefSteady() |> reefSteady() |> reefSteady()

# Rescale algae/detritus to a realistic absolute scale, now that diet,
# biomass and growth are tuned -- see vignette("steady-state-recipe")'s
# final step for the literature targets and why this belongs last.
params <- rescale_algae(params, target_biomass / algae_biomass(params))
detritus_lifetime(params) <- target_lifetime

# Final diagnostic plots
plotBiomassVsSpecies(params)
plotRefugeProfile(params)
plotSpectra(params, biomass = TRUE, total = TRUE)
plotDiet(params)
plotGrowthCurves(params)
plotPredMort(params)

# Save!
caribbean_3_model <- reefSteady(params)
save(caribbean_3_model, file = "data/caribbean_3_model.rda")
```

</details>
