---
title: "How mizerReef extends mizer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How mizerReef extends mizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(mizer)
library(mizerReef)
params <- caribbean_3_model
```

# Overview

mizerReef is an extension package for mizer that adds three ecological
processes to a standard multispecies size-spectrum model, all motivated by
structurally complex habitats such as coral reefs:

- **predation refuge** — structural complexity lets some prey hide from some
  predators, reducing the encounter and mortality rates that would otherwise
  apply,
- **algae and detritus** — two unstructured (not size-structured) benthic
  resource pools that herbivorous and detritivorous species feed on,
- **senescence mortality** — an extra size-dependent mortality term
  representing non-predation death close to a species' maximum size.

The package uses the mechanisms described in
`vignette("creating-extension-packages", package = "mizer")`:

| Extension mechanism | Used for |
|---|---|
| `.onLoad` + `registerExtension()` | Register the package with mizer so params objects know which extensions they need |
| S4 marker classes + S3 dispatch | Define `mizerReef`/`mizerReefSim` so reef-specific methods run automatically |
| `project*` rate methods | Override Encounter, FeedingLevel, PredMort and Mort to add refuge, satiation, and senescence effects |
| `setComponent()` | Add algae and detritus as scalar dynamical components |
| `utils::upgrade()` | Migrate objects saved with older mizerReef versions |

All of these are wired together inside `newReefParams()`, the single entry
point for building a reef model.

# The `mizerReef` marker class

mizerReef defines two S4 marker subclasses:

```r
setClass("mizerReef",    contains = "MizerParams")
setClass("mizerReefSim", contains = "MizerSim")
```

These classes carry no extra slots — all reef-specific state (refuge, algae
and detritus parameters) lives in `other_params(params)`. The classes exist
purely to trigger S3 dispatch, exactly as described for marker classes in
`vignette("creating-extension-packages", package = "mizer")`.

## Registration via `.onLoad`

```r
.onLoad <- function(libname, pkgname) {
    mizer::registerExtension(pkgname, requirement = "sizespectrum/mizerReef")
}
```

This registers mizerReef with mizer's extension chain as soon as the package
is loaded. `newReefParams()` then records the chain and coerces the result to
the marker class at the very end of construction:

```r
params@extensions <- mizer::getRegisteredExtensions()
params <- mizer::coerceToExtensionClass(params)
```

Recording the chain in `params@extensions` is what lets `project()` produce a
`mizerReefSim` automatically (via `MizerSim()`'s own call to
`coerceToExtensionClass()`), and what lets `readParams()` restore the correct
class — and warn about missing or outdated extensions — when a saved object
is reloaded in a later session.

### A known gap: bundled example models and multiple extensions

mizerReef ships two pre-built example models as package data
(`caribbean_10_model`, `caribbean_3_model`). `.onLoad` also contains a
`makeActiveBinding()` call intended to re-coerce these objects to the
correct class if a *second* mizer extension is loaded alongside mizerReef,
since a bundled `.rda` is fixed at save time and can't know in advance
what else will be loaded in the same session. In the current mizer/
mizerReef versions this mechanism does not fire — `exists()` on the
bundled object inside `.onLoad` is always `FALSE` at that point, so the
active binding is never installed. This is not mizerReef-specific: the
identical dead-code pattern exists in mizerShelf's own current source for
its bundled `NWMed_params` object.

For everyday single-extension use this is harmless, since both bundled
objects already carry the correct class from when they were saved. It
would only matter if you loaded mizerReef together with a second
dispatching extension on one of these two specific objects — for example,
the mizerMR combination shown in `vignette("using-multiple-resources")`.
That combination does work (mizer's extension chain, a separate mechanism,
correctly promotes the object's class when both extensions are
registered), but the underlying `.onLoad` dead-code gap described above is
still unfixed upstream; see `inst/to-do-list.txt` for the current state.

## Methods dispatched on `mizerReef`

| Method | What it adds |
|---|---|
| `projectEncounter.mizerReef()` | Recomputes encounter for refuge-blocked predators using vulnerability-reduced prey abundance |
| `projectFeedingLevel.mizerReef()` | Gives predators without a satiation response unlimited intake capacity |
| `projectPredMort.mizerReef()` | Discounts predation mortality from refuge-blocked predators by prey vulnerability |
| `projectMort.mizerReef()` | Adds senescence mortality on top of the standard mortality |
| `getBiomass.mizerReefSim()` | Adds algae and detritus biomass to the species biomasses |
| `removeSpecies.mizerReef()` | Updates the algae/detritus encounter-rate matrices `rho` |
| `upgrade.mizerReef()` | Migrates objects created with mizerReef < 2.1.0 to the current layout |

Every one of these methods calls `NextMethod()` at least once, so multiple
extension packages can modify the same rate or generic without silently
overwriting each other's contribution — *provided every extension involved
follows the same convention*. This is not automatic just because a package is
registered as a mizer extension: the chaining mechanism only walks the class
hierarchy and calls whichever method comes next, it does not merge two
extensions' contributions on its own. An extension that overrides a rate
without calling `NextMethod()` (rather than dispatching through the chain)
will simply overwrite whichever extension was loaded before it, with no
combination and no warning. See the caveat in
`vignette("using-multiple-resources")` for a concrete example (mizerReef and
therMizer both modify the encounter rate, but do not currently compose).
The next two sections look at mizerReef's own rate methods in detail, because
they illustrate both the easy case (a purely additive modification) and a
harder case that needed more thought.

# Predation refuge: a multiplicative rate modification

Three species parameters control how refuge affects a species:

- `refuge_user` — does this species hide in refuge at all?
- `blocked_pred` — is this species' *foraging* hindered by refuge (i.e. is it
  a predator that struggles to reach prey hiding in structure)?
- `satiation` — does this species show a satiation response, or can it always
  find as much food as it can physically ingest?

`reefVulnerable()` computes, for each species and size, the proportion of
individuals *not* protected by refuge — using one of four profile methods
(`"sigmoidal"`, `"binned"`, `"competitive"`, `"noncomplex"`) set with
`setRefuge()`. `"competitive"` is the most detailed: refuge is a limited
resource that same-size individuals compete for, so vulnerability depends on
the actual local density of competitors at each time step, not just on size.

```{r plot-vulnerable, fig.width=6, fig.height=4}
plotVulnerable(params)
```

## Why this can't be a simple additive override

mizerShelf's `getBiomass.mizerShelf()` adds detritus and carrion biomass *on
top of* the standard result — a purely additive modification that a single
`NextMethod()` call expresses perfectly. Predation refuge is different: it
*discounts* the prey available to `blocked_pred` predators before the
encounter and predation-mortality integrals are computed, i.e. it enters
**multiplicatively**, and only for a subset of predators. A single
`NextMethod()` call cannot express "recompute this integral with different
inputs for these rows only".

`projectEncounter.mizerReef()` and `projectPredMort.mizerReef()` solve this by
calling the rate calculation twice — once with unmodified inputs to get the
standard result, and once more with a vulnerability-modified input to get the
correction — and then combining the two by predator row. Both calls go through
`NextMethod()`, so both stay fully composable:

```r
projectEncounter.mizerReef <- function(params, n, n_pp, n_other, t = 0, ...) {
    blocked_pred <- params@species_params$blocked_pred == TRUE
    if (!any(blocked_pred)) {
        return(NextMethod())
    }
    vulnerable <- reefVulnerable(params, n, n_pp, n_other, t,
        new_rd = reefDegrade(params, n, n_pp, n_other, t, ...))

    # Standard encounter (used as-is for predators unaffected by refuge)
    encounter <- NextMethod()
    # Encounter recomputed with vulnerability-reduced prey abundance (used
    # for predators whose foraging is blocked by refuge). Reassigning the
    # formal `n` and then calling NextMethod() bare sends the reduced prey
    # abundance down the full extension chain.
    n <- vulnerable * n
    encounter_vul <- NextMethod()
    encounter[blocked_pred, ] <- encounter_vul[blocked_pred, ]
    encounter
}
```

`projectPredMort.mizerReef()` follows the same pattern, using the linearity of
the predation-mortality calculation in `pred_rate` to isolate the
contribution of `blocked_pred` predators by zeroing out every other
predator's row before recomputing:

```r
projectPredMort.mizerReef <- function(params, n, n_pp, n_other, t = 0,
                                      pred_rate, ...) {
    blocked_pred <- params@species_params$blocked_pred == TRUE
    if (!any(blocked_pred)) {
        return(NextMethod())
    }
    vulnerable <- reefVulnerable(params, n, n_pp, n_other, t,
        new_rd = reefDegrade(params, n, n_pp, n_other, t, ...))

    pm <- NextMethod()
    pred_rate[!blocked_pred, ] <- 0
    pm_blocked <- NextMethod()
    pm + (vulnerable - 1) * pm_blocked
}
```

Because *both* calls go through `NextMethod()` — the standard one and the
vulnerability correction — a lower extension package's contribution to
Encounter or PredMort is preserved for every predator, blocked or not. The
trick is that a bare `NextMethod()` forwards the *current* values of the
method's formal arguments as bound in this frame, not the values the generic
was originally called with. So reassigning `n <- vulnerable * n` (or zeroing
rows of `pred_rate`) before the second bare call is enough to recompute the
integral with modified inputs *through the full chain*, with no direct call to
mizer's base rate function needed.

## A `NextMethod()` pitfall worth knowing about

The obvious way to recompute with a modified input is to pass an override
directly to `NextMethod()`, e.g. `NextMethod(n = vulnerable * n)`. This looks
like it should work — R does let you override a single argument this way — but
testing surfaced a serious problem: once a generic has more than two formal
arguments, passing an explicit named override to `NextMethod()` corrupts the
matching of the *other* arguments. A minimal reproduction:

```r
f <- function(params, n, n_pp, n_other, t = 0, ...) UseMethod("f")
f.default <- function(params, n, n_pp, n_other, t = 0, ...) {
    cat("default sees length(n_pp) =", length(n_pp), "\n")
}
f.foo <- function(params, n, n_pp, n_other, t = 0, ...) {
    NextMethod()               # fine
    NextMethod(n = n * 2)      # n_pp comes out wrong!
}
```

Calling this with a 2×3 matrix `n` and a length-5 vector `n_pp` prints the
correct `length(n_pp) = 5` for the first call, but `length(n_pp) = 6` for the
second — `n_pp` has silently been overwritten by (part of) the new `n`. This
is why the two rate methods above never pass an argument to `NextMethod()`.
Instead they reassign the formal in the local frame (`n <- vulnerable * n`)
and then call `NextMethod()` bare: the modified value still propagates down
the chain, but because no named argument is supplied, the matching of the
remaining arguments is left intact. Reassigning-then-calling-bare gives you
the composability of the chain *and* the modified input, with none of the
argument-corruption.

# Algae and detritus: two unstructured resource components

Algae and detritus are each registered as a scalar *other* component via
`setComponent()` inside `newReefParams()`:

```r
params <- mizer::setComponent(
    params, "algae",
    initial_value = 1,
    dynamics_fun = "algae_dynamics",
    encounter_fun = "encounter_contribution",
    component_params = list(rho = rho_alg, capacity = ..., growth = ...))

params <- mizer::setComponent(
    params, "detritus",
    initial_value = 1,
    dynamics_fun = "detritus_dynamics",
    encounter_fun = "encounter_contribution",
    component_params = list(rho = rho_det, capacity = ..., ...))
```

Both share the same `encounter_contribution()` helper, which adds
$\rho_i(w)\,B$ to species $i$'s encounter rate, where $B$ is the current
biomass of the component and $\rho_i(w)$ is a species- and size-dependent
encounter-rate coefficient stored in `other_params(params)`:

```r
encounter_contribution <- function(params, n_other, component, ...) {
    if (component == "algae") {
        params@other_params$algae_params$rho * n_other[[component]]
    } else if (component == "detritus") {
        params@other_params$detritus_params$rho * n_other[[component]]
    } else {
        params@other_params[[component]]$rho * n_other[[component]]
    }
}
```

`rho` is initialised in `newReefParams()` so that each species can reach
feeding level `f0` at maximum body size once its algae, detritus, and normal
prey encounter are all accounted for.

Both components follow the same dynamical structure — production balanced
against biomass-proportional consumption, solved analytically within each
time step to avoid Euler-step instabilities:

$$\frac{dB}{dt} = P - c\,B \quad\Longrightarrow\quad
B(t+dt) = B(t)\,e^{-c\,dt} + \frac{P}{c}\left(1-e^{-c\,dt}\right)$$

For algae, production is a constant growth rate; for detritus, production
also includes senescence and external decomposition inputs. Both have
getters/setters for inspecting and tuning the steady-state balance, e.g.:

```{r detritus-lifetime}
detritus_lifetime(params)
algae_biomass(params)
```

## `getBiomass.mizerReefSim()`

Standard `getBiomass()` only returns species biomasses. mizerReef's method —
following the same pattern as mizerShelf's `getBiomass.mizerShelf()` — appends
the algae and detritus biomass by reading them directly out of `n_other`:

```r
getBiomass.mizerReefSim <- function(object, ...) {
    sim <- object
    b <- unclass(NextMethod())
    dimname_names <- names(dimnames(b))

    comp_mat <- matrix(unlist(sim@n_other),
        nrow = nrow(sim@n_other), ncol = ncol(sim@n_other))
    dimnames(comp_mat) <- dimnames(sim@n_other)

    b <- cbind(b, comp_mat[rownames(b), , drop = FALSE])
    names(dimnames(b)) <- dimname_names

    mizer::ArrayTimeBySpecies(b, value_name = "Biomass", units = "g",
                              params = sim@params)
}
```

Because `plotBiomass()` calls `getBiomass()` internally, this single method is
enough to make biomass plots include algae and detritus alongside fish
species — and it works regardless of whether `mizer` or `mizerReef` was
loaded first, since it is triggered by S3 dispatch on the object's class
rather than by masking mizer's function.

```{r plot-biomass, fig.width=7, fig.height=4}
sim <- project(params, t_max = 5, progress_bar = FALSE)
plotBiomass(sim)
```

# Senescence and satiation: `projectMort` and `projectFeedingLevel`

Two of the four rate methods are the simple, purely additive/input-transform
case that `NextMethod()` handles cleanly.

`projectMort.mizerReef()` adds senescence mortality, `reefSenMort()`, on top
of the standard mortality:

```r
projectMort.mizerReef <- function(params, n, n_pp, n_other, t = 0,
                                  f_mort, pred_mort, ...) {
    mort <- NextMethod()
    if (isTRUE(params@other_params$include_sen_mort)) {
        mort <- mort + reefSenMort(params, ...)
    }
    mort
}
```

`projectFeedingLevel.mizerReef()` reassigns `params@intake_max` to `Inf` for
species without a satiation response *before* delegating — a formal
argument reassigned inside a method body is correctly picked up by a bare
`NextMethod()` call (unlike the explicit-override case described above):

```r
projectFeedingLevel.mizerReef <- function(params, n, n_pp, n_other, t = 0,
                                          encounter, ...) {
    params@intake_max[params@species_params$satiation == FALSE] <- Inf
    fl <- NextMethod()
    fl[is.na(fl)] <- 0
    fl
}
```

# `removeSpecies.mizerReef()`

Removing a species must also drop its row from the algae and detritus `rho`
matrices, which live outside the standard mizer slots that
`removeSpecies.MizerParams()` already knows how to trim:

```r
removeSpecies.mizerReef <- function(params, species, ...) {
    keep <- !valid_species_arg(params, species, return.logical = TRUE)
    p <- NextMethod()
    p@other_params$algae_params$rho <-
        p@other_params$algae_params$rho[keep, , drop = FALSE]
    p@other_params$detritus_params$rho <-
        p@other_params$detritus_params$rho[keep, , drop = FALSE]
    p
}
```

```{r remove-species}
p2 <- removeSpecies(params, "predators")
species_params(p2)$species
```

# Upgrading objects saved with older mizerReef versions

Versions of mizerReef before 2.1.0 stored refuge, algae, and detritus
parameters in three custom S4 slots (`refuge_params`, `algae_params`,
`detritus_params`) rather than in `other_params()`. Since a dispatching
extension's data must live in `other_params()` rather than in new S4 slots
(see the checklist in `vignette("creating-extension-packages", package =
"mizer")`), those slots were removed from the class definition — which means
an object saved with an older version needs to be migrated before it can be
used. `upgrade.mizerReef()` does this automatically:

```r
upgrade.mizerReef <- function(object, ...) {
    tryCatch({
        old_refuge <- .hasSlot(object, "refuge_params")
        old_algae  <- .hasSlot(object, "algae_params")
        old_det    <- .hasSlot(object, "detritus_params")
        if (old_refuge && is.null(object@other_params$refuge_params)) {
            object@other_params$refuge_params  <- slot(object, "refuge_params")
        }
        if (old_algae && is.null(object@other_params$algae_params)) {
            object@other_params$algae_params   <- slot(object, "algae_params")
        }
        if (old_det && is.null(object@other_params$detritus_params)) {
            object@other_params$detritus_params <- slot(object, "detritus_params")
        }
    }, error = function(e) NULL)
    object
}
```

This method is idempotent and never touches the version stamp, per the
upgrade contract described in `vignette("creating-extension-packages",
package = "mizer")`: mizer's own orchestrator (triggered by `validParams()`,
which runs on essentially every entry point) calls it once and re-stamps the
object afterwards.

# How the pieces fit together in `newReefParams()`

```r
newReefParams <- function(species_params, method, method_params, ...) {
    # Standard multispecies model
    params <- newMultispeciesParams(species_params, ...)

    # Predation refuge: species parameters + refuge profile
    params <- setRefuge(params, method = method, method_params = method_params, ...)
    params <- getRefuge(params, ...)

    # Algae and detritus: scalar dynamical components
    params <- setAlgaeParams(params, ...)
    params <- setDetritusParams(params, ...)
    # ... compute rho so each species reaches feeding level f0 ...
    params <- mizer::setComponent(params, "algae", ...,
                                  dynamics_fun = "algae_dynamics",
                                  encounter_fun = "encounter_contribution")
    params <- mizer::setComponent(params, "detritus", ...,
                                  dynamics_fun = "detritus_dynamics",
                                  encounter_fun = "encounter_contribution")

    # Senescence and external mortality parameters
    params <- setExtMortParams(params, ...)

    # Register the extension chain and promote to the mizerReef S4 class.
    # Rate overrides (Encounter, FeedingLevel, PredMort, Mort) are handled by
    # the project*.mizerReef methods shown above, not by setRateFunction(),
    # so they compose with other extension packages.
    params@extensions <- mizer::getRegisteredExtensions()
    params <- mizer::coerceToExtensionClass(params)
    params
}
```

The marker-class promotion happens last, after every other slot has been
populated — exactly as `vignette("creating-extension-packages", package =
"mizer")` recommends, so that no earlier step in construction accidentally
dispatches to a reef-specific method on a not-yet-fully-built object.

# Summary

mizerReef illustrates the following extension mechanisms working together:

1. **`.onLoad` + `registerExtension()`** — announces mizerReef to mizer's
   extension chain as soon as the package is loaded.

2. **S4 marker classes + S3 dispatch** — `mizerReef` and `mizerReefSim` ensure
   that reef-specific methods (`projectEncounter`, `projectFeedingLevel`,
   `projectPredMort`, `projectMort`, `getBiomass`, `removeSpecies`) are
   dispatched automatically, with every method calling `NextMethod()` at
   least once so the standard mizer pipeline, and any other extension package
   stacked below mizerReef, keeps working.

3. **`setComponent()`** — algae and detritus are scalar dynamical variables
   whose biomass evolves each time step and whose contribution to the
   encounter rate is automatically included via a shared
   `encounter_contribution()` helper.

4. **A multiplicative rate override, handled without a full rewrite** —
   predation refuge cannot be expressed as a purely additive `NextMethod()`
   correction the way mizerShelf's detritus and carrion biomass can. mizerReef
   instead calls `NextMethod()` *twice* — once with unmodified inputs for the
   composable standard result, and once more with vulnerability-modified
   inputs to get the correction — and combines the two by predator row, so
   both calls (and any extension package stacked below mizerReef) stay fully
   composable. This is a pattern worth reusing for any extension whose
   modification depends on a subset of predators or a transformed input,
   rather than a term added on top.

5. **`utils::upgrade()`** — migrates objects created with mizerReef versions
   that stored data in custom S4 slots rather than in `other_params()`.

Together, these mechanisms let mizerReef add predation refuge, benthic
resource coupling, and senescence mortality to any mizer model without
modifying the mizer source code, and without the model's behaviour depending
on the order in which `mizer` and `mizerReef` happen to be loaded.
