---
title: "Functional Group Example"
subtitle: "Karpata Reef Steady State Model Description"
author: ""
bibliography: references.bib
csl: ecology.csl
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functional Group Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(dpi=400)
```

```{r, include = FALSE}
library(ggplot2)
library(dplyr)
library(knitr)
library(mizer)
library(mizerExperimental)
library(mizerReef)

data(caribbean_10_model)
params <- caribbean_10_model
```

\newpage

# Model Description

::: {.content-visible when-format="html"}
[Skip to Size Spectrum Dynamics](#size-spectrum-dynamics)
:::

The model includes `r nrow(params@species_params)-1` species groups of fish as well as a general group of benthic invertebrates. Species groups were assigned based on functional traits, body size, diet, and interactions with habitat structure. Estimates of observed biomass for each group were used to tune reproduction and resource consumption parameters so that steady state abundances agree with empirical observations. The process of establishing a steady state that agrees with empirical observations is nontrivial. The R script used to tune the steady state parameters is included at the end of this file. 

Biomass estimates were based on data collected from Karpata Reef in Bonaire, a site with low fishing levels in a marine reserve [@Dryden2016; @Rogers_2014; @Newman_2015; @Williams_2015; @Williams_2016]. Observations of fish smaller than 10 cm were removed from analysis because their abundance is not well captured by visual survey methods. The cutoff weight for observed biomass estimates was calculated using published length-weight relationships [@FishBase]. The cutoff size is given in the table below along with the observed biomass per square meter in grams.

```{r, echo = FALSE}
params@species_params %>%
    filter(!is.na(biomass_observed)) %>%
    select(species,biomass_observed, biomass_cutoff) %>%
    kable(row.names = FALSE,
          col.names = c("Species Group", 
                        "Observed Biomass [g/m^2]", "Cutoff Size [g]"))
```

# Size Spectrum Dynamics

::: {.content-visible when-format="html"}
[Skip to Resources](#resources)
:::

These dynamics expand on those used in standard Mizer models, see @mizer.

## Growth

The growth of organisms is dependent on the energy they are able to obtain from consumed food resources.

### Predator-prey encounter rate  
  
The rate $E_{i}(w)$ at which a predator of species $i$ and weight $w$ encounters food is dictated by the predation power $\gamma_i(w)$, interaction matrix $\theta_{ij}$, the vulnerability to predation $V_{ij}(w_p)$, and the size selectivity of predator, given by the predation kernel $\phi_i(w,w_p)$.

#### Interaction Matrix  
  
The $\theta_{ij}$ matrix sets the interaction strength between predator group $i$ prey group $j$. The predator/prey interaction matrix has entries between 0 (if the groups can not interact) and 1, see the figure below. All organisms were assumed to interact equally, with the exception of nocturnal invertivores which are known to hunt off the reef at night.

```{r, echo = FALSE, fig.height = 6, fig.width = 7, fig.cap = "Interaction strength between each predator group (rows) and prey group (columns) in the 10-group Karpata Reef model."}
ggplot(melt(params@interaction),
       aes(x = prey, y = reorder(predator, desc(predator)), fill = value)) +
    geom_tile(colour = "black", linewidth = 0.25) +
    scale_fill_gradient(low = "white", high = "blue", limit = c(0, 1),
                        guide = "legend") +
    labs(y = "Predator", x = "Prey", fill = "Interaction") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

#### Predation Kernel   
  
The parameters for the predation kernels were based on trait-based studies of prey size selectivity, diet studies that estimate predator prey mass ratio, and home range size estimates from @Nash_2014. Values were tuned to achieve expected diet compositions. All estimates fall within observed ranges. All groups use a lognormal predation kernel. The parameters are given in the table below. 

```{r, echo = FALSE, results='asis'}
params@species_params %>%
    filter(pred_kernel_type == "lognormal") %>%
    select(beta, sigma, gamma) %>%
    kable()
```

### Vulnerability to Predation   
```{r, include = FALSE}
refuge_params <- params@other_params$refuge_params
```

The refuge function, $R_j(w_p)$, describes the proportion of fish of size $w_p$ in prey group $j$ that are hidden from predators. $1 - R_j(w_p)$ is then the proportion of fish of weight $w_p$ and prey group $j$ that are vulnerable to consumption by predators.  This model uses the `r print(params@other_params$refuge_params$method)` method to determine the **refuge profile**, the set of proportions that describe refuge availability across the entire size range of model fish (@Rogers_2018). The proportion of prey of weight $w_p$ and group $j$ with access to refuge $R_j(w_p)$ is given by

$$
R_j(w_p) = min\left \{R_{max} \, , \frac{\tau\cdot\eta_k}{\sum_{i}\int_{w_{k-1}}^{w_k} N_i(w)~dw}
            ~~~~~~~ w_p \in (~w_{k-1}, w_k~] \right \}
$$ {#eq-refuge_data}

The parameter $\tau$ is the proportion of fish with access to refuge that are expected to utilize it, $\eta_k$ is the density ($no./m^{2}$) of refuges in size range $(w_{k-1}, w_k]$ and $\sum_{i} \int_{w_{k-1}}^{w_k} N_i(w)~dw$ gives the total density ($no./m^{2}$) of fish from any group in size range $(w_{k-1}, w_k]$. This represents the density of competitors for refuges in size class $k$. With this method, the refuge profile is density dependent. 

All fish are assumed to utilize refuge ($\tau = 1$). Fish smaller than `r paste(refuge_params$w_settle,"g")` are assumed to be larval reef fish that have not yet settled to the reef. A maximum proportion of `r paste(100*refuge_params$max_protect,"%")` of fish are protected at any given time. 

<details>
<summary>Refuge length bins and densities used in the steady state (Karpata Reef, FORCE data)</summary>

```{r, echo = FALSE}
#|
params@other_params$refuge_params$method_params %>%
    mutate(refuge_density = round(refuge_density,3)) %>%
    kable(row.names = FALSE,
          col.names = c("Start of Bin (cm)",
                        "End of Bin (cm)",
                        "Refuge Density (no./m^2)"))
```

</details>

<details>
<summary>How each species group interacts with predation refuge</summary>

Each species group utilises benthic structures differently depending on their specific traits. The table below indicates how each species group interacts with structures that provide predation refuge.

```{r, echo = FALSE, results='asis'}
#|
params@species_params %>%
    select(refuge_user, blocked_pred) %>%
    mutate(refuge_user   = case_when(refuge_user   == TRUE ~ 'Yes',
                                     refuge_user   == FALSE ~ '×')) %>%
    mutate(blocked_pred = case_when(blocked_pred == TRUE ~ '×',
                                    blocked_pred == FALSE ~ 'Yes')) %>%
    kable(col.names = c("Group", "Uses refuge?", "Accesses prey in refuge?"))
```

</details>

the figure below shows the density-dependent refuge profile produced by the competitive method for the simple trait-based Bonaire model at steady state.

```{r, echo = FALSE, fig.height = 6, fig.width = 7, fig.cap = "Proportion of each refuge-using group protected from predation, by body length, at steady state, for the 10-group Karpata Reef model."}
plotRefugeProfile(params)
```

\newpage

### Consumption  
  
Invertebrate consumption of the detrital resource and planktivory are subject to a Holling functional response type II to represent satiation. This relationship is defined by the maximum intake rate, which increases allometrically with body size at rate $n$. The parameter $h$ is the max consumption rate for an invertebrate consumer of size 1 gram. Values for $h$ were chosen so that invertebrates are neither too starved nor totally satiated.

Only a proportion $\alpha_i$ of consumed biomass is retained, while a proportion $1-\alpha_i$ is expelled in the form of feces, which contribute to the detritus. See the table below for the max consumption rate for invertebrates.

No maximum consumption rate is imposed for predatory or herbivorous groups.

```{r, echo = FALSE}
params@species_params %>%
    filter(satiation == TRUE) %>%
    select(h, alpha, n) %>%
    kable()
```

#### Metabolic Losses  
  
The energy losses to metabolic needs are comprised of two components. Standard metabolism occurs at rate $k_{s.i}$, scaling allometrically with body size at rate $p$. The units of the coefficients $k_{s.i}$ are $\text{grams}^{1-p}$ per year. Losses due to activity and movement occur at rate $k_i$ in grams per year, scaling with body size at rate $1$.

```{r, results='asis', echo = FALSE}
params@species_params %>%
    select(ks, p, k) %>%
    kable()
```

#### Energy Invested into Reproduction  
  
A proportion $\psi_i(w)$ of the energy available for growth and reproduction is used for reproduction. This proportion changes from zero below the weight $w_{mat.i}$ of maturation to one at the maximum weight $w_{max.i}$, where all available energy is used for reproduction.

Maturation length and age data were based on the most observed species from each species group in the FORCE data set [@Dryden2016; @Rogers_2014; @Newman_2015; @Williams_2015; @Williams_2016].

```{r, echo = FALSE}
params@species_params %>%
    select(w_mat, w_max, age_mat) %>%
    kable()
```

#### Somatic growth  
  
Energy that is left over after metabolism and reproduction is invested in somatic growth. The growth rate of an individual of from group $i$ and weight $w$ is 
$$
  g_i(w) = E_{r.i}(w)\left(1-\psi_i(w)\right).
$$ {#eq-growth} 
When food supply does not cover the requirements of metabolism and activity, growth and reproduction stops. There is no negative growth or starvation mortality.

The values for the model parameters were chosen so that the resulting growth curves would be close to von Bertalanffy growth curves. The parameters in the table below were taken from the literature.

```{r, echo = FALSE}
params@species_params %>%
    select(k_vb, w_max, a, b) %>%
    kable()
```

Here the parameters $a$ and $b$ are parameters for the allometric weight-length relationship $w = a l^b$ where $w$ is measured in grams and $l$ is measured in centimetres.


## Mortality {#mortality}  
  
The mortality rate $\mu_i(w)$ of an individual of group $i$ and weight
$w$ has three sources: predation mortality $\mu_{p.i}(w)$, external
mortality $\mu_{ext.i}(w)$, fishing mortality $\mu_{f.i}(w)$. External
mortality is composed of residual natural mortality $\mu_{nat.i}(w)$,
and senescence $\mu_{sen.i}(w)$. The mortality rate for group $i$ is
then given by: $$
\mu_i(w)=\mu_{p.i}(w)+\mu_{ext.i}(w)+\mu_{f.i}(w)
$$ {#eq-mort}

### Predation mortality  
  
All consumption by fish translates into corresponding predation
mortalities on the ingested prey individuals. the rate at which all
predators from $j$ consume prey of size $w_p$ is $$
  \mathtt{pred\_rate}_j(w_p) = \int \phi_j(w,w_p) \gamma_j(w) N_j(w) \, dw.
$$ {#eq-pred_rate}

The mortality rate due to predation is then obtained as $$
  \mu_{p.i}(w_p) = \sum_j \mathtt{pred\_rate}_j(w_p)\, V_{ji}(w_p)\, \theta_{ji}.
$$ {#eq-mup}

### Fishing Mortality  
  
Like in `mizer`, fishing mortality in `mizerReef` is imposed  by fishing gears. The total per-capita fishing mortality (1/year) is obtained by summing over the mortality from all gears,

$$\mu_{f.i}(w) = \sum_g F_{g,i}(w)$$

where the fishing mortality $F_{g,i}(w)$ imposed by gear $g$ on group $i$ at size $w$ is calculated as:

$$F_{g,i}(w) = S_{g,i}(w) Q_{g,i} E_{g}$$

The constant $S$ is the selectivity by group, gear and size, $Q$ is the catchability by group and gear and $E$ is the fishing effort by gear.

Fishing parameters for mizerReef models can be set up with `setFishing()`, which contains the details of how to set up gears with different selectivities and
catchabilities for each group.

Fishing mortality $\mu_{f.i}(w)$ is calculated with the function `getFMort()`.
The vignettes were run with following fishing parameters (the table below).

```{r, echo = FALSE}
params@gear_params %>%
    filter(catchability > 0) %>%
    select(species, knife_edge_size, catchability) %>%
    kable(row.names = FALSE,
          col.names = c("Group", "Minimum fishing size [g]", 
                        "Catchability [1/year]"))
```

### External Mortality   
  
```{r, results = 'asis', echo = FALSE}
ext_mort <- params@other_params[['ext_mort_params']]
```

#### Residual natural mortality  
  
Mortality caused by illness, fishing, or predators not explicitly
included in the model is captured by $\mu_{nat.i}(w)$, which is
independent of group identities and abundances. It is assumed
to decrease allometrically with body size:

$$
\mu_{nat.i}(w) = \mu_{nat} w^{1 - n}
$$ {#eq-z0} where $\mu_{nat}$ is the residual natural mortality rate and
$n$ is the allometric scaling exponent. We use a residual mortality rate of 
$\mu_{nat} =$ `r ext_mort$nat_mort` per year at size 1 gram.

#### Senescence mortality   
  
Senescence mortality $\mu_{sen.i}(w)$ is intended to capture mortality
due to illness or old age. It is independent of group abundances.
Senescence mortality is assumed to increase allometrically with body
size. The rate of senescence mortality is given by:

$$
\mu_{sen.i}(w) =  k_{sen} \left(\frac{log_{10}(w)}
                                     {log_{10}(w_{max.i})}\right)^{p_{sen}}
$$ {#eq-extmort} where $k_{sen}$ is the rate of senescence mortality,
$p_{sen}$ defines the slope of the senescence curve, and $w_{max.i}$ is
the maximum body size of species group $i$ in grams.

The senescence mortality rate for mortalities due to illness and old age is 
set to $k_{sen} =$ `r ext_mort$sen_prop` per year at size 1 gram and increases
allometrically with body size with slope $p_{sen} =$ `r ext_mort$sen_curve`.

## Reproduction {#reproduction}
The reproduction parameters $\epsilon_i$ and $R_{max.i}$ are not directly observable. The values were instead chosen so as to produce steady-state abundances of the groups that are in line with observations and to give reasonable values for the reproduction level.

the table below gives the steady-state reproduction level which is defined as the ratio between the actual reproduction rate $R_i$ and the maximal possible reproduction rate $R_{\max.i}$.

```{r, echo = FALSE}
df <- params@species_params
#df$reproduction_level <- getReproductionLevel(params)
df %>%
    select(w_min, erepro, R_max) %>% # , reproduction_level) %>%
    kable()
```

# Resources {#resources}

::: {.content-visible when-format="html"}
[Skip to Tuning the Steady State](#tuning-the-steady-state)
:::

The fish spectrum is fed by three background resources: plankton , algae, and detritus. Small individuals of all species will feed on plankton while only herbivorous groups and invertebrates feed on algae or detritus. The general mathematical form of the algae and detritus dynamics -- a linear production/consumption balance for each pool, solved analytically each time step -- was adapted from @mizerShelf [@deJuan2023]. This is not a direct port: mizerShelf represents its detritus/carrion pools by rescaling the total biomass of mizer's size-structured background resource spectrum, whereas mizerReef's algae and detritus are genuinely unstructured scalar pools with their own dedicated encounter-rate matrices, added as independent components via `setComponent()`. mizerReef also adds coral-reef-specific behaviour with no equivalent in mizerShelf, most notably that algae production is treated as a fixed rate of primary production decoupled from consumer demand rather than tuned to match consumption (see `algae_consumption()`'s documentation for the ecological rationale and citations).

### Plankton {#plankton}
```{r, echo = FALSE}
wcutoff <- params@resource_params$w_pp_cutoff
lambda <- params@resource_params$lambda
kappa <- params@resource_params$kappa
```

The plankton spectrum $N_P(w)$ tracks the abundance all planktonic food sources. This spectrum starts at a smaller size than the fish spectrum in order to provide food for the smallest individuals (larvae) of the fish spectrum. 

The plankton spectrum ranges from $w_0=`r signif(min(params@w_full), digits = 1)`$ to $w_{cutoff}=`r wcutoff`$ grams. The steady state abundance of plankton at size 1 gram is $\kappa=`r signif(kappa, digits = 3)`[g/m^{-2}]$. The slope of the plankton spectrum is set to $\lambda=`r signif(lambda, digits = 3)`$.

### Algae
The algal resource is described only by its total biomass $B_A$. Feeding on algae is not is not size-based. Herbivores can feed on algae of any size. In the steady state the total 
algal biomass per square meter is $B_A = `r signif(algae_biomass(params), digits = 3)`$ grams.

#### Algal consumption  
  
For each consumer species $i$, a parameter $\rho_{A.i}$ determines the rate at which individuals of that species encounter algal biomass. The parameters $\rho_{i.A}$ have units of $g^{-n}$ per year. They are non-zero only for species that consume at least some algae. The preference $\theta_{i.A}$ for algae is a value between 0 and 1 specifying the proportion of consumer diets that are comprised of algal matter.

```{r, echo = FALSE}
params@species_params %>%
    select(rho_algae, interaction_algae) %>%
    rename(rho = rho_algae) %>%
    filter(rho > 0) %>%
    kable()
```

#### Algal production  
  
```{r, echo = FALSE, include=FALSE}
pA <- params@other_params$algae$growth
```

The rate at which algal biomass is produced by the ecosystem is given by a constant growth rate with units of grams per unit area per unit time. At steady state algal production is `r signif(pA,digits = 4)` grams per square meter per year.


### Detritus
Detritus is consumed by herbivores and benthic invertebrates. Feeding on detritus is not size-based as fish can feed on detritus particles of any size. The detritus resource is described only by its total biomass $B_D$. In the steady state the total detrital biomass per square meter is $B_D = `r signif(detritus_biomass(params), digits = 4)`$ grams.

#### Detrital consumption  
  
For each consumer species $i$, a parameter $\rho_{D.i}$ determines the rate at which individuals of that species encounter detritus. The parameters $\rho_{i.D}$ have units of $g^{-n}$ per year. They are non-zero only for species that consume at least some detritus. The preference $\theta_{i.D}$ for detritus is a value between 0 and 1 specifying the proportion of consumer diets comprised of detritus.

```{r, echo = FALSE}
params@species_params %>%
    select(rho_detritus,interaction_detritus) %>%
    rename(rho = rho_detritus) %>%
    filter(rho > 0) %>%
    kable()
```

#### Detrital production  
  
```{r, echo = FALSE}
#| include: false
e_d <- params@other_params$detritus$ext_decomp
s_d <- params@other_params$detritus$sen_decomp
r <- getDetritusProduction(params)
pf <- r[1]
pd <- r[2]
dext <- params@other_params$detritus$external
```
Detritus comes from defecation and decomposing dead organisms that die as a result of sources of external mortality. At steady state detritus from decomposing dead organisms is `r paste(round(pd,2))` g/year and defecation produces `r paste(round(pf,2))` g/year. A proportion `ext_decomp` of the external mortality and a proportion `sen_decomp` of senescence mortality decomposes to detritus.  The proportion of detritus that sinks from external mortality is `r paste(100*e_d,"%")` and the proportion that decomposes from senescence mortality is `r paste(100*s_d,"%")`. These values are based on estimates from @Hatcher_1988.

External detritus is generated by unmodelled sources or sinks in from the pelagic zone. At steady state external detritus production is `r paste(round(dext,2))` g/year. This value is set so that steady state abundances match empirical observations. Where it is negative, feces and mortality are producing more detritus than can be consumed. Detrital biomass is assumed to be washed away by wave action in this case.

# Tuning the Steady State
The following R script was used to tune the steady state parameters for this model.

<details>
<summary>Show the full steady-state calibration script</summary>

```{r, echo = TRUE, eval = FALSE}
# Setting up a Caribbean coral reef mizer model with multiple resources
# Model steady state calibration
## Setup - load packages ----------------------------------
library(ggplot2)
library(mizer)
library(mizerExperimental)
library(mizerReef)
library(assertthat)
library(here)

## Load parameters -----------------------------------
karpata_10plus  <- read.csv(here("inst/data-csv/caribbean_10_species.csv"))
karpata_int     <- read.csv(here("inst/data-csv/caribbean_10_interaction.csv"),
                            row.names = 1)
karpata_refuge  <- karpata_refuge
tuning_profile  <- tuning_profile

# Herbivores consume plankton at small sizes and 
#   transition to detritus and algae as they grow
# Invertebrates consume plankton and detritus,
#   with the proportion of detritus increasing with size

## Set model ----------------------------------------
params <- newReefParams(species_params = karpata_10plus,
                        interaction = karpata_int,
                        method = "binned",
                        method_params = tuning_profile)

## Reduce density dependent of reproduction ----------------
rdi <- rep(0.5, dim(karpata_int)[1])

params <- setBevertonHolt(params, reproduction_level = rdi)
getReproductionLevel(params)

## Project to first steady state -------------------------------
params <- params |>
    reefSteady() |> reefSteady() |> reefSteady() |> reefSteady() |>
    reefSteady() |> reefSteady() 

## Calibrate biomasses and growth ---------------------------------
# Match observed species group biomasses
params <- calibrateReefBiomass(params)
params <- matchBiomasses(params)
params <- reefSteady(params)

# Match observed growth rates
params <- matchReefGrowth(params)
params <- reefSteady(params)

# Iterate to refine biomass
params <- params |>
    calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
    reefSteady()|>
    calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
    reefSteady()|>
    calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
    reefSteady()

# Check biomass match
plotBiomassVsSpecies(params) # spot on

# Check match with observed age at maturity
age_mat_observed = karpata_10plus$age_mat
age_mat_model = age_mat(params)
data.frame(age_mat_model, age_mat_observed)
# Closer than needed

# Check predation mortality, feeding levels, and diets
plotPredMort(params) + facet_wrap(~Species)
plotFeedingLevel(params)
plotDiet(params) + scale_x_log10(limits = c(1, 1e4))
plotSpectra(params, power = 1)

## Now switch to competitive method ----------------------------
params <- newRefuge(params,
                    new_method = "competitive",
                    new_method_params = karpata_refuge)

# Match biomasses again
params <- params |>
    matchBiomasses()|> reefSteady()|> 
    matchBiomasses()|> reefSteady()|>
    matchBiomasses()|> reefSteady()|>
    matchBiomasses()|> reefSteady() 

# Make sure new refuge is in place
plotVulnerable(params)

plotBiomassVsSpecies(params) # spot on

# Check match with observed age at maturity
age_mat_observed = karpata_10plus$age_mat
age_mat_model = age_mat(params)
data.frame(age_mat_model, age_mat_observed)
# Still look good

## Check resulting spectra and tune resources------------------------

# Resource looks low - should match sheldon's spectrum
# looks fairly straight not bad but some bumps
plotSpectra(params, total = TRUE, power = 1)
plotSpectra(params, total = TRUE, power = 2)

# plot feeding level to check if resource is too low
plotFeedingLevel(params, species = "inverts")

# Invert feeding level is relatively stable through life, non-linearities
#   are probably due to refuge

# Tune reproduction -----------------------------------------------
# We do not have yield or catch data - can't tune size distribution
# First attempt to set very low to see what the minimum values are
params <- setBevertonHolt(params, erepro = 0.0001)
# Now set setting erepro same for all species, as low as possible
params <- setBevertonHolt(params, erepro = 0.35)
# Project back to steady
params <- reefSteady(params)
# Check reproduction level (value between 0 and 1) - should be higher for
# larger, slow growing species and low for small, fast growing ones
getReproductionLevel(params)
# A reproduction level closer to one means reproduction rate is 
# almost totally independent of the investment into reproduction
# These are near one for all species except farming damsels

# Check comparison of density dependent & independent reduction
getRDI(params) / getRDD(params)
# Reproduction is density independent for almost all species

# Let's increase reproduction level back to 0.5 so there is still some
# density dependence
params <- setBevertonHolt(params, reproduction_level = rdi)

# Iterate to get back to steady state
params <- params |>
    reefSteady()|>
    reefSteady()|>
    reefSteady()

# Check new reproduction - these look better
rep <- getReproductionLevel(params)
getRDI(params) / getRDD(params)

# Check new spectra & plots
plotSpectra(params, total = TRUE, power = 2)
plotPredMort(params) + facet_wrap(~Species)
plotFeedingLevel(params)
plotDiet(params) + scale_x_log10(limits = c(1, 1e4))
plotSpectra(params, power = 1)

# Save!
caribbean_10_model <- reefSteady(params)
```

</details>