---
title: "Functional Group Example"
subtitle: "Karpata Reef Steady State Model Description"
author: ""
bibliography: references.bib
csl: ecology.csl
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Functional Group Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(dpi=400)
```

```{r, include = FALSE}
library(ggplot2)
library(dplyr)
library(knitr)
library(mizer)
library(mizerExperimental)
library(mizerReef)

data(caribbean_10_model)
params <- caribbean_10_model
```

\newpage

# Model Description

::: {.content-visible when-format="html"}
[Skip to Resources](#resources)
:::

`caribbean_10_model` is mizerReef's field-calibrated flagship example: `r nrow(params@species_params)-1`
fish species groups plus a general group of benthic invertebrates,
parameterised against field data from Karpata Reef, Bonaire. Species groups
were assigned based on functional traits, body size, diet, and interactions
with habitat structure. It is the finer-resolution counterpart to the
[3-group Caribbean model](caribbean_3_model-description.html), which
reproduces the coarser trait-based structure of @Rogers_2018 using the same
marine-reserve dataset described below. This page documents this model's
specific parameters and how they were tuned; for the general mizerReef
formulation these parameters plug into, see `vignette("model-description")`.

Biomass estimates were based on data collected from Karpata Reef in Bonaire,
a site with low fishing levels in a marine reserve [@Dryden2016;
@Rogers_2014; @Newman_2015; @Williams_2015; @Williams_2016]. Observations of
fish smaller than 10 cm were removed from analysis because their abundance is
not well captured by visual survey methods. The cutoff weight for observed
biomass estimates was calculated using published length-weight relationships
[@FishBase]. The cutoff size is given in the table below along with the
observed biomass per square meter in grams.

```{r, echo = FALSE}
params@species_params %>%
    filter(!is.na(biomass_observed)) %>%
    select(species,biomass_observed, biomass_cutoff) %>%
    kable(row.names = FALSE,
          col.names = c("Species Group",
                        "Observed Biomass [g/m^2]", "Cutoff Size [g]"))
```

## Interaction matrix

The $\theta_{ij}$ matrix sets the interaction strength between predator group
$i$ and prey group $j$, with entries between 0 (groups cannot interact) and
1. All organisms were assumed to interact equally, with the exception of
nocturnal invertivores which are known to hunt off the reef at night.

```{r, echo = FALSE, fig.height = 6, fig.width = 7, fig.cap = "Interaction strength between each predator group (rows) and prey group (columns) in the 10-group Karpata Reef model."}
ggplot(melt(params@interaction),
       aes(x = prey, y = reorder(predator, desc(predator)), fill = value)) +
    geom_tile(colour = "black", linewidth = 0.25) +
    scale_fill_gradient(low = "white", high = "blue", limit = c(0, 1),
                        guide = "legend") +
    labs(y = "Predator", x = "Prey", fill = "Interaction") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

## Predation kernel

The parameters for the predation kernels were based on trait-based studies of
prey size selectivity, diet studies that estimate predator-prey mass ratio,
and home range size estimates from @Nash_2014, tuned to achieve expected diet
compositions. All estimates fall within observed ranges. All groups use a
lognormal predation kernel.

```{r, echo = FALSE, results='asis'}
params@species_params %>%
    filter(pred_kernel_type == "lognormal") %>%
    select(beta, sigma, gamma) %>%
    kable()
```

## Vulnerability to predation

```{r, include = FALSE}
refuge_params <- params@other_params$refuge_params
```

This model uses the `r print(params@other_params$refuge_params$method)`
refuge method (see `vignette("model-description")`'s
[refuge profiles](model-description.html#refuge-profiles) section for the
general formulation and @Rogers_2018 for its derivation). All fish are
assumed to utilize refuge ($\tau = 1$). Fish smaller than
`r paste(refuge_params$w_settle,"g")` are assumed to be larval reef fish that
have not yet settled to the reef. A maximum proportion of
`r paste(100*refuge_params$max_protect,"%")` of fish are protected at any
given time.

<details>
<summary>Refuge length bins and densities used in the steady state (Karpata Reef, FORCE data)</summary>

```{r, echo = FALSE}
#|
params@other_params$refuge_params$method_params %>%
    mutate(refuge_density = round(refuge_density,3)) %>%
    kable(row.names = FALSE,
          col.names = c("Start of Bin (cm)",
                        "End of Bin (cm)",
                        "Refuge Density (no./m^2)"))
```

</details>

<details>
<summary>How each species group interacts with predation refuge</summary>

Each species group utilises benthic structures differently depending on
their specific traits.

```{r, echo = FALSE, results='asis'}
#|
params@species_params %>%
    select(refuge_user, blocked_pred) %>%
    mutate(refuge_user   = case_when(refuge_user   == TRUE ~ 'Yes',
                                     refuge_user   == FALSE ~ '×')) %>%
    mutate(blocked_pred = case_when(blocked_pred == TRUE ~ '×',
                                    blocked_pred == FALSE ~ 'Yes')) %>%
    kable(col.names = c("Group", "Uses refuge?", "Accesses prey in refuge?"))
```

</details>

The figure below shows the density-dependent refuge profile produced by the
competitive method at steady state.

```{r, echo = FALSE, fig.height = 6, fig.width = 7, fig.cap = "Proportion of each refuge-using group protected from predation, by body length, at steady state, for the 10-group Karpata Reef model."}
plotRefugeProfile(params)
```

\newpage

## Consumption

Invertebrate consumption of the detrital resource and planktivory are
subject to a Holling type II functional response (see
`vignette("model-description")`'s [Consumption and
satiation](model-description.html#consumption-and-satiation) section for the
general formula). $h$ is the max consumption rate for an invertebrate
consumer of size 1 gram, chosen so that invertebrates are neither too
starved nor totally satiated; $\alpha_i$ is the proportion of consumed
biomass retained, with $1-\alpha_i$ expelled as faeces, which contribute to
detritus. No maximum consumption rate is imposed for predatory or herbivorous
groups.

```{r, echo = FALSE}
params@species_params %>%
    filter(satiation == TRUE) %>%
    select(h, alpha, n) %>%
    kable()
```

### Metabolic losses

Standard metabolism occurs at rate $k_{s.i}$; losses due to activity and
movement occur at rate $k_i$ (see `vignette("model-description")` for the
general growth/reproduction formulation).

```{r, results='asis', echo = FALSE}
params@species_params %>%
    select(ks, p, k) %>%
    kable()
```

### Maturation and growth

Maturation length and age data were based on the most observed species from
each species group in the FORCE data set [@Dryden2016; @Rogers_2014;
@Newman_2015; @Williams_2015; @Williams_2016].

```{r, echo = FALSE}
params@species_params %>%
    select(w_mat, w_max, age_mat) %>%
    kable()
```

The values for the growth parameters below were chosen so that the resulting
growth curves would be close to von Bertalanffy growth curves; $a$ and $b$
are the allometric weight-length parameters $w = a l^b$ ($w$ in grams, $l$ in
centimetres), taken from the literature. There is no negative growth or
starvation mortality.

```{r, echo = FALSE}
params@species_params %>%
    select(k_vb, w_max, a, b) %>%
    kable()
```

## Mortality {#mortality}

See `vignette("model-description")`'s [Mortality](model-description.html#mortality)
section for the general formulation. This model's residual-natural- and
senescence-mortality parameters:

```{r, results = 'asis', echo = FALSE}
ext_mort <- params@other_params[['ext_mort_params']]
```

We use a residual natural mortality rate of $\mu_{nat} =$
`r ext_mort$nat_mort` per year at size 1 gram, and senescence mortality
parameters $k_{sen} =$ `r ext_mort$sen_prop` (`sen_prop`) and $p_{sen} =$
`r ext_mort$sen_curve` (`sen_curve`, the exponent governing how steeply
mortality climbs as individuals approach their maximum size), both based on
estimates from @Hatcher_1988.

### Fishing mortality

Fishing parameters for mizerReef models can be set up with `setFishing()`.
The vignettes were run with the following fishing parameters:

```{r, echo = FALSE}
params@gear_params %>%
    filter(catchability > 0) %>%
    select(species, knife_edge_size, catchability) %>%
    kable(row.names = FALSE,
          col.names = c("Group", "Minimum fishing size [g]",
                        "Catchability [1/year]"))
```

## Reproduction {#reproduction}

The reproduction parameters $\epsilon_i$ and $R_{max.i}$ are not directly
observable. Their values were instead chosen so as to produce steady-state
abundances of the groups that are in line with observations and to give
reasonable values for the reproduction level -- the ratio between the actual
reproduction rate $R_i$ and the maximal possible reproduction rate
$R_{\max.i}$.

```{r, echo = FALSE}
df <- params@species_params
df %>%
    select(w_min, erepro, R_max) %>%
    kable()
```

# Resources {#resources}

::: {.content-visible when-format="html"}
[Skip to Tuning the Steady State](#tuning-the-steady-state)
:::

The fish spectrum is fed by three background resources: plankton, algae, and
detritus. Small individuals of all species feed on plankton, while only
herbivorous groups and invertebrates feed on algae or detritus. See
`vignette("model-description")`'s [unstructured resource
dynamics](model-description.html#unstructured-resource-dynamics) section for
the general formulation and its relationship to mizerShelf.

### Plankton {#plankton}
```{r, echo = FALSE}
wcutoff <- params@resource_params$w_pp_cutoff
lambda <- params@resource_params$lambda
kappa <- params@resource_params$kappa
```

The plankton spectrum ranges from $w_0=`r signif(min(params@w_full), digits = 1)`$
to $w_{cutoff}=`r wcutoff`$ grams. The steady state abundance of plankton at
size 1 gram is $\kappa=`r signif(kappa, digits = 3)` (g/m^{2})$, with slope
$\lambda=`r signif(lambda, digits = 3)`$.

### Algae {#algae}
The algal resource is described only by its total biomass $B_A$, and feeding
on it is not size-based. In the steady state the total algal biomass per
square meter is $B_A = `r signif(algae_biomass(params), digits = 3)`$ grams.

```{r, echo = FALSE}
params@species_params %>%
    select(rho_algae, interaction_algae) %>%
    rename(rho = rho_algae) %>%
    filter(rho > 0) %>%
    kable()
```

```{r, echo = FALSE, include=FALSE}
pA <- params@other_params$algae$growth
```

Algal *production* -- a fixed, literature-informed constant, `algae_growth_initial`
in `setAlgaeParams()` -- is `r signif(pA,digits = 4)` grams per square meter
per year at steady state. Note this value is not retuned to match
consumption the way detritus production is (see below); see
`vignette("model-description")`'s Algae section for why.

### Detritus
Detritus is consumed by herbivores and benthic invertebrates, also not
size-based. In the steady state the total detrital biomass per square meter
is $B_D = `r signif(detritus_biomass(params), digits = 4)`$ grams.

```{r, echo = FALSE}
params@species_params %>%
    select(rho_detritus,interaction_detritus) %>%
    rename(rho = rho_detritus) %>%
    filter(rho > 0) %>%
    kable()
```

```{r, echo = FALSE}
#| include: false
e_d <- params@other_params$detritus$ext_decomp
s_d <- params@other_params$detritus$sen_decomp
r <- getDetritusProduction(params)
pf <- r[1]
pd <- r[2]
dext <- params@other_params$detritus$external
```
Detritus production comes from three sources: defecation, decomposing dead
organisms, and external input. At steady state, decomposing dead organisms
contribute `r paste(round(pd,2))` g/year and defecation contributes
`r paste(round(pf,2))` g/year -- `r paste(100*e_d,"%")` of external mortality
and `r paste(100*s_d,"%")` of senescence mortality is assumed to decompose to
detritus (estimates from @Hatcher_1988). The remaining external-input term is
solved for so that, unlike algae, detritus *production* matches current
consumption: at steady state it is `r paste(round(dext,2))` g/year. Where
this is negative, faeces and mortality alone already produce more detritus
than is consumed, and the excess is assumed to be washed away.

# Tuning the Steady State

The following R script was used to tune the steady state parameters for this
model (see `vignette("steady-state-recipe")` for the general recipe this
follows, including the final algae/detritus absolute-scale rescaling step).

<details>
<summary>Show the full steady-state calibration script</summary>

```{r, echo = TRUE, eval = FALSE}
# Setting up a Caribbean coral reef mizer model with multiple resources
# Model steady state calibration
## Setup - load packages ----------------------------------
library(ggplot2)
library(mizer)
library(mizerExperimental)
library(mizerReef)
library(assertthat)
library(here)

## Load parameters -----------------------------------
karpata_10plus  <- read.csv(here("inst/data-csv/caribbean_10_species.csv"))
karpata_int     <- read.csv(here("inst/data-csv/caribbean_10_interaction.csv"),
                            row.names = 1)
karpata_refuge  <- karpata_refuge
tuning_profile  <- tuning_profile

# Herbivores consume plankton at small sizes and
#   transition to detritus and algae as they grow
# Invertebrates consume plankton and detritus,
#   with the proportion of detritus increasing with size

## Set model, with info_level = 1 to keep only the reports that say
## something went differently from what was asked, not every default fill ---
params <- newReefParams(species_params = karpata_10plus,
                        interaction = karpata_int,
                        method = "binned",
                        method_params = tuning_profile,
                        info_level = 1)

## Reduce density dependent of reproduction ----------------
rdi <- rep(0.5, dim(karpata_int)[1])

params <- setBevertonHolt(params, reproduction_level = rdi)
getReproductionLevel(params)

## Project to first steady state -------------------------------
params <- params |>
    reefSteady() |> reefSteady() |> reefSteady() |> reefSteady() |>
    reefSteady() |> reefSteady()

## Calibrate biomasses and growth ---------------------------------
# Match observed species group biomasses
params <- calibrateReefBiomass(params)
params <- matchBiomasses(params)
params <- reefSteady(params)

# Match observed growth rates
params <- matchReefGrowth(params)
params <- reefSteady(params)

# Iterate to refine biomass
params <- params |>
    calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|>
    reefSteady()|>
    calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|>
    reefSteady()|>
    calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|>
    reefSteady()

# Check biomass match
plotBiomassVsSpecies(params) # spot on

# Check match with observed age at maturity
age_mat_observed = karpata_10plus$age_mat
age_mat_model = age_mat(params)
data.frame(age_mat_model, age_mat_observed)
# Closer than needed

# Check predation mortality, feeding levels, and diets
plotPredMort(params) + facet_wrap(~Species)
plotFeedingLevel(params)
plotDiet(params) + scale_x_log10(limits = c(1, 1e4))
plotSpectra(params, biomass = TRUE)

## Now switch to competitive method ----------------------------
params <- newRefuge(params,
                    new_method = "competitive",
                    new_method_params = karpata_refuge)

# Match biomasses again
params <- params |>
    matchBiomasses()|> reefSteady()|>
    matchBiomasses()|> reefSteady()|>
    matchBiomasses()|> reefSteady()|>
    matchBiomasses()|> reefSteady()

# Make sure new refuge is in place
plotVulnerable(params)

plotBiomassVsSpecies(params) # spot on

# Check match with observed age at maturity
age_mat_observed = karpata_10plus$age_mat
age_mat_model = age_mat(params)
data.frame(age_mat_model, age_mat_observed)
# Still look good

## Check resulting spectra and tune resources------------------------

# Resource looks low - should match sheldon's spectrum
# looks fairly straight not bad but some bumps
plotSpectra(params, total = TRUE, biomass = TRUE)
plotSpectra(params, total = TRUE, biomass = TRUE, per_log_size = TRUE)

# plot feeding level to check if resource is too low
plotFeedingLevel(params, species = "inverts")

# Invert feeding level is relatively stable through life, non-linearities
#   are probably due to refuge

# Tune reproduction -----------------------------------------------
# We do not have yield or catch data - can't tune size distribution
# First attempt to set very low to see what the minimum values are
params <- setBevertonHolt(params, erepro = 0.0001)
# Now set setting erepro same for all species, as low as possible
params <- setBevertonHolt(params, erepro = 0.35)
# Project back to steady
params <- reefSteady(params)
# Check reproduction level (value between 0 and 1) - should be higher for
# larger, slow growing species and low for small, fast growing ones
getReproductionLevel(params)
# A reproduction level closer to one means reproduction rate is
# almost totally independent of the investment into reproduction
# These are near one for all species except farming damsels

# Check comparison of density dependent & independent reduction
getRDI(params) / getRDD(params)
# Reproduction is density independent for almost all species

# Let's increase reproduction level back to 0.5 so there is still some
# density dependence
params <- setBevertonHolt(params, reproduction_level = rdi)

# Iterate to get back to steady state
params <- params |>
    reefSteady()|>
    reefSteady()|>
    reefSteady()

# Check new reproduction - these look better
rep <- getReproductionLevel(params)
getRDI(params) / getRDD(params)

# Check new spectra & plots
plotSpectra(params, total = TRUE, biomass = TRUE, per_log_size = TRUE)
plotPredMort(params) + facet_wrap(~Species)
plotFeedingLevel(params)
plotDiet(params) + scale_x_log10(limits = c(1, 1e4))
plotSpectra(params, biomass = TRUE)

# Rescale algae/detritus to a realistic absolute scale, now that diet,
# biomass and growth are tuned -- see vignette("steady-state-recipe")'s
# final step for the literature targets and why this belongs last.
params <- rescale_algae(params, target_biomass / algae_biomass(params))
detritus_lifetime(params) <- target_lifetime

# Save!
caribbean_10_model <- reefSteady(params)
```

</details>
