---
title: "MizerReef Degradation Simulations"
author: ""
bibliography: references.bib
output:
  bookdown::html_document2:
    toc: true
    code_folding: show
    toc_depth: 3
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{MizerReef Degradation Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

# Introduction

This vignette demonstrates how to set up and run MizerReef simulations with 
habitat degradation. Degradation simulations allow you to model the effects 
of disturbances like coral bleaching events on reef fish communities by 
reducing refuge availability over time and optionally adjusting algal 
dynamics.

MizerReef includes functionality to simulate gradual habitat degradation 
following an acute disturbance, with flexible parameterization for different 
recovery trajectories and algae dynamics responses.

# Prerequisites

Before running degradation simulations, you need:

1. A tuned steady-state MizerReef model
2. A degradation trajectory (scaling matrix)
3. Understanding of your desired simulation timeframe

```{r load-packages, message=FALSE, warning=FALSE, eval=FALSE}
library(mizerReef)
library(mizer)
library(ggplot2)
library(dplyr)
library(knitr)

# Load example model and degradation data
data("karpata_model")
data("rubble_scale")
data("algae_scale") 
data("recovery_scale")
```

# Understanding Degradation Parameters

## Degradation Trajectories

MizerReef includes three pre-defined degradation trajectories based on 
different post-bleaching scenarios:

```{r examine-trajectories, eval=FALSE}
# Examine the structure of degradation scaling matrices
cat("Rubble scale dimensions:", dim(rubble_scale), "\n")
cat("Algae scale dimensions:", dim(algae_scale), "\n")
cat("Recovery scale dimensions:", dim(recovery_scale), "\n")

# Show first few columns of each trajectory
kable(rubble_scale[, 1:5], 
      caption = "First 5 years of rubble trajectory (refuge scaling factors)",
      digits = 3)

kable(algae_scale[, 1:5], 
      caption = "First 5 years of algae trajectory (refuge scaling factors)",
      digits = 3)

kable(recovery_scale[, 1:5], 
      caption = "First 5 years of recovery trajectory (refuge scaling factors)",
      digits = 3)
```

**Trajectory interpretation:**
- **Rubble**: Severe degradation with little recovery
- **Algae**: Moderate degradation with algal dominance  
- **Recovery**: Initial degradation followed by habitat recovery

Each matrix has:
- **Rows**: Refuge size bins (matching your competitive refuge profile)
- **Columns**: Time steps (Column 1 = bleaching year, 2+ = recovery years)
- **Values**: Scaling factors applied to previous year's refuge density

## Algae Dynamics Adjustments

When algae_boost = TRUE, you can simulate the response of algal resources 
to habitat degradation:

- **algae_growth_boost**: Multipliers for algae growth rate
- **algae_capacity_boost**: Multipliers for algae carrying capacity

```{r algae-params-example, eval=FALSE}
# Default algae adjustment parameters
default_growth_boost <- c(1.11, 1.11, 1.11, 1.11)
default_capacity_boost <- c(2.0)

cat("Default growth boost (4 years):", default_growth_boost, "\n")
cat("Default capacity boost (1 year):", default_capacity_boost, "\n")
```

# Setting Up Basic Degradation Simulation

## Step 1: Prepare the Model for Degradation

First, set up your tuned model with degradation parameters using 
setDegradation():

```{r setup-degradation, eval=FALSE}
# Start with a tuned steady-state model
params_degraded <- karpata_model

# Add degradation parameters - rubble trajectory example
params_degraded <- setDegradation(
  params_degraded,
  trajectory = "rubble",           # Trajectory name (for documentation)
  deg_scale = rubble_scale,        # Degradation scaling matrix
  bleach_time = 2,                 # Year when bleaching occurs
  algae_boost = FALSE              # No algae adjustments initially
)

# Check that degradation parameters were stored
cat("Degradation enabled:", params_degraded@refuge_params$degrade, "\n")
cat("Bleaching time:", params_degraded@refuge_params$t_bleach, "\n")
cat("Trajectory:", params_degraded@refuge_params$trajectory, "\n")
```

## Step 2: Add Carrying Capacity Dynamics

For realistic simulations, add carrying capacity dynamics to unstructured 
resources using setURcapacity():

```{r add-carrying-capacity, eval=FALSE}
# Add carrying capacity dynamics
params_degraded <- setURcapacity(params_degraded, cap = 1.5)

# Re-equilibrate the model
params_degraded <- reefSteady(params_degraded)

cat("Carrying capacity dynamics enabled:", 
    params_degraded@other_params$use_UR_cc, "\n")
```

## Step 3: Run the Simulation

Now run a projection to see the effects of degradation:

```{r basic-simulation, eval=FALSE}
# Run 20-year simulation 
sim_rubble <- project(params_degraded, t_max = 20, dt = 0.1)

# Plot total biomass over time
plotBiomass(sim_rubble, total = TRUE) +
  ggtitle("Total fish biomass with rubble trajectory") +
  geom_vline(xintercept = 2, color = "red", linetype = "dashed",
             alpha = 0.7) +
  annotate("text", x = 2.2, y = max(getBiomass(sim_rubble)), 
           label = "Bleaching", color = "red", hjust = 0)
```

```{r species-biomass, eval=FALSE}
# Plot species-specific biomass
plotBiomass(sim_rubble) +
  ggtitle("Species biomass with rubble trajectory") +
  geom_vline(xintercept = 2, color = "red", linetype = "dashed",
             alpha = 0.7)
```

# Comparing Different Trajectories

Let's compare the three different degradation trajectories:

```{r compare-trajectories, eval=FALSE}
# Set up models for each trajectory
params_algae <- setDegradation(
  karpata_model,
  trajectory = "algae",
  deg_scale = algae_scale,
  bleach_time = 2,
  algae_boost = FALSE
) %>% setURcapacity() %>% reefSteady()

params_recovery <- setDegradation(
  karpata_model,
  trajectory = "recovery", 
  deg_scale = recovery_scale,
  bleach_time = 2,
  algae_boost = FALSE
) %>% setURcapacity() %>% reefSteady()

# Run simulations
sim_algae <- project(params_algae, t_max = 20, dt = 0.1)
sim_recovery <- project(params_recovery, t_max = 20, dt = 0.1)

# Extract total biomass data for comparison
biomass_rubble <- plotBiomass(sim_rubble, total = TRUE, return_data = TRUE)
biomass_algae <- plotBiomass(sim_algae, total = TRUE, return_data = TRUE)
biomass_recovery <- plotBiomass(sim_recovery, total = TRUE, return_data = TRUE)

# Combine data
biomass_comparison <- rbind(
  biomass_rubble %>% mutate(Trajectory = "Rubble"),
  biomass_algae %>% mutate(Trajectory = "Algae"),
  biomass_recovery %>% mutate(Trajectory = "Recovery")
)

# Plot comparison
ggplot(biomass_comparison, aes(x = Year, y = Biomass, color = Trajectory)) +
  geom_line(size = 1.2) +
  geom_vline(xintercept = 2, color = "red", linetype = "dashed", 
             alpha = 0.7) +
  labs(title = "Comparison of degradation trajectories",
       subtitle = "Total fish community biomass over time",
       y = "Total biomass (g/m²)",
       x = "Year") +
  theme_minimal() +
  annotate("text", x = 2.2, y = max(biomass_comparison$Biomass), 
           label = "Bleaching", color = "red", hjust = 0, size = 3)
```

# Adding Algae Dynamics

Now let's explore how algae dynamics affect the simulation by enabling 
algae_boost:

```{r algae-boost-simulation, eval=FALSE}
# Set up model with algae boost enabled
params_algae_boost <- setDegradation(
  karpata_model,
  trajectory = "rubble",
  deg_scale = rubble_scale,
  bleach_time = 2,
  algae_boost = TRUE,                           # Enable algae adjustments
  algae_growth_boost = c(1.5, 1.3, 1.1, 1.0), # 4 years of growth boost
  algae_capacity_boost = c(3.0, 2.0)           # 2 years of capacity boost
) %>% setURcapacity() %>% reefSteady()

# Run simulation
sim_algae_boost <- project(params_algae_boost, t_max = 20, dt = 0.1)

# Compare with and without algae boost
biomass_no_boost <- plotBiomass(sim_rubble, total = TRUE, return_data = TRUE) %>%
  mutate(Algae_Boost = "No boost")
biomass_with_boost <- plotBiomass(sim_algae_boost, total = TRUE, 
                                  return_data = TRUE) %>%
  mutate(Algae_Boost = "With boost")

biomass_algae_comparison <- rbind(biomass_no_boost, biomass_with_boost)

ggplot(biomass_algae_comparison, aes(x = Year, y = Biomass, 
                                     color = Algae_Boost)) +
  geom_line(size = 1.2) +
  geom_vline(xintercept = 2, color = "red", linetype = "dashed", 
             alpha = 0.7) +
  labs(title = "Effect of algae boost on degradation simulation",
       subtitle = "Total fish biomass with and without algae dynamics boost",
       y = "Total biomass (g/m²)",
       x = "Year",
       color = "Algae dynamics") +
  theme_minimal()
```

Let's also examine how algae boost affects herbivorous species specifically:

```{r herbivore-response, eval=FALSE}
# Extract herbivore biomass (assuming 'parrotfish' are primary herbivores)
herbivore_no_boost <- getBiomass(sim_rubble)[, "parrotfish"]
herbivore_with_boost <- getBiomass(sim_algae_boost)[, "parrotfish"]

herbivore_data <- data.frame(
  Year = rep(as.numeric(dimnames(getBiomass(sim_rubble))$time), 2),
  Biomass = c(herbivore_no_boost, herbivore_with_boost),
  Algae_Boost = rep(c("No boost", "With boost"), 
                    each = length(herbivore_no_boost))
)

ggplot(herbivore_data, aes(x = Year, y = Biomass, color = Algae_Boost)) +
  geom_line(size = 1.2) +
  geom_vline(xintercept = 2, color = "red", linetype = "dashed", 
             alpha = 0.7) +
  labs(title = "Herbivore response to algae boost",
       subtitle = "Parrotfish biomass with and without algae dynamics boost",
       y = "Parrotfish biomass (g/m²)",
       x = "Year",
       color = "Algae dynamics") +
  theme_minimal()
```

# Custom Degradation Trajectories

You can create custom degradation trajectories by building your own scaling 
matrix:

```{r custom-trajectory, eval = FALSE}
# Create a custom "slow recovery" trajectory
# Start with severe degradation, then gradual recovery
n_bins <- nrow(rubble_scale)  # Number of refuge size bins
n_years <- 15                 # Simulation years after bleaching

# Initialize matrix
custom_scale <- matrix(1.0, nrow = n_bins, ncol = n_years)

# Year 1 (bleaching): Severe impact (different for each size class)
custom_scale[, 1] <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7)

# Years 2-5: Continued degradation
for (year in 2:5) {
  custom_scale[, year] <- 0.95  # 5% annual decline
}

# Years 6-15: Gradual recovery
recovery_rate <- seq(0.98, 1.02, length.out = 10)  # Gradual improvement
for (year in 6:15) {
  custom_scale[, year] <- recovery_rate[year - 5]
}

# Show the custom trajectory
kable(custom_scale[, 1:8], 
      caption = "Custom slow recovery trajectory (first 8 years)",
      digits = 3)

# Set up and run simulation with custom trajectory
params_custom <- setDegradation(
  karpata_model,
  trajectory = "slow_recovery",
  deg_scale = custom_scale,
  bleach_time = 2,
  algae_boost = FALSE
) %>% setURcapacity() %>% reefSteady()

sim_custom <- project(params_custom, t_max = 20, dt = 0.1)

# Plot custom trajectory results
plotBiomass(sim_custom, total = TRUE) +
  ggtitle("Custom slow recovery trajectory") +
  geom_vline(xintercept = 2, color = "red", linetype = "dashed", alpha = 0.7)
```

# Analyzing Simulation Results

## Refuge Dynamics Over Time

Let's examine how refuge availability changes during the simulation:

```{r refuge-dynamics, eval=FALSE}
# Extract refuge densities at different time points
time_points <- c(0, 2, 5, 10, 15)  # Before bleaching, at bleaching, recovery

# Function to get refuge density at a specific time
get_refuge_at_time <- function(sim, time_point) {
  # This would need to be implemented based on how refuge data is stored
  # For now, we'll create a conceptual example
  params_at_time <- sim@params
  # In practice, you'd extract refuge density from simulation results
  return(data.frame(
    Size_bin = 1:nrow(rubble_scale),
    Refuge_density = runif(nrow(rubble_scale), 0.1, 1.0),
    Time = time_point
  ))
}

# Create example data showing refuge decline and potential recovery
refuge_data <- do.call(rbind, lapply(time_points, function(t) {
  base_density <- exp(-((1:7 - 4)^2) / 8)  # Bell curve
  
  if (t < 2) {
    # Before bleaching
    density_multiplier <- 1.0
  } else if (t == 2) {
    # At bleaching
    density_multiplier <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7)
  } else {
    # After bleaching - depends on trajectory
    years_since <- t - 2
    density_multiplier <- pmax(0.1, 0.5 + 0.05 * years_since)
  }
  
  data.frame(
    Size_bin = 1:7,
    Refuge_density = base_density * density_multiplier,
    Time = paste("Year", t),
    Year_numeric = t
  )
}))

ggplot(refuge_data, aes(x = Size_bin, y = Refuge_density, 
                        color = factor(Year_numeric))) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Refuge density changes over time",
       subtitle = "Example showing degradation and partial recovery",
       x = "Refuge size bin",
       y = "Refuge density (no./m²)",
       color = "Year") +
  theme_minimal()
```

## Species-Specific Responses

Examine which species are most affected by degradation:

```{r species-responses, eval=FALSE}
# Calculate proportional change in biomass from pre- to post-bleaching
initial_biomass <- getBiomass(sim_rubble)[1, ]  # Year 0
final_biomass <- getBiomass(sim_rubble)[nrow(getBiomass(sim_rubble)), ]  # Final

biomass_change <- data.frame(
  Species = names(initial_biomass),
  Initial = as.numeric(initial_biomass),
  Final = as.numeric(final_biomass),
  Change_percent = ((as.numeric(final_biomass) - as.numeric(initial_biomass)) / 
                    as.numeric(initial_biomass)) * 100
)

# Plot species vulnerability
ggplot(biomass_change, aes(x = reorder(Species, Change_percent), 
                           y = Change_percent,
                           fill = Change_percent < 0)) +
  geom_col() +
  coord_flip() +
  labs(title = "Species vulnerability to habitat degradation",
       subtitle = "Percent change in biomass from year 0 to year 20",
       x = "Species",
       y = "Biomass change (%)",
       fill = "Declined") +
  theme_minimal() +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue"),
                    guide = "none")

# Show the data
kable(biomass_change, 
      caption = "Biomass changes by species",
      digits = 2)
```

# Best Practices and Tips

## Simulation Setup

1. **Always start with a well-tuned steady state** before adding degradation
2. **Use carrying capacity dynamics** (setURcapacity()) for realistic 
   resource responses
3. **Choose appropriate time steps** (dt) - smaller for complex dynamics
4. **Plan your simulation duration** to capture both immediate and long-term 
   effects

## Parameter Selection

```{r parameter-tips, eval=FALSE}
# Example of systematic parameter exploration
bleach_times <- c(1, 2, 5)  # Different timing of disturbance
algae_growth_levels <- list(
  "Low" = c(1.1, 1.05, 1.0),
  "Medium" = c(1.3, 1.2, 1.1),
  "High" = c(1.5, 1.4, 1.2)
)

# Demonstration of parameter space
params_demo <- expand.grid(
  Bleach_time = bleach_times,
  Algae_boost_level = names(algae_growth_levels),
  Trajectory = c("rubble", "algae", "recovery")
)

kable(head(params_demo), 
      caption = "Example parameter combinations for systematic exploration")
```

## Interpreting Results

- **Immediate effects** (years 0-3): Direct impact of habitat loss
- **Medium-term dynamics** (years 3-8): Population responses and potential 
  compensatory growth
- **Long-term outcomes** (years 8+): Community restructuring and new equilibria

# Complete Example Workflow

Here's a complete workflow for setting up and analyzing a degradation 
simulation:

```{r complete-workflow, eval=FALSE}
# 1. Start with tuned model
params_base <- karpata_model

# 2. Set up degradation simulation
params_sim <- setDegradation(
  params_base,
  trajectory = "rubble",
  deg_scale = rubble_scale,
  bleach_time = 2,
  algae_boost = TRUE,
  algae_growth_boost = c(1.2, 1.1, 1.05, 1.0),
  algae_capacity_boost = c(2.5, 1.5)
)

# 3. Add carrying capacity dynamics
params_sim <- setURcapacity(params_sim, cap = 1.5)

# 4. Re-equilibrate
params_sim <- reefSteady(params_sim)

# 5. Run simulation
sim_result <- project(params_sim, t_max = 25, dt = 0.1)

# 6. Analyze results
# Total biomass
plot_total <- plotBiomass(sim_result, total = TRUE)

# Species biomass
plot_species <- plotBiomass(sim_result)

# Productivity
plot_productivity <- plotProductivity(sim_result)

# Diet composition changes
plot_diet <- plotDiet(sim_result, time_range = c(0, 5, 10, 20))

# 7. Export results
write.csv(getBiomass(sim_result), "simulation_biomass.csv")
```

# Further Reading

- [Getting started with MizerReef](mizerReef.html) - Basic model setup
- [MizerReef Steady State Recipe](steady-state-recipe.html) - Model tuning
- [Karpata Model Description](karpata_model-description.html) - Example model
- `setDegradation()` function documentation for parameter details
- `reefDegrade()` function documentation for degradation mechanics

```{r session-info, echo=FALSE}
sessioninfo::session_info()
```