---
title: "Running MizerReef Simulations"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running MizerReef Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r plot-sizing, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4, dpi=300)
```

## Overview

This vignette picks up once you have a tuned `MizerReef` model at steady
state -- see `vignette("mizerReef")` for how to build and tune one. It
covers two things you typically want to do with a tuned model: changing the
refuge profile to explore how habitat structure affects dynamics, and
projecting a model forward in time with fishing pressure or a changing
habitat.

```{r loading-packages, warning = FALSE, message=FALSE}
library(mizer)
library(mizerExperimental)
library(mizerReef)
```

## Changing the refuge profile {.section-change-refuge}

::: {.content-visible when-format="html"}
[Skip to Running a simulation](#running-a-simulation)
:::

Changing the refuge profile allows you to explore how habitat structure affects model dynamics, such as biomass and productivity. This is useful for simulating habitat degradation or restoration scenarios.

**Workflow:**

1. Use `newRefuge()` to change the refuge profile in your model.
2. Run `reefSteady()` several times to reach a new steady state.
3. Compare results (e.g., biomass and productivity) using built-in plotting functions.

```{r change-refuge, echo = TRUE, eval=TRUE, warning=FALSE, message=FALSE, fig.cap="Figure 3. The biomass (left) and productivity (right) for a model with no predation refuge (non-complex) and a model with predation refuge based on data from Karpata Reef in Bonaire (complex). Colours represent species groups.", fig.alt="Two-panel plot: left panel shows total biomass for flat vs complex reef, right panel shows productivity for flat vs complex reef. Each panel compares model output for two habitat types.", out.width="80%"}
# Change to a non-complex (no refuge) profile
non_complex <- newRefuge(caribbean_10_model, new_method = "noncomplex")

# Run to steady state
non_complex <- non_complex |> reefSteady() |> reefSteady() |> reefSteady()

# Compare biomass and productivity between models. Invertebrates aren't
# included in the productivity calculation, so only plot2TotalBiomass()'s
# legend has the complete set of species - keep that one and drop the
# other, rather than collecting both (which would duplicate the legend
# since the two plots' fill scales don't have identical levels).
all_biom11 <- plot2TotalBiomass(non_complex, caribbean_10_model,
                                name1 = "Flat",
                                name2 = "Complex",
                                stack = TRUE) +
    ggplot2::theme_bw() +
    ggplot2::guides(alpha = "none") +
    ggplot2::theme(legend.position = "bottom")

all_prod11 <- plot2Productivity(non_complex, caribbean_10_model,
                                name1 = "Flat",
                                name2 = "Complex",
                                stack = TRUE) +
    ggplot2::theme_bw() +
    ggplot2::guides(alpha = "none") +
    ggplot2::theme(legend.position = "none")

patchwork::wrap_plots(all_biom11, all_prod11)
```

As we can see in Figure 3, removing refuge availability reduces total biomass substantially. Total productivity, by contrast, is only modestly affected here -- refuge changes *which* species contribute most to production more than it changes the total. Biomass and productivity can decouple like this because refuge protects juveniles from predation, letting populations skew toward larger, slower-growing individuals: standing biomass goes up, but production per unit biomass goes down.

> <details><summary>💡 **Tip:** If your results look odd after changing the refuge profile, run `reefSteady()`.</summary> It needs to run enough times to reach a new steady state. To learn more about modifying refuge profiles and their parameters, see [setRefuge()].</details>


For more information on the example data used in this vignette, see the [mizerReef model description vignette](karpata_model-description.html).


## Running a simulation {.section-running-simulation}

::: {.content-visible when-format="html"}
[Skip to Links and further reading](#links-and-further-reading)
:::

Once you have a tuned model at steady state, you can project it forward in
time with mizer's `project()` function, exactly as you would for a standard
mizer model. This section shows two common scenarios: adding fishing
pressure, and letting a parameter (here, the refuge profile) change part-way
through a simulation.

### Simulating fishing pressure

`caribbean_10_model` already has gear parameters set up, so you can project
it forward at a chosen fishing effort. Comparing an unfished run (`effort =
0`) against a fished run shows the effect of fishing on total biomass:

```{r fishing-sim, echo=TRUE, eval=TRUE}
sim_unfished <- project(caribbean_10_model, effort = 0, t_max = 20,
                        progress_bar = FALSE)
sim_fished <- project(caribbean_10_model, effort = 1, t_max = 20,
                      progress_bar = FALSE)

# Total biomass after 20 years, unfished vs. fished
c(unfished = sum(mizer::getBiomass(sim_unfished)[21, ]),
  fished = sum(mizer::getBiomass(sim_fished)[21, ]))
```

```{r plot-fished-biomass, echo=TRUE, eval=TRUE, fig.cap="Figure 4. Total biomass over time for the fished simulation.", fig.alt="Line plot of total biomass by species over 20 years of simulation with fishing effort 1."}
plotBiomass(sim_fished) +
    ggplot2::theme_bw()
```

Yield (the catch taken by the fishery) can be plotted the same way:

```{r plot-fished-yield, echo=TRUE, eval=TRUE, fig.cap="Yield over time for the fished simulation.", fig.alt="Line plot of fishing yield by species over 20 years of simulation with fishing effort 1."}
plotYield(sim_fished) +
    ggplot2::theme_bw()
```

See mizer's own [effort and fishing mortality
articles](https://sizespectrum.org/mizer/articles/mizer.html) for more on
setting up gears, selectivity, and effort schedules.

### Simulating habitat decline

Refuge parameters can also be changed between projection steps, so a
simulation can represent a habitat that is gradually declining (or
recovering) rather than staying fixed. The example below uses the
sigmoidal method (the simplest of the three refuge methods) purely to
illustrate the *mechanism*: shrink the refuge threshold length and the
maximum protected proportion a little each year for five years, projecting
one year at a time and carrying the model's state forward with
`mizer::finalParams()`. The refuge is then left at its final, most-degraded
setting for ten more years so the community has time to settle into a new
steady state.

```{r degradation-mechanism, echo=TRUE, eval=TRUE}
params <- newRefuge(caribbean_10_model, new_method = "sigmoidal",
                    new_method_params = list(L_refuge = 10, prop_protect = 0.8))
sim <- project(params, t_max = 1, progress_bar = FALSE)
params <- mizer::finalParams(sim)
params_yr1 <- params

# Refuge threshold length and maximum protection shrinking over 5 years,
# then held fixed for 10 more years to let the community re-equilibrate
L_seq <- c(10, 8, 6, 4, 2)
prop_seq <- c(0.8, 0.6, 0.4, 0.2, 0.05)
n_years <- length(L_seq) + 10

biomass_trend <- numeric(n_years)
productivity_trend <- numeric(n_years)
biomass_trend[1] <- sum(mizer::getBiomass(params_yr1))
productivity_trend[1] <- sum(getProductivity(params_yr1))

for (i in 2:n_years) {
    if (i <= length(L_seq)) {
        params <- newRefuge(params, new_method = "sigmoidal",
                            new_L_refuge = L_seq[i], new_prop_protect = prop_seq[i])
    }
    sim <- project(params, t_max = 1, progress_bar = FALSE)
    params <- mizer::finalParams(sim)
    biomass_trend[i] <- sum(mizer::getBiomass(sim)[dim(sim@n)[1], ])
    productivity_trend[i] <- sum(getProductivity(params))
}
params_yr15 <- params
```

```{r plot-degradation-trend, echo=TRUE, eval=TRUE, fig.cap="Total biomass and productivity over fifteen years as refuge shrinks then holds at its final degraded state, each shown as a percentage of its year-1 value so the two can share one axis.", fig.alt="Line plot of total biomass and productivity over fifteen years as the refuge threshold length and maximum protected proportion shrink then hold, ending at different levels relative to their starting values."}
# Normalise to year 1 so biomass and productivity (different units) can
# share one y-axis and be compared directly on the same plot.
trend_data <- data.frame(
    year = rep(seq_len(n_years), 2),
    pct = c(biomass_trend / biomass_trend[1] * 100,
           productivity_trend / productivity_trend[1] * 100),
    metric = rep(c("Biomass", "Productivity"), each = n_years)
)

ggplot2::ggplot(trend_data, ggplot2::aes(x = year, y = pct, color = metric,
                                         shape = metric, linetype = metric)) +
    ggplot2::geom_line() +
    ggplot2::geom_point(size = 2) +
    ggplot2::scale_color_manual(values = c(Biomass = "#1B9E77", Productivity = "#D95F02")) +
    ggplot2::labs(x = "Year", y = "% of year-1 value", color = NULL, shape = NULL, linetype = NULL) +
    ggplot2::theme_bw()
```

Biomass ends up lower than where it started (about 90% of its year-1 value),
but productivity actually settles at a *higher* new steady state (about
125%). Aggregate totals like these can hide a lot of detail, though - looking
at individual species groups tells a very different story:

```{r plot-species-collapse, echo=TRUE, eval=TRUE, fig.cap="Species-level biomass before (year 1) and after (year 15) refuge loss, log scale. Several groups collapse to near zero.", fig.alt="Grouped bar chart of biomass by species group, comparing year 1 and year 15, on a log scale. Several bars for year 15 are clipped at the floor of the axis, indicating functional extinction."}
species_names <- species_params(caribbean_10_model)$species
species_comparison <- data.frame(
    species = rep(species_names, 2),
    year = rep(c("Year 1", "Year 15"), each = length(species_names)),
    biomass = c(as.numeric(mizer::getBiomass(params_yr1)),
               as.numeric(mizer::getBiomass(params_yr15)))
)

ggplot2::ggplot(species_comparison, ggplot2::aes(x = species, y = biomass, fill = year)) +
    ggplot2::geom_col(position = "dodge") +
    ggplot2::scale_y_log10(limits = c(1e-3, NA), oob = scales::squish) +
    ggplot2::labs(x = "Species Group", y = expression("Biomass (g/m"^2*", log scale)"), fill = NULL) +
    ggplot2::theme_bw() +
    ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

Four of the ten species groups - Engulfers, Eels, Nocturnal Invertivores and
Planktivores, all species that rely on refuge for protection - collapse to
functionally zero biomass. Parrotfish (the dominant group by biomass) barely
changes, and Invertebrates increase substantially, which is why the
*aggregate* biomass and productivity trends above look like a moderate
decline rather than a multi-species collapse. `plotRelativeContribution()`
makes the same point from a different angle, comparing how much each species
group contributes to abundance, biomass, and productivity before and after:

```{r plot-relative-contribution, echo=TRUE, eval=TRUE, fig.cap="Relative contribution of each species group to abundance, biomass, and productivity, comparing year 1 (with refuge) and year 15 (refuge lost).", fig.alt="Two side-by-side stacked bar charts comparing the relative contribution of each species group to abundance, biomass and productivity, one for year 1 and one for year 15. The abundance panel shows the Planktivores' share present in year 1 has vanished by year 15."}
patchwork::wrap_plots(
    plotRelativeContribution(params_yr1) + ggplot2::ggtitle("Year 1 (with refuge)") + ggplot2::theme_bw(),
    plotRelativeContribution(params_yr15) + ggplot2::ggtitle("Year 15 (refuge lost)") + ggplot2::theme_bw()
) +
  patchwork::plot_layout(guides = "collect") &
  ggplot2::theme(
    legend.position = "bottom",
    legend.title = ggplot2::element_blank(),
    legend.text = ggplot2::element_text(size = 8),
    legend.key.height = grid::unit(0.35, "cm"),
    legend.key.width = grid::unit(0.6, "cm"),
    legend.spacing.x = grid::unit(0.1, "cm")
  )
```

The Planktivores' share of total abundance in year 1 (the blue band) is
entirely gone by year 15 - direct visual confirmation of the collapse shown
in the previous figure, this time in terms of each group's relative share
rather than its absolute biomass.

> <details><summary>💡 **Tip:** This is a mechanism demo, not a calibrated
> analysis.</summary> Switching refuge methods abruptly (as done here) skips
> the tuning workflow described in [Tuning the steady
> state](mizerReef.html#tuning-the-steady-state); for a real
> analysis, follow that recipe after switching methods, not just after
> creating the model. `MizerReef` also supports fully specified,
> data-driven habitat-degradation trajectories via [setDegradation()] and
> [reefDegrade()] for the `"competitive"` refuge method — see their
> reference pages for details.</details>


## Links and further reading {.section-links}

**MizerReef documentation and tutorials:**

- [Getting started with MizerReef](mizerReef.html): Building, tuning and exploring your first model -- start here if you haven't already
- [MizerReef documentation](https://cmbeese.github.io/mizerReef/): Main package documentation and reference manual
- [MizerReef model description vignette](karpata_model-description.html): Detailed explanation of model structure and example workflows
- [MizerReef Steady State recipe](steady-state-recipe.html): Step-by-step guide for tuning models to steady state, including the diet and resource-scale tuning covered in `vignette("tuning-diet-composition")`
- [Example models and built-in data](https://cmbeese.github.io/mizerReef/reference/index.html#example-models): Reference for example species, interaction matrices, and refuge profiles

**Plotting and function references:**

- [MizerReef summary plots](https://cmbeese.github.io/mizerReef/reference/index.html#summary-plots): List of available summary and diagnostic plots
- [setRefuge() function documentation](https://cmbeese.github.io/mizerReef/reference/setRefuge.html): Details on modifying refuge profiles and parameters
- [Mizer plotting results reference](https://sizespectrum.org/mizer/reference/index.html#plotting-results): Reference for plotting functions in mizer

**General mizer resources:**

- [Official mizer getting started guide](https://sizespectrum.org/mizer/articles/mizer.html): General introduction to mizer

**Further research:**

- [Modelling Coral Reef Futures: Exploring the role of structural complexity in sustaining ecosystem services (PhD Thesis, VUW)](https://openaccess.wgtn.ac.nz/articles/thesis/Modelling_Coral_Reef_Futures_Exploring_the_role_of_structural_complexity_in_sustaining_ecosystem_services/26421523?file=48064144): In-depth research and context for MizerReef



## Session info {.section-session-info}

```{r session-info, echo = FALSE}
sessionInfo()
```
