---
title: "MizerReef Steady State Recipe"
author: ""
bibliography: references.bib
output:
  bookdown::html_document2:
    toc: true
    code_folding: show
    toc_depth: 3
    fig_caption: true
vignette: >
  %\VignetteIndexEntry{MizerReef Steady State Recipe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

Achieving a steady state that matches observed biomasses and growth rates is 
a critical step in `MizerReef` model development. This vignette provides a 
step-by-step recipe for tuning a `MizerReef` model to reach a satisfactory 
steady state, using the Karpata reef example dataset.

The tuning process follows the 5-step recipe described in the 
[mizer blog](https://blog.mizer.sizespectrum.org/posts/
2021-08-20-a-5-step-recipe-for-tuning-the-model-steady-state/), 
with additional reef-specific steps to adjust unstructured resource parameters 
and refuge dynamics.

# Prerequisites

Before starting, ensure you have:

1. Species parameter data with reef-specific columns
2. An interaction matrix defining predator-prey relationships  
3. Observed biomass data for calibration
4. A suitable refuge profile for initial tuning

```{r load-packages, message=FALSE, warning=FALSE, eval=FALSE}
library(mizerReef)
library(mizer)
library(ggplot2)
library(dplyr)
library(knitr)

# Load example datasets
data("karpata_species")
data("karpata_int")
data("tuning_profile")
data("karpata_refuge")
```

# Step 1: Create Initial Parameters Object

The first step is to create a `MizerReefParams` object using your species 
parameters and interaction matrix. For initial tuning, we use a simple binned 
refuge method rather than the full competitive method.

**Why use a tuning profile first?** The competitive refuge method is 
density-dependent, which makes it difficult to achieve a stable steady state 
before the biomasses are calibrated to observed values. We start with a 
simpler binned profile that approximates the final refuge shape so that 
we can tune abundances first.

```{r create-initial-params, eval=FALSE}
# Create initial parameters object with tuning refuge profile
params <- newReefParams(group_params = karpata_species,
                        interaction = karpata_int,
                        method = "binned",
                        method_params = tuning_profile)

# Display the basic model structure
print(params)
```

Let's examine the species groups and their key parameters:

```{r species-overview, eval=FALSE}
# Show key species parameters
kable(karpata_species[, c("species", "w_max", "w_mat", "refuge_user", 
                          "bad_pred", "biomass_observed")],
      caption = "Key species parameters for the Karpata model")
```

# Step 2: Reduce Density-Dependence of Reproduction

To achieve a stable initial steady state, we reduce the density-dependence 
of reproduction by setting reproduction levels to an intermediate value 
(typically 0.5).

**Why reduce density-dependence?** High density-dependence can cause 
oscillations during the tuning process. By reducing it initially, we create a 
more stable foundation for tuning other parameters. Many complex coastal reef
ecosystems are highly interconnected, so reproduction is unlikely to be
strongly density-dependent.

```{r reduce-reproduction, eval=FALSE}
# Set reproduction levels to 0.5 for all species
rdi <- rep(0.5, nrow(karpata_species))
names(rdi) <- karpata_species$species

params <- setBevertonHolt(params, reproduction_level = rdi)

# Check the reproduction levels
reproduction_levels <- getReproductionLevel(params)
kable(data.frame(Species = names(reproduction_levels), 
                 Reproduction_Level = reproduction_levels),
      caption = "Initial reproduction levels")
```

# Step 3: Reach Initial Steady State

Project the model forward to reach an initial steady state. This may require 
multiple iterations of `reefSteady()`.

```{r initial-steady-state, eval=FALSE}
# Project to steady state - multiple calls ensure convergence
params <- params |>
    reefSteady() |> reefSteady() |> reefSteady() |> 
    reefSteady() |> reefSteady() |> reefSteady()

# Check convergence by plotting spectra
plotSpectra(params, power = 1) + 
  ggtitle("Initial steady state spectra")
```

# Step 4: Calibrate Biomasses and Growth

This is the core tuning step where we iteratively match observed biomasses 
and growth rates.

## Step 4a: Match Observed Biomasses

Use `calibrateReefBiomass()` to adjust reproduction parameters so that 
modeled biomasses match observations.

```{r calibrate-biomass, eval=FALSE}
# Calibrate biomasses to match observations
params <- calibrateReefBiomass(params)
params <- matchBiomasses(params)
params <- reefSteady(params)

# Plot comparison of observed vs modeled biomasses
plotBiomassVsSpecies(params) + 
  ggtitle("Biomass comparison after initial calibration")
```

## Step 4b: Match Growth Rates

Adjust growth-related parameters to match observed growth rates using 
`matchReefGrowth()`.

```{r match-growth, eval=FALSE}
# Match growth rates
params <- matchReefGrowth(params)
params <- reefSteady(params)

# Check age at maturity matches observations
age_mat_observed <- karpata_species$age_mat
age_mat_model <- age_mat(params)

age_comparison <- data.frame(
  Species = karpata_species$species,
  Observed = age_mat_observed,
  Model = age_mat_model,
  Difference = age_mat_model - age_mat_observed
)

kable(age_comparison, 
      caption = "Comparison of observed vs modeled age at maturity",
      digits = 2)
```

## Step 4c: Iterative Refinement

Repeat the biomass and growth matching process to fine-tune the model.

```{r iterative-tuning, eval=FALSE}
# Iterate to refine biomass and growth matches
params <- params |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady()

# Final biomass comparison
plotBiomassVsSpecies(params) + 
  ggtitle("Final biomass comparison after iterative tuning")
```

Let's examine the model performance after tuning:

```{r check-performance, eval=FALSE}
# Check feeding levels
plotFeedingLevel(params) + 
  ggtitle("Feeding levels after tuning")

# Check predation mortality
plotPredMort(params) + 
  facet_wrap(~Species) +
  ggtitle("Predation mortality by species")
```

# Step 5: Switch to Competitive Refuge Method

Once we have a stable steady state with the tuning refuge profile, we can 
switch to the full competitive refuge method, which includes 
density-dependent refuge dynamics.

**Why switch methods?** The competitive method provides more realistic 
refuge dynamics but requires a stable baseline to work from.

```{r switch-refuge-method, eval=FALSE}
# Switch to competitive refuge method
params <- newRefuge(params,
                    new_method = "competitive", 
                    new_method_params = karpata_refuge)

# Check the new refuge profile
plotRefuge(params) + 
  ggtitle("Refuge profile with competitive method")
```

After switching refuge methods, we need to re-match biomasses as the refuge 
dynamics will have changed:

```{r rematch-after-refuge, eval=FALSE}
# Re-match biomasses with new refuge method
params <- params |>
    matchBiomasses() |> reefSteady() |> 
    matchBiomasses() |> reefSteady() |>
    matchBiomasses() |> reefSteady() |>
    matchBiomasses() |> reefSteady()

# Check vulnerability with new method
plotVulnerable(params) + 
  ggtitle("Vulnerability to predation with competitive refuge")

# Final biomass check
plotBiomassVsSpecies(params) + 
  ggtitle("Biomass match with competitive refuge method")
```

# Step 6: Fine-tune Reproduction Parameters

The final step involves tuning reproduction parameters to achieve the 
desired reproduction levels and density-dependence.

```{r tune-reproduction, eval=FALSE}
# Set reproduction efficiency to allow some density dependence
params <- setBevertonHolt(params, erepro = 0.35)
params <- reefSteady(params)

# Check reproduction levels - should be realistic for each species
final_reproduction_levels <- getReproductionLevel(params)
kable(data.frame(Species = names(final_reproduction_levels), 
                 Reproduction_Level = final_reproduction_levels),
      caption = "Final reproduction levels",
      digits = 3)

# Check ratio of density-independent to density-dependent reproduction
rdi_rdd_ratio <- getRDI(params) / getRDD(params)
kable(data.frame(Species = names(rdi_rdd_ratio), 
                 RDI_RDD_Ratio = rdi_rdd_ratio),
      caption = paste0("Ratio of density-independent to ",
                       "density-dependent reproduction"),
      digits = 2)
```

## Final Model Validation

Let's examine the final model performance across multiple metrics:

```{r final-validation, eval=FALSE}
# Overall spectra
plotSpectra(params, total = TRUE, power = 2) + 
  ggtitle("Final total spectra")

# Diet composition
plotDiet(params) + 
  scale_x_log10(limits = c(1, 1e4)) +
  ggtitle("Diet composition by size")

# Growth curves
plotGrowthCurves(params) + 
  ggtitle("Growth curves")
```

# Summary

This recipe provides a systematic approach to tuning `MizerReef` models:

1. **Start simple**: Use a binned refuge method for initial tuning
2. **Reduce complexity**: Lower reproduction density-dependence initially  
3. **Iterative matching**: Repeatedly match biomasses and growth rates
4. **Add complexity**: Switch to competitive refuge method when stable
5. **Final tuning**: Adjust reproduction parameters for realistic dynamics

## Key Functions Used

- `newReefParams()`: Create initial parameters object
- `setBevertonHolt()`: Set reproduction parameters
- `reefSteady()`: Project to steady state
- `calibrateReefBiomass()`: Match observed biomasses
- `matchReefGrowth()`: Match growth rates  
- `newRefuge()`: Switch refuge methods
- Various plotting functions for validation

## Complete Tuning Script

Here's the complete script used for the Karpata model:

```{r complete-script, eval=FALSE}
# Complete tuning script for Karpata model
library(mizerReef)
library(mizer)

# Load data
data("karpata_species")
data("karpata_int") 
data("tuning_profile")
data("karpata_refuge")

# 1. Create initial parameters
params <- newReefParams(group_params = karpata_species,
                        interaction = karpata_int,
                        method = "binned",
                        method_params = tuning_profile)

# 2. Reduce density dependence
rdi <- rep(0.5, nrow(karpata_species))
params <- setBevertonHolt(params, reproduction_level = rdi)

# 3. Initial steady state
params <- params |>
    reefSteady() |> reefSteady() |> reefSteady() |> 
    reefSteady() |> reefSteady() |> reefSteady()

# 4. Calibrate biomasses and growth
params <- calibrateReefBiomass(params)
params <- matchBiomasses(params)
params <- reefSteady(params)
params <- matchReefGrowth(params)
params <- reefSteady(params)

# Iterative refinement
params <- params |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady()

# 5. Switch to competitive refuge
params <- newRefuge(params,
                    new_method = "competitive",
                    new_method_params = karpata_refuge)

# Re-match with new refuge
params <- params |>
    matchBiomasses() |> reefSteady() |> 
    matchBiomasses() |> reefSteady() |>
    matchBiomasses() |> reefSteady() |>
    matchBiomasses() |> reefSteady()

# 6. Fine-tune reproduction
params <- setBevertonHolt(params, erepro = 0.35)
params <- setBevertonHolt(params, reproduction_level = rdi)
params <- params |>
    reefSteady() |> reefSteady() |> reefSteady()

# Save final model
karpata_model <- reefSteady(params)
```

## Further Reading

- [Getting started with MizerReef](mizerReef.html)
- [Karpata Model Description](karpata_model-description.html)
- [Mizer steady state tuning blog](https://blog.mizer.sizespectrum.org/posts/
  2021-08-20-a-5-step-recipe-for-tuning-the-model-steady-state/)

```{r session-info, echo=FALSE}
sessioninfo::session_info()
```
