---
title: "MizerReef Steady State Recipe"
author: ""
bibliography: references.bib
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MizerReef Steady State Recipe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

Achieving a steady state that matches observed biomasses and growth rates is 
a critical step in `MizerReef` model development. This vignette provides a 
step-by-step recipe for tuning a `MizerReef` model to reach a satisfactory 
steady state, using the Karpata reef example dataset.

The tuning process follows the 5-step recipe described in the 
[mizer blog](https://blog.mizer.sizespectrum.org/posts/
2021-08-20-a-5-step-recipe-for-tuning-the-model-steady-state/), 
with additional reef-specific steps to adjust unstructured resource parameters 
and refuge dynamics.

# Prerequisites

Before starting, ensure you have:

1. Species parameter data with reef-specific columns
2. An interaction matrix defining predator-prey relationships  
3. Observed biomass data for calibration
4. A suitable refuge profile for initial tuning

```{r load-packages, message=FALSE, warning=FALSE, eval=TRUE}
library(mizerReef)
library(mizer)
library(mizerExperimental)
library(ggplot2)
library(dplyr)
library(knitr)

# Load example datasets
data("caribbean_10_species")
data("caribbean_10_interaction")
data("tuning_profile")
data("karpata_refuge")
```

# Step 1: Create the Initial Parameters Object

The first step is to create a `MizerReefParams` object using your species 
parameters and interaction matrix. For initial tuning, we use a simple binned 
refuge method rather than the full competitive method.

**Why use a tuning profile first?** The competitive refuge method is
density-dependent, which makes it difficult to achieve a stable steady state
before the biomasses are calibrated to observed values. We start with a
simpler constant binned profile so that
we can tune abundances first.

```{r create-initial-params, eval=TRUE}
# Create initial parameters object with tuning refuge profile
params <- newReefParams(species_params = caribbean_10_species,
                        interaction = caribbean_10_interaction,
                        method = "binned",
                        method_params = tuning_profile)
```

These messages let us know when the package set default values for some
parameters such as the search volume, natural mortality, and maximum
consumption rates. You can change these defaults later if you have better
estimates using `species_params()` or one of the `set*()` functions.

With a full species table, this wall of messages gets long. `newReefParams()`
(and every other model-building function in mizerReef) takes an `info_level`
argument to control it: `info_level = 1` keeps only the reports that say
something went differently from what you actually asked for (dropping the
routine default-value fills), and `info_level = 0` is silent. For example:

```{r create-initial-params-quiet, eval=FALSE}
params <- newReefParams(species_params = caribbean_10_species,
                        interaction = caribbean_10_interaction,
                        method = "binned",
                        method_params = tuning_profile,
                        info_level = 1)
```

Let's examine the species groups and their key parameters:

```{r species-overview, eval=FALSE}
# Show key species parameters
kable(caribbean_10_species[, c("species", "w_max", "w_mat", "refuge_user", 
                          "blocked_pred", "biomass_observed")],
      caption = "Key species parameters for the Karpata model")
```

# Step 2: Find an Initial Steady State

Project the model forward to reach an initial steady state. This may require 
multiple iterations of `reefSteady()`.

```{r initial-steady-state, eval=TRUE}
# Project to steady state - multiple calls ensure convergence
params <- params |>
    reefSteady() |> reefSteady() |> reefSteady() |> 
    reefSteady() |> reefSteady() |> reefSteady()

# Check convergence by plotting spectra
plotSpectra(params, biomass = TRUE) +
  ggtitle("Initial steady state spectra")
```

If you're having trouble reaching a steady state, consider providing the species parameter `R_max` or adjusting the `reproduction_level` like this:

```{r reduce-reproduction, eval=TRUE}
# Set reproduction levels to 0.5 for all species
rdi <- rep(0.5, nrow(caribbean_10_species))
names(rdi) <- caribbean_10_species$species

params <- setBevertonHolt(params, reproduction_level = rdi)

# Check the reproduction levels
reproduction_levels <- getReproductionLevel(params)
kable(data.frame(Species = names(reproduction_levels), 
                 Reproduction_Level = reproduction_levels),
      caption = "Initial reproduction levels")
```

**Why use density-dependence?** Pure density-independence (the mizer 
default) can cause instability in any model, but models with reef-specific non-linearities 
(multiple resources, refuge dynamics) are especially prone to this. Moderate density-dependence provides 
stabilizing feedback without creating the steep response curves that cause 
oscillations. 

<details>
<summary>**Click for detailed explanation of reproduction dynamics**</summary>

## Understanding Reproduction Level and Convergence

`reproduction_level` controls the functional form of the Beverton-Holt
stock-recruitment relationship, from 0 (maximum compensation: recruitment
saturates quickly against spawning biomass) to 1 (mizer's default, no
compensation: recruitment is directly proportional to spawning biomass).

Reef models add several sources of non-linearity beyond a standard mizer
model -- coupled algae/detritus/plankton dynamics, refuge mechanisms whose
effect on predation mortality is itself density-dependent, and the iterative
`calibrateReefBiomass()`/`matchReefGrowth()` loops used to reach them. With
no reproductive compensation (`reproduction_level = 1`), there's nothing to
damp these interacting non-linearities, and `reefSteady()`'s search for an
equilibrium (where it repeatedly adjusts `erepro` to drive births toward
deaths) can fail to converge or oscillate. Moderate compensation
(`reproduction_level = 0.5`) provides just enough stabilizing feedback to
damp this without itself being steep enough to cause the solver to
overshoot, which is what happens as compensation gets stronger still
(`reproduction_level` near 0). This isn't purely a numerical convenience,
either: many reef fish have an open recruitment system, with larvae
dispersing widely and local recruitment only weakly tied to local spawning
stock, which is itself closer to moderate than to zero or maximum
compensation.

**Recommended workflow**: use `reproduction_level = 0.5` throughout
calibration for stable convergence while matching biomasses and growth
rates. Afterwards, you can move it toward 1 if you believe recruitment is
very weakly density-dependent for your system, or toward 0 to test recovery
dynamics from overfishing -- but moderate values are usually both the most
numerically stable and the most biologically defensible default for reef
fish with typical connectivity.

</details>

# Step 3: Calibrate Biomasses and Growth

This is the core tuning step where we iteratively match observed biomasses 
and growth rates.

## Step 4a: Match Observed Biomasses

Use `calibrateReefBiomass()` to adjust reproduction parameters so that 
modeled biomasses match observations.

```{r calibrate-biomass, eval=FALSE}
# Calibrate biomasses to match observations
params <- calibrateReefBiomass(params)
params <- matchBiomasses(params)
params <- reefSteady(params)

# Plot comparison of observed vs modeled biomasses
plotBiomassVsSpecies(params) + 
  ggtitle("Biomass comparison after initial calibration")
```

## Step 4b: Match Growth Rates

Adjust growth-related parameters to match observed growth rates using 
`matchReefGrowth()`.

```{r match-growth, eval=FALSE}
# Match growth rates
params <- matchReefGrowth(params)
params <- reefSteady(params)

# Check age at maturity matches observations
age_mat_observed <- caribbean_10_species$age_mat
age_mat_model <- age_mat(params)

age_comparison <- data.frame(
  Species = caribbean_10_species$species,
  Observed = age_mat_observed,
  Model = age_mat_model,
  Difference = age_mat_model - age_mat_observed
)

kable(age_comparison, 
      caption = "Comparison of observed vs modeled age at maturity",
      digits = 2)
```

## Step 4c: Iterative Refinement

Repeat the biomass and growth matching process to fine-tune the model.

```{r iterative-tuning, eval=FALSE}
# Iterate to refine biomass and growth matches
params <- params |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady()

# Final biomass comparison
plotBiomassVsSpecies(params) + 
  ggtitle("Final biomass comparison after iterative tuning")
```

Let's examine the model performance after tuning:

```{r check-performance, eval=FALSE}
# Check feeding levels
plotFeedingLevel(params) + 
  ggtitle("Feeding levels after tuning")

# Check predation mortality
plotPredMort(params) + 
  facet_wrap(~Species) +
  ggtitle("Predation mortality by species")
```

# Step 5: Switch to Competitive Refuge Method

Once we have a stable steady state with the tuning refuge profile, we can 
switch to the full competitive refuge method, which includes 
density-dependent refuge dynamics.

**Why switch methods?** The competitive method provides more realistic 
refuge dynamics but requires a stable baseline to work from.

```{r switch-refuge-method, eval=TRUE}
# Switch to competitive refuge method
params <- newRefuge(params,
                    new_method = "competitive", 
                    new_method_params = karpata_refuge)

# Check the new refuge profile
plotRefugeProfile(params) +
  ggtitle("Refuge profile with competitive method")
```

After switching refuge methods, we need to re-match biomasses as the refuge 
dynamics will have changed:

```{r rematch-after-refuge, eval=TRUE}
# Re-match biomasses with new refuge method
params <- params |>
    matchBiomasses() |> reefSteady() |> 
    matchBiomasses() |> reefSteady() |>
    matchBiomasses() |> reefSteady() |>
    matchBiomasses() |> reefSteady()

# Check vulnerability with new method
plotVulnerable(params) + 
  ggtitle("Vulnerability to predation with competitive refuge")

# Final biomass check
plotBiomassVsSpecies(params) + 
  ggtitle("Biomass match with competitive refuge method")
```

# Step 6: Fine-tune Reproduction Parameters

The final step involves tuning reproduction parameters to achieve the 
desired reproduction levels and density-dependence.

```{r tune-reproduction, eval=FALSE}
# Set reproduction efficiency to allow some density dependence
params <- setBevertonHolt(params, erepro = 0.35)
params <- reefSteady(params)

# Check reproduction levels - should be realistic for each species
final_reproduction_levels <- getReproductionLevel(params)
kable(data.frame(Species = names(final_reproduction_levels), 
                 Reproduction_Level = final_reproduction_levels),
      caption = "Final reproduction levels",
      digits = 3)

# Check ratio of density-independent to density-dependent reproduction
rdi_rdd_ratio <- getRDI(params) / getRDD(params)
kable(data.frame(Species = names(rdi_rdd_ratio), 
                 RDI_RDD_Ratio = rdi_rdd_ratio),
      caption = paste0("Ratio of density-independent to ",
                       "density-dependent reproduction"),
      digits = 2)
```

# Step 7: Rescale Algae and Detritus to a Realistic Absolute Scale

Steps 3-6 above tune algae and detritus *relative* to the fish sub-model --
`reefSteady()` calls `tuneUR()`/`tuneUR_cc()` internally, which balances
each pool's production against consumption (see
`vignette("model-description")`'s Algae and Detritus sections for exactly
what's held fixed for each). That balance says nothing about the pools'
*absolute* scale, though, and reaching it usually leaves algae and detritus
biomass far from a realistic standing stock: every biomass-correction step
above (`calibrateReefBiomass()` in particular) necessarily rescales the
species' interaction strength with algae/detritus (`rho_algae`/
`rho_detritus`) to preserve feeding level as fish abundances are corrected --
the same principle mizer itself uses for `gamma`/search volume -- and that
repeated rescaling drives the *absolute* biomass scale down over the course
of calibration, even though nothing about it is wrong.

The fix is a final, deliberately non-disruptive correction:
`rescale_algae()`/`rescale_detritus()` multiply a pool's biomass by a factor
and divide the consuming species' interaction strength by the same factor,
so **total consumption -- and therefore the already-tuned fish
sub-model -- is completely unaffected**. `detritus_lifetime() <-` is a
convenience wrapper around `rescale_detritus()` that targets a turnover time
(biomass / consumption rate) instead of a raw factor; there is no equivalent
setter for algae, so target its standing biomass directly with
`rescale_algae()`.

This is why the step belongs at the *end* of the recipe, not earlier:
because it doesn't touch consumption, it doesn't need to be redone after a
biomass or growth match, and running it before one would just be undone by
the next `calibrateReefBiomass()` call.

```{r rescale-algae-detritus, eval=FALSE}
# Detritus: target a literature turnover time directly.
# e.g. Max, Hamilton, Gaines & Warner (2013, Mar. Ecol. Prog. Ser. 482:181-195)
# measured a 2.8-4.9 hour benthic detritus residence time at Palmyra Atoll;
# here we target the midpoint, expressed in years.
detritus_lifetime(params) <- 3.85 / (24 * 365)

# Algae: target a literature standing biomass directly with rescale_algae().
# Grazed Caribbean algal turf standing stock is commonly cited on the order
# of 10-100 g dry weight/m^2 -- convert to whatever mass units your own
# biomass_observed data uses (usually wet weight, via a site-specific
# dry:wet ratio) before using it as `target_algae_biomass` below, rather
# than reusing a number tuned for a different dataset.
params <- rescale_algae(params, target_algae_biomass / algae_biomass(params))

# Neither call requires re-running reefSteady() -- they don't change
# consumption, only the reported biomass and rho -- but it's good practice
# to confirm the biomass match is still exact:
plotBiomassObservedVsModel(params)
```

## Final Model Validation

Let's examine the final model performance across multiple metrics:

```{r final-validation, eval=FALSE}
# Overall spectra
plotSpectra(params, total = TRUE, biomass = TRUE, per_log_size = TRUE) + 
  ggtitle("Final total spectra")

# Diet composition
plotDiet(params) + 
  scale_x_log10(limits = c(1, 1e4)) +
  ggtitle("Diet composition by size")

# Growth curves
plotGrowthCurves(params) + 
  ggtitle("Growth curves")
```

# Summary

This recipe provides a systematic approach to tuning `MizerReef` models:

1. **Start simple**: Use a binned refuge method for initial tuning
2. **Set intermediate reproduction**: Add moderate density-dependence (0.5) for stability
3. **Iterative matching**: Repeatedly match biomasses and growth rates
4. **Add complexity**: Switch to competitive refuge method when stable
5. **Final tuning**: Adjust reproduction parameters for realistic dynamics
6. **Rescale resources**: Correct algae/detritus to a realistic absolute scale, last, since it doesn't disturb anything tuned before it

## Key Functions Used

- `newReefParams()`: Create initial parameters object
- `setBevertonHolt()`: Set reproduction parameters
- `reefSteady()`: Project to steady state
- `calibrateReefBiomass()`: Match observed biomasses
- `matchReefGrowth()`: Match growth rates
- `newRefuge()`: Switch refuge methods
- `rescale_algae()`, `detritus_lifetime()<-`: Correct absolute resource scale without disturbing consumption
- Various plotting functions for validation

## Complete Tuning Script

Here's the complete script used for the Karpata model:

```{r complete-script, eval=FALSE}
# Complete tuning script for Karpata model
library(mizerReef)
library(mizer)

# Load data
data("caribbean_10_species")
data("caribbean_10_interaction") 
data("tuning_profile")
data("karpata_refuge")

# 1. Create initial parameters
params <- newReefParams(species_params = caribbean_10_species,
                        interaction = caribbean_10_interaction,
                        method = "binned",
                        method_params = tuning_profile)

# 2. Set intermediate reproduction level (moderate density-dependence)
rdi <- rep(0.5, nrow(caribbean_10_species))
params <- setBevertonHolt(params, reproduction_level = rdi)

# 3. Initial steady state
params <- params |>
    reefSteady() |> reefSteady() |> reefSteady() |> 
    reefSteady() |> reefSteady() |> reefSteady()

# 4. Calibrate biomasses and growth
params <- calibrateReefBiomass(params)
params <- matchBiomasses(params)
params <- reefSteady(params)
params <- matchReefGrowth(params)
params <- reefSteady(params)

# Iterative refinement
params <- params |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady()

# 5. Switch to competitive refuge
params <- newRefuge(params,
                    new_method = "competitive",
                    new_method_params = karpata_refuge)

# Re-match with new refuge
params <- params |>
    matchBiomasses() |> reefSteady() |> 
    matchBiomasses() |> reefSteady() |>
    matchBiomasses() |> reefSteady() |>
    matchBiomasses() |> reefSteady()

# 6. Fine-tune reproduction
params <- setBevertonHolt(params, erepro = 0.35)
params <- setBevertonHolt(params, reproduction_level = rdi)
params <- params |>
    reefSteady() |> reefSteady() |> reefSteady()

# 7. Rescale algae/detritus to a realistic absolute scale. Does not disturb
# consumption, so no further reefSteady() calls are needed for this step.
detritus_lifetime(params) <- 3.85 / (24 * 365)
params <- rescale_algae(params, target_algae_biomass / algae_biomass(params))

# Save final model
caribbean_10_model <- reefSteady(params)
```

## Further Reading

- [Getting started with MizerReef](mizerReef.html)
- [Karpata Model Description](karpata_model-description.html)
- [Mizer steady state tuning blog](https://blog.mizer.sizespectrum.org/posts/
  2021-08-20-a-5-step-recipe-for-tuning-the-model-steady-state/)

```{r session-info, echo=FALSE}
sessionInfo()
```
