---
title: "MizerReef Steady State Recipe"
author: ""
bibliography: references.bib
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MizerReef Steady State Recipe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

Achieving a steady state that matches observed biomasses and growth rates is 
a critical step in `MizerReef` model development. This vignette provides a 
step-by-step recipe for tuning a `MizerReef` model to reach a satisfactory 
steady state, using the Karpata reef example dataset.

The tuning process follows the 5-step recipe described in the 
[mizer blog](https://blog.mizer.sizespectrum.org/posts/
2021-08-20-a-5-step-recipe-for-tuning-the-model-steady-state/), 
with additional reef-specific steps to adjust unstructured resource parameters 
and refuge dynamics.

# Prerequisites

Before starting, ensure you have:

1. Species parameter data with reef-specific columns
2. An interaction matrix defining predator-prey relationships  
3. Observed biomass data for calibration
4. A suitable refuge profile for initial tuning

```{r load-packages, message=FALSE, warning=FALSE, eval=FALSE}
library(mizerReef)
library(mizer)
library(ggplot2)
library(dplyr)
library(knitr)

# Load example datasets
data("caribbean_10_species")
data("caribbean_10_interaction")
data("tuning_profile")
data("karpata_refuge")
```

# Step 1: Create Initial Parameters Object

The first step is to create a `MizerReefParams` object using your species 
parameters and interaction matrix. For initial tuning, we use a simple binned 
refuge method rather than the full competitive method.

**Why use a tuning profile first?** The competitive refuge method is 
density-dependent, which makes it difficult to achieve a stable steady state 
before the biomasses are calibrated to observed values. We start with a 
simpler binned profile that approximates the final refuge shape so that 
we can tune abundances first.

```{r create-initial-params, eval=FALSE}
# Create initial parameters object with tuning refuge profile
params <- newReefParams(species_params = caribbean_10_species,
                        interaction = caribbean_10_interaction,
                        method = "binned",
                        method_params = tuning_profile)

# Display the basic model structure
print(params)
```

Let's examine the species groups and their key parameters:

```{r species-overview, eval=FALSE}
# Show key species parameters
kable(caribbean_10_species[, c("species", "w_max", "w_mat", "refuge_user", 
                          "blocked_pred", "biomass_observed")],
      caption = "Key species parameters for the Karpata model")
```

# Step 2: Set Intermediate Reproduction Level

To achieve a stable initial steady state, we set reproduction levels to an 
intermediate value (typically 0.5), which adds moderate density-dependence 
compared to the mizer default of pure density-independence.

**Why use moderate density-dependence?** Pure density-independence (the mizer 
default) can cause instability when combined with reef-specific non-linearities 
(multiple resources, refuge dynamics). Moderate density-dependence provides 
stabilizing feedback without creating the steep response curves that cause 
oscillations. This is also biologically realistic for reef systems with open 
recruitment and larval connectivity.

<details>
<summary>**Click for detailed explanation of reproduction dynamics**</summary>

## Understanding Reproduction Level and Convergence

### What does reproduction_level control?

The `reproduction_level` parameter controls the **functional form** of the 
Beverton-Holt stock-recruitment relationship. It ranges from 0 (maximum 
compensation) to 1 (no compensation):

- **reproduction_level = 1** (mizer default): Purely density-**independent**. 
  Recruitment is directly proportional to spawning biomass with no compensatory 
  feedback. Linear response.
  
- **reproduction_level = 0.5** (intermediate): Moderate density-**dependence**. 
  Recruitment saturates somewhat as spawning biomass increases, providing 
  compensatory feedback but not extreme non-linearity.
  
- **reproduction_level = 0** (maximum): Strongly density-**dependent**. 
  Recruitment saturates quickly, creating steep non-linear response curves. 
  Maximum compensation.

### Why does pure density-independence (default) fail in reef models?

During `steady()` or `reefSteady()`, the algorithm iteratively adjusts `erepro` 
(egg production efficiency) to drive each species toward equilibrium where 
births equal deaths. 

**With pure density-independence (reproduction_level = 1, the mizer default):**

1. **No stabilizing feedback**: Biomass fluctuations have no compensatory 
   response in recruitment.
   
2. **Compounded instability**: Reef models add multiple non-linearities (refuge 
   dynamics, multiple resources). Without reproductive compensation to dampen 
   fluctuations, these non-linearities interact chaotically.
   
3. **Failed convergence**: The solver can't find a stable equilibrium because 
   small perturbations propagate without feedback.

**With moderate density-dependence (reproduction_level = 0.5):**

- **Stabilizing feedback**: If biomass drops, recruitment compensates somewhat, 
  preventing runaway collapses.
  
- **Smooth response**: The compensation is gradual, not steep, so the solver 
  can iterate without overshooting.
  
- **Convergence**: The system finds a stable equilibrium where reproductive 
  feedback balances the reef non-linearities.

**With strong density-dependence (reproduction_level near 0):**

- **Too steep**: The compensation curve becomes so non-linear that the solver 
  overshoots targets, creating oscillations again.

### Why are reef models especially sensitive?

MizerReef models add multiple sources of non-linearity:

- **Multiple resources** (algae, detritus, plankton) with their own coupled 
  dynamics
- **Refuge mechanisms** that non-linearly affect predation mortality based on 
  density
- **Iterative calibration loops** (`calibrateReefBiomass`, `matchReefGrowth`) 
  that compound small instabilities

With **no density-dependence** (mizer default), there's no compensatory 
feedback to stabilize these interacting non-linearities. The system has too 
many sources of instability and no damping mechanism.

With **moderate density-dependence** (0.5), reproductive compensation provides 
just enough stabilizing feedback without adding another steep non-linearity. 
This balances the non-linearity budget and allows convergence.

With **strong density-dependence** (near 0), the steep compensation curve 
itself becomes a destabilizing non-linearity, causing overshooting.

### What does reproduction_level do outside of steady state?

During forward projections (`project()`), `reproduction_level` controls how 
recruitment responds to biomass perturbations:

- **Low (high density-dependence)**: Strong compensatory dynamics. Overfished 
  stocks recover faster because recruitment is maintained even at low spawning 
  biomass. Boom-bust cycles are naturally damped.
  
- **High (low density-dependence)**: Recruitment tracks spawning biomass more 
  directly with less compensation. Populations are more vulnerable to collapse 
  but also track environmental drivers more closely.

### Is this biologically realistic for reefs?

For many reef fish species, **moderate density-dependence is often more 
realistic** than either pure density-independence or strong compensation:

- **Larval dispersal**: Reef fish larvae disperse widely in ocean currents. 
  Local recruitment is weakly (but not completely) tied to local spawning stock 
  because recruits come from both local and distant populations (open 
  recruitment system). This creates weak to moderate density-dependence.
  
- **Connectivity**: Reef ecosystems are highly interconnected through larval 
  exchange. Local density-dependent feedbacks exist but are dampened by 
  connectivity, resulting in moderate (not strong) compensation.
  
- **Environmental variability**: Recruitment is driven by both spawning stock 
  and oceanographic conditions. Pure density-independence (default mizer) would 
  ignore biological reality that spawning stock matters *somewhat*.

Therefore, setting `reproduction_level` to 0.5 is not just numerically 
convenient—it often better reflects reef fish biology than either extreme 
(pure density-independence or maximum compensation).

### Recommended workflow

1. **During calibration**: Set `reproduction_level` to 0.5 (moderate 
   density-dependence) to ensure convergence while matching biomasses and 
   growth rates. This provides stabilizing feedback without extreme 
   non-linearity.
   
2. **After calibration**: You can optionally increase `reproduction_level` 
   toward 1 (less compensation) if you believe recruitment is very weakly 
   density-dependent, or decrease it toward 0 (stronger compensation) if you 
   need to test recovery dynamics from overfishing.
   
3. **For reef models**: Keeping `reproduction_level` at moderate values 
   (around 0.5) throughout is usually both numerically stable and biologically 
   defensible for most reef fish with moderate connectivity.

</details>

```{r reduce-reproduction, eval=FALSE}
# Set reproduction levels to 0.5 for all species
rdi <- rep(0.5, nrow(caribbean_10_species))
names(rdi) <- caribbean_10_species$species

params <- setBevertonHolt(params, reproduction_level = rdi)

# Check the reproduction levels
reproduction_levels <- getReproductionLevel(params)
kable(data.frame(Species = names(reproduction_levels), 
                 Reproduction_Level = reproduction_levels),
      caption = "Initial reproduction levels")
```

# Step 3: Reach Initial Steady State

Project the model forward to reach an initial steady state. This may require 
multiple iterations of `reefSteady()`.

```{r initial-steady-state, eval=FALSE}
# Project to steady state - multiple calls ensure convergence
params <- params |>
    reefSteady() |> reefSteady() |> reefSteady() |> 
    reefSteady() |> reefSteady() |> reefSteady()

# Check convergence by plotting spectra
plotSpectra(params, power = 1) + 
  ggtitle("Initial steady state spectra")
```

# Step 4: Calibrate Biomasses and Growth

This is the core tuning step where we iteratively match observed biomasses 
and growth rates.

## Step 4a: Match Observed Biomasses

Use `calibrateReefBiomass()` to adjust reproduction parameters so that 
modeled biomasses match observations.

```{r calibrate-biomass, eval=FALSE}
# Calibrate biomasses to match observations
params <- calibrateReefBiomass(params)
params <- matchBiomasses(params)
params <- reefSteady(params)

# Plot comparison of observed vs modeled biomasses
plotBiomassVsSpecies(params) + 
  ggtitle("Biomass comparison after initial calibration")
```

## Step 4b: Match Growth Rates

Adjust growth-related parameters to match observed growth rates using 
`matchReefGrowth()`.

```{r match-growth, eval=FALSE}
# Match growth rates
params <- matchReefGrowth(params)
params <- reefSteady(params)

# Check age at maturity matches observations
age_mat_observed <- caribbean_10_species$age_mat
age_mat_model <- age_mat(params)

age_comparison <- data.frame(
  Species = caribbean_10_species$species,
  Observed = age_mat_observed,
  Model = age_mat_model,
  Difference = age_mat_model - age_mat_observed
)

kable(age_comparison, 
      caption = "Comparison of observed vs modeled age at maturity",
      digits = 2)
```

## Step 4c: Iterative Refinement

Repeat the biomass and growth matching process to fine-tune the model.

```{r iterative-tuning, eval=FALSE}
# Iterate to refine biomass and growth matches
params <- params |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady()

# Final biomass comparison
plotBiomassVsSpecies(params) + 
  ggtitle("Final biomass comparison after iterative tuning")
```

Let's examine the model performance after tuning:

```{r check-performance, eval=FALSE}
# Check feeding levels
plotFeedingLevel(params) + 
  ggtitle("Feeding levels after tuning")

# Check predation mortality
plotPredMort(params) + 
  facet_wrap(~Species) +
  ggtitle("Predation mortality by species")
```

# Step 5: Switch to Competitive Refuge Method

Once we have a stable steady state with the tuning refuge profile, we can 
switch to the full competitive refuge method, which includes 
density-dependent refuge dynamics.

**Why switch methods?** The competitive method provides more realistic 
refuge dynamics but requires a stable baseline to work from.

```{r switch-refuge-method, eval=FALSE}
# Switch to competitive refuge method
params <- newRefuge(params,
                    new_method = "competitive", 
                    new_method_params = karpata_refuge)

# Check the new refuge profile
plotRefugeProfile(params) +
  ggtitle("Refuge profile with competitive method")
```

After switching refuge methods, we need to re-match biomasses as the refuge 
dynamics will have changed:

```{r rematch-after-refuge, eval=FALSE}
# Re-match biomasses with new refuge method
params <- params |>
    matchBiomasses() |> reefSteady() |> 
    matchBiomasses() |> reefSteady() |>
    matchBiomasses() |> reefSteady() |>
    matchBiomasses() |> reefSteady()

# Check vulnerability with new method
plotVulnerable(params) + 
  ggtitle("Vulnerability to predation with competitive refuge")

# Final biomass check
plotBiomassVsSpecies(params) + 
  ggtitle("Biomass match with competitive refuge method")
```

# Step 6: Fine-tune Reproduction Parameters

The final step involves tuning reproduction parameters to achieve the 
desired reproduction levels and density-dependence.

```{r tune-reproduction, eval=FALSE}
# Set reproduction efficiency to allow some density dependence
params <- setBevertonHolt(params, erepro = 0.35)
params <- reefSteady(params)

# Check reproduction levels - should be realistic for each species
final_reproduction_levels <- getReproductionLevel(params)
kable(data.frame(Species = names(final_reproduction_levels), 
                 Reproduction_Level = final_reproduction_levels),
      caption = "Final reproduction levels",
      digits = 3)

# Check ratio of density-independent to density-dependent reproduction
rdi_rdd_ratio <- getRDI(params) / getRDD(params)
kable(data.frame(Species = names(rdi_rdd_ratio), 
                 RDI_RDD_Ratio = rdi_rdd_ratio),
      caption = paste0("Ratio of density-independent to ",
                       "density-dependent reproduction"),
      digits = 2)
```

## Final Model Validation

Let's examine the final model performance across multiple metrics:

```{r final-validation, eval=FALSE}
# Overall spectra
plotSpectra(params, total = TRUE, power = 2) + 
  ggtitle("Final total spectra")

# Diet composition
plotDiet(params) + 
  scale_x_log10(limits = c(1, 1e4)) +
  ggtitle("Diet composition by size")

# Growth curves
plotGrowthCurves(params) + 
  ggtitle("Growth curves")
```

# Summary

This recipe provides a systematic approach to tuning `MizerReef` models:

1. **Start simple**: Use a binned refuge method for initial tuning
2. **Set intermediate reproduction**: Add moderate density-dependence (0.5) for stability
3. **Iterative matching**: Repeatedly match biomasses and growth rates
4. **Add complexity**: Switch to competitive refuge method when stable
5. **Final tuning**: Adjust reproduction parameters for realistic dynamics

## Key Functions Used

- `newReefParams()`: Create initial parameters object
- `setBevertonHolt()`: Set reproduction parameters
- `reefSteady()`: Project to steady state
- `calibrateReefBiomass()`: Match observed biomasses
- `matchReefGrowth()`: Match growth rates  
- `newRefuge()`: Switch refuge methods
- Various plotting functions for validation

## Complete Tuning Script

Here's the complete script used for the Karpata model:

```{r complete-script, eval=FALSE}
# Complete tuning script for Karpata model
library(mizerReef)
library(mizer)

# Load data
data("caribbean_10_species")
data("caribbean_10_interaction") 
data("tuning_profile")
data("karpata_refuge")

# 1. Create initial parameters
params <- newReefParams(species_params = caribbean_10_species,
                        interaction = caribbean_10_interaction,
                        method = "binned",
                        method_params = tuning_profile)

# 2. Set intermediate reproduction level (moderate density-dependence)
rdi <- rep(0.5, nrow(caribbean_10_species))
params <- setBevertonHolt(params, reproduction_level = rdi)

# 3. Initial steady state
params <- params |>
    reefSteady() |> reefSteady() |> reefSteady() |> 
    reefSteady() |> reefSteady() |> reefSteady()

# 4. Calibrate biomasses and growth
params <- calibrateReefBiomass(params)
params <- matchBiomasses(params)
params <- reefSteady(params)
params <- matchReefGrowth(params)
params <- reefSteady(params)

# Iterative refinement
params <- params |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady() |>
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |> 
    reefSteady()

# 5. Switch to competitive refuge
params <- newRefuge(params,
                    new_method = "competitive",
                    new_method_params = karpata_refuge)

# Re-match with new refuge
params <- params |>
    matchBiomasses() |> reefSteady() |> 
    matchBiomasses() |> reefSteady() |>
    matchBiomasses() |> reefSteady() |>
    matchBiomasses() |> reefSteady()

# 6. Fine-tune reproduction
params <- setBevertonHolt(params, erepro = 0.35)
params <- setBevertonHolt(params, reproduction_level = rdi)
params <- params |>
    reefSteady() |> reefSteady() |> reefSteady()

# Save final model
caribbean_10_model <- reefSteady(params)
```

## Further Reading

- [Getting started with MizerReef](mizerReef.html)
- [Karpata Model Description](karpata_model-description.html)
- [Mizer steady state tuning blog](https://blog.mizer.sizespectrum.org/posts/
  2021-08-20-a-5-step-recipe-for-tuning-the-model-steady-state/)

```{r session-info, echo=FALSE}
sessionInfo()
```
