---
title: "Tuning Diet Composition"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tuning Diet Composition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(mizer)
library(mizerExperimental)
library(mizerReef)
library(dplyr)
```

# Overview

A mizerReef model's diet composition -- how much of each consumer's intake
comes from other fish, from plankton, from algae, or from detritus -- is set
by several parameters working together, and it interacts with the biomass
and growth calibration covered in `vignette("steady-state-recipe")`. This
vignette focuses on that interaction specifically: where diet composition
actually lives in a mizerReef model, how it's kept in balance as you change
it, and how to read `plotDiet()` to tell whether a diet is realistic.

It assumes you already have a model near steady state (see
`vignette("mizerReef")` and `vignette("steady-state-recipe")`); the worked
example below uses the bundled `caribbean_3_model`.

```{r load-model}
data("caribbean_3_model")
params <- caribbean_3_model
```

# Where diet composition lives

A consumer's diet in mizerReef is split across up to four food sources: other
fish (governed by the interaction matrix $\theta_{ij}$), plankton, algae, and
detritus. The first three of these follow the same mizer/mizerReef pattern --
a preference column between 0 and 1 -- but they're set through two different
mechanisms, because algae and detritus are mizerReef's own invention and
plankton is not:

- **`interaction_resource`** (plankton) is a plain **mizer** species
  parameter, part of the same machinery mizer uses for any background
  resource. Set it the ordinary mizer way -- as a column in your species
  table, or with `species_params(params)$interaction_resource <- value` --
  there's no mizerReef-specific setter for it, because none is needed.
- **`interaction_algae`** and **`interaction_detritus`** are mizerReef's own
  columns, since mizer has no built-in notion of either resource. Set them
  through `setAlgaeParams()`/`setDetritusParams()`'s `UR_interaction`
  argument (which validates the values and can set several resources'
  interaction columns at once), or as species parameter columns directly, the
  same as `interaction_resource`.

That's the asymmetry in full: `setAlgaeParams()`/`setDetritusParams()` (and
`setURcapacity()` for the optional carrying-capacity variant) exist to
configure resources mizer doesn't know about. They aren't a gap relative to
`interaction_resource` -- mizer's own species-parameter machinery already
covers that one.

Two more parameters set the *power* of algae/detritus consumption, as
opposed to the *preference* for it: `rho_algae` and `rho_detritus`
(mass-specific consumption-rate coefficients). A species with
`interaction_algae > 0` but `rho_algae = 0` still encounters no algae in
practice -- `rho` and `interaction` multiply together in the encounter rate
(see `vignette("model-description")`'s Encounter rate section).

```{r diet-params-table}
params@species_params %>%
    select(species, interaction_algae, rho_algae,
           interaction_detritus, rho_detritus) %>%
    knitr::kable()
```

## Defaults `setAlgaeParams()`/`setDetritusParams()` choose for you

Besides `interaction_algae`/`interaction_detritus` defaulting to 0 (shown
above), `setAlgaeParams()`/`setDetritusParams()` fill in several
model-level defaults -- not tied to any one species -- whenever left unset.
All are mizerReef's own, and all are ordinary arguments you can override:

```{r ur-defaults-table, echo=FALSE}
knitr::kable(
  data.frame(
    Argument = c("algae_growth_initial", "algae_capacity",
                 "detritus_capacity", "sen_decomp", "ext_decomp",
                 "use_UR_cc"),
    Default = c("2000 g/m²/year", "1", "1", "0.8", "0.2", "FALSE"),
    Description = c(
      "Algae's fixed production rate -- see the note below on why this one is worth checking against your own system.",
      "Algae carrying capacity, only used if use_UR_cc = TRUE.",
      "Detritus carrying capacity, only used if use_UR_cc = TRUE.",
      "Proportion of senescence-mortality biomass that decomposes to detritus.",
      "Proportion of residual-natural-mortality biomass that decomposes to detritus.",
      "Whether algae/detritus use the logistic, carrying-capacity-limited dynamics instead of the default unlimited linear ones."
    )
  ),
  caption = "mizerReef's default algae/detritus settings."
)
```

`algae_growth_initial` is the one most worth checking rather than accepting
by default: it's held fixed throughout tuning (see below), so it acts as an
absolute anchor for the whole algae pool's scale, not just a starting guess
like most of the others here. It (like `sen_decomp`/`ext_decomp`) comes from
literature on *coral-reef algal turf and detritus* specifically (see
`?setAlgaeParams`) -- for a non-reef system, or even a reef with a different
dominant primary producer, look up your own system's value rather than
keeping this one.

Both bundled example models (`caribbean_3_model`, `caribbean_10_model`) use
these defaults as-is for everything except `ext_decomp` (set to 0.8, not the
0.2 default -- see `vignette("caribbean_3_model-description")` for the
literature source).

# Keeping algae and detritus in balance: `tuneUR()`

Changing a diet-composition parameter changes how much algae or detritus is
consumed, which knocks those two pools off their own steady state -- so a
diet change needs to be followed by re-tuning them, not just the fish.
`reefSteady()` does this automatically (it calls `tuneUR()`, or `tuneUR_cc()`
for carrying-capacity models, once the fish sub-model has converged), but
it's worth understanding exactly what each pool holds fixed, because the two
are **not** symmetric:

- **Algae's *production*** (`algae_growth_initial` in `setAlgaeParams()`,
  called "growth" in the API but genuinely a production rate, in
  g/m²/year) **is a fixed, literature-informed constant that is never
  retuned to match consumption.** `tuneUR()` instead solves for algae's
  *biomass*: at steady state, production $P_A$ and consumption $c_A B_A$
  balance, so $B_A = P_A / c_A$. Turn up grazing pressure (more consumers,
  or a higher `interaction_algae`/`rho_algae`) and the modelled algae
  biomass goes *down* to match -- production itself never moves.
- **Detritus is tuned the opposite way.** `tuneUR()` leaves detritus's
  *current* biomass alone and instead solves for the *external-input* term
  of its production -- the flux needed, on top of faeces and decomposing
  carcasses, to keep the current biomass a steady state under current
  consumption. So it's detritus's *production*, not its biomass, that
  chases whatever consumption you set up -- the reverse of algae.

Both of these are a **relative-scale** balance, done automatically as part
of ordinary steady-state tuning. They say nothing about whether the
resulting *absolute* biomass is realistic -- that's a separate, optional
final step (`rescale_algae()`, `detritus_lifetime()<-`) covered in
`vignette("steady-state-recipe")`'s last step, deliberately not duplicated
here since it's independent of diet composition: it holds total consumption
fixed while it corrects the scale.

# Diet composition and growth: iterate, don't set once

Changing a diet parameter changes realised encounter rates, which changes
growth -- so a diet change usually needs another pass of
`matchReefGrowth()`/`calibrateReefBiomass()` afterwards, exactly as in the
main calibration recipe. Treat diet tuning as one more lever inside that
same iterative loop, not a step you do once at the end:

```{r diet-tuning-loop, eval=FALSE}
# After adjusting interaction_algae/interaction_detritus/rho_algae/rho_detritus:
params <- params |>
    reefSteady() |>                 # re-balances algae/detritus (tuneUR())
    calibrateReefBiomass() |> matchBiomasses() |> matchReefGrowth() |>
    reefSteady()
```

# Reading `plotDiet()`

`plotDiet()` shows, for each predator and size, the proportion of intake
coming from each prey/resource category. A few things to look for:

```{r plot-diet, fig.cap="Diet composition by predator and size for caribbean_3_model."}
plotDiet(params)
```

- **A category near 0% or 100% across the whole size range** for a species
  that should have a mixed diet usually means its `interaction_*`/`rho_*`
  values for that resource are at their defaults (typically 0) rather than
  set deliberately -- check the table above rather than assuming the model
  has "decided" something ecologically.
- **An ontogenetic shift that doesn't match expectations** (e.g. a species
  that should transition from plankton to algae/detritus with size but
  doesn't) usually means the allometric exponents on encounter
  ($m_{alg}$/$m_{det}$, or the predation kernel for fish prey) need
  adjusting, not just the preference values.
- **Compare against `plotFeedingLevel()`** alongside diet: a species pinned
  near a feeding level of 1 (fully satiated) can have a diet share that
  looks reasonable in proportion while still indicating the *absolute* food
  availability is off -- diet composition and feeding level are different
  questions, and a model can get one right while the other is still wrong.

```{r plot-feeding-level, fig.cap="Feeding level by species and size for caribbean_3_model."}
plotFeedingLevel(params)
```

# Worked example: shifting a herbivore's diet

Suppose `herbivores` in `caribbean_3_model` should draw relatively more of
their diet from detritus and less from algae than the current calibration
gives them. Adjust `interaction_algae`/`interaction_detritus`, then re-tune:

```{r diet-shift-example}
sp <- species_params(params)
sp["herbivores", "interaction_algae"]
sp["herbivores", "interaction_detritus"]

params_shifted <- params
species_params(params_shifted)["herbivores", "interaction_algae"] <- 0.5
species_params(params_shifted)["herbivores", "interaction_detritus"] <- 0.5
params_shifted <- reefSteady(params_shifted)
```

```{r plot-diet-comparison, fig.cap="Herbivore diet before (top) and after (bottom) shifting interaction_algae/interaction_detritus.", fig.height=8}
patchwork::wrap_plots(
    plotDiet(params, species = "herbivores") + ggplot2::ggtitle("Original"),
    plotDiet(params_shifted, species = "herbivores") + ggplot2::ggtitle("Shifted"),
    ncol = 1
)
```

Note that `reefSteady()` alone only re-balances algae and detritus around
the *same* fish abundances -- it doesn't re-match biomass or growth to
observations. A real diet retuning would follow this with the
`calibrateReefBiomass()`/`matchBiomasses()`/`matchReefGrowth()` loop shown
above before treating the result as a finished, calibrated model.

# See also

- `vignette("steady-state-recipe")` for the full calibration recipe this
  fits into, including the final absolute-scale rescaling step.
- `vignette("model-description")`'s [Encounter
  rate](model-description.html#encounter-rate) and [unstructured resource
  dynamics](model-description.html#unstructured-resource-dynamics) sections
  for the underlying formulas.
- `?setAlgaeParams`, `?setDetritusParams`, `?tuneUR`, `?matchReefGrowth`,
  `?plotDiet`

```{r session-info, echo = FALSE}
sessionInfo()
```
