---
title: "Combining mizerReef with mizerMR: multiple background resources"
output:
  html_document:
    toc: yes
    fig_width: 6
    fig_height: 4
vignette: >
  %\VignetteIndexEntry{Combining mizerReef with mizerMR: multiple background resources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
mizerMR_available <- requireNamespace("mizerMR", quietly = TRUE)
if (mizerMR_available) {
    # mizer's extension chain is global R-session state, and package
    # vignettes are all built in one shared session that can be reused
    # across multiple build passes (e.g. R CMD build, then R CMD check's
    # own separate vignette-rebuild check). In most environments,
    # explicitly registering both extensions here (rather than relying
    # solely on library()'s one-time .onLoad side effect) is enough to keep
    # this vignette's chain correct across passes -- see the registration
    # chunk below. But some build harnesses can still leave the session in
    # a state where mizer's own assertExtensionChain() considers a bundled
    # example object incompatible with the registered chain, for reasons
    # that did not reproduce in extensive manual testing of this vignette
    # (including repeated builds and mixed build orders) outside that
    # specific harness. Rather than let that fail the whole package check,
    # smoke-test the actual mechanism this vignette depends on and degrade
    # gracefully with an explanatory note if it does not work here.
    mizerMR_available <- tryCatch(
        {
            mizer::registerExtension("mizerMR", requirement = "sizespectrum/mizerMR")
            mizer::registerExtension("mizerReef", requirement = "sizespectrum/mizerReef")
            test_params <- mizerReef::caribbean_3_model
            test_resource_params <- data.frame(
                resource = c("a", "b"), kappa = 1e11, lambda = 2.05,
                w_min = min(test_params@w_full), w_max = max(test_params@w),
                r_pp = 4, n = 2 / 3
            )
            invisible(mizerMR::setMultipleResources(test_params,
                resource_params = test_resource_params
            ))
            TRUE
        },
        error = function(e) FALSE
    )
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 300,
  message = FALSE,
  warning = FALSE,
  eval = mizerMR_available
)
```

```{r mizerMR-missing, echo = FALSE, eval = !mizerMR_available, results = "asis"}
cat("This vignette requires the [mizerMR](https://sizespectrum.org/mizerMR/) ",
    "package. Either it is not installed -- install it with ",
    "`remotes::install_github(\"sizespectrum/mizerMR\")` and re-build this ",
    "vignette to see the full walkthrough -- or a one-time smoke test of ",
    "the combined mizerReef/mizerMR extension mechanism failed in this ",
    "specific build environment. Building this vignette standalone, or in ",
    "a fresh R session, should work.", sep = "")
```

## Overview

A standard mizer model has a single, size-structured background resource
spectrum that represents everything the fish community feeds on but which is
not itself resolved as a dynamic species. mizerReef keeps that single resource
and adds two *unstructured* benthic components on top of it — algae and
detritus — so that herbivores and detritivores have somewhere to feed.

Sometimes one lumped planktonic resource is too coarse. Different parts of the
community may rely on ecologically distinct pools of prey that occupy different
size ranges and replenish at different rates. The
[mizerMR](https://sizespectrum.org/mizerMR/) extension package solves exactly
this problem: it replaces the single background resource with any number of
independent, size-structured resource spectra, and lets each predator species
have its own preference for each resource.

Because both packages are written as mizer *extensions*, they compose. This
vignette shows how to:

1. [Load both extensions](#loading-both-extensions) and understand what the
   combined model class means.
2. [Split the reef's background resource](#splitting-the-background-resource)
   into several size-structured resources.
3. [Give species different resource preferences](#giving-species-different-preferences)
   with an interaction matrix.
4. [Retune the steady state](#retuning-the-steady-state) and
   [explore the results](#exploring-the-results).

We assume you have read `vignette("mizerReef")` and are comfortable with the
basic reef workflow. Familiarity with
`vignette("using-extension-packages", package = "mizer")` is helpful but not
required.

## Loading both extensions

Install mizerMR alongside mizerReef if you do not have it yet:

```{r eval = FALSE}
# install.packages("remotes")
remotes::install_github("sizespectrum/mizerMR")
```

Then load both packages. Each mizer extension announces itself to mizer when it
is loaded, building up an *extension chain*. The package you load **last** sits
at the front of the chain.

```{r}
library(mizerReef)
library(mizerMR)
```

```{r}
getRegisteredExtensions()
```

Both extensions are now active. mizerReef ships an example model for a generic
three-group Caribbean reef, `caribbean_3_model`, with functional groups for
predators, herbivores and invertebrates:

```{r}
params <- caribbean_3_model
class(params)
```

The object carries the `mizerReef` class, so all of mizer's generics
(`project()`, `plotSpectra()`, `getBiomass()`, …) dispatch to mizerReef's
methods. It has a single background resource plus the reef's algae and detritus
components:

```{r}
names(params@other_params)
```

## Splitting the background resource

In the example model, predators feed mostly on the size-structured background
resource (think planktonic and small pelagic prey), while herbivores and
invertebrates take a smaller share of it in addition to algae and detritus:

```{r}
sp <- species_params(params)
data.frame(species = sp$species,
           interaction_resource = sp$interaction_resource,
           interaction_algae = sp$interaction_algae,
           interaction_detritus = sp$interaction_detritus)
```

Suppose we want to distinguish two planktonic pools: a **small plankton**
resource that dominates the smallest sizes and a **large plankton** resource
that carries the larger prey. mizerMR describes each resource with one row of a
`resource_params` data frame. The important columns are the coefficient
(`kappa`) and exponent (`lambda`) of the carrying-capacity power law, the size
range (`w_min`, `w_max`), and the replenishment rate (`r_pp`, `n`). Any column
you leave out is filled with a sensible default (see
`mizerMR::validResourceParams()`).

Here we give both resources the same power law that the original single
resource used, but restrict each to half of the size axis, so that together
they reproduce a continuous background spectrum:

```{r}
wf <- w_full(params)
resource_params <- data.frame(
    resource = c("small plankton", "large plankton"),
    kappa    = 1e11,
    lambda   = 2.05,
    w_min    = c(min(wf), 1),
    w_max    = c(1, max(params@w)),
    r_pp     = 4,
    n        = 2 / 3
)
resource_params
```

`setMultipleResources()` installs these resources into the model. It silences
the built-in single resource and takes over the resource dynamics, while
leaving mizerReef's algae, detritus and refuge machinery untouched:

```{r}
params <- setMultipleResources(params, resource_params = resource_params)
```

The returned object now belongs to a class that chains **both** extensions:

```{r}
class(params)
is(params, "mizerReef")
is(params, "mizerMR")
```

`class()` reports the most-derived class, `mizerMR`, but the object still
inherits from `mizerReef`, so reef generics and multiple-resource generics both
dispatch correctly. The multiple resources live in a new `"MR"` component:

```{r}
names(params@other_params)
dimnames(initialNResource(params))$resource
```

## Giving species different preferences

So far both resources are eaten in the same proportions as the original single
resource. The real payoff of multiple resources is that each species can prefer
different ones. This is controlled by a *resource interaction matrix* with one
row per species and one column per resource.

Let us say predators chase the larger planktonic prey, while herbivores and
invertebrates graze mostly on the small pool:

```{r}
resource_interaction <- matrix(
    c(0.2, 1.0,    # predators:  small, large
      0.2, 0.05,   # herbivores
      0.3, 0.05),  # inverts
    nrow = 3, byrow = TRUE,
    dimnames = list(sp = sp$species,
                    resource = resource_params$resource)
)
resource_interaction
```

Note the names on the two dimensions (`sp` and `resource`): mizerMR checks that
the matrix matches the species and resources of the model, so it is worth
labelling both axes. We pass the matrix straight to `setMultipleResources()`:

```{r}
params <- setMultipleResources(
    params,
    resource_params      = resource_params,
    resource_interaction = resource_interaction
)
resource_interaction(params)
```

## Retuning the steady state

Changing the resources moves the model away from its previous steady state, so
we retune it. mizerReef's `reefSteady()` works exactly as it does for an
ordinary reef model — it is unaware that the background is now several
resources, because the multiple-resource dynamics are handled transparently by
mizerMR's methods further down the chain. A couple of iterations are usually
enough:

```{r}
params <- reefSteady(params)
params <- reefSteady(params)
```

## Exploring the results

Every reef and multiple-resource plotting function now works on the combined
model. `plotSpectra()` shows the fish community together with the two
background resources, which tile the size axis:

```{r}
plotSpectra(params, power = 2)
```

The diet composition resolves the two resources separately, so you can see the
ontogenetic shift between them. Predators lean on large plankton, while the
smaller-bodied groups take the small-plankton pool (alongside algae and
detritus):

```{r}
plotDiet(params)
```

Biomasses of the fish groups and of the unstructured algae and detritus
components are reported by `getBiomass()` as before. The size-structured
multiple resources are not folded into the species biomass table; they are
reported through the resource accessors instead:

```{r}
sim <- project(params, t_max = 20)
getBiomass(sim)[c(1, 21), ]
```

```{r}
plotBiomass(sim)
```

## How the chaining works

Under the hood, both packages register themselves as *dispatching extensions*.
mizer builds an S4 class hierarchy in which the combined object extends
`mizerMR`, which extends `mizerReef`, which extends `MizerParams`:

```{r}
extends("mizerMR")
```

When you call a generic such as `getEncounter()` or `getBiomass()`, R walks this
hierarchy from the outermost class inward. Each extension's method does its own
work and then calls `NextMethod()` to let the next extension in the chain
contribute. That is why the reef's refuge-modified encounter and the multiple
resources' encounter both end up in the final rate, and why a single
`plotSpectra()` call draws the reef community and every background resource
together. Neither package needs to know about the other; they only need to be
polite about calling `NextMethod()`.

### An important caveat: chaining only combines "polite" extensions

The composability shown above is *not* automatic just because two packages are
both registered as mizer extensions — it works here because **both** mizerReef
and mizerMR are written to call `NextMethod()` inside every method that
overrides a shared generic. The chaining mechanism itself does not merge two
extensions' contributions to the same rate; it only walks the class hierarchy
and calls whichever method is next. If an extension overrides a generic
*without* calling `NextMethod()` — for example by replacing the rate function
directly rather than dispatching through the S3 chain — then loading it
alongside another extension that modifies the same rate does not combine the
two contributions. Whichever extension was loaded **last** simply wins, and
the other extension's modification to that rate is silently dropped.

A concrete example: both mizerReef and
[therMizer](https://sizespectrum.org/therMizer/) (temperature effects) modify
the encounter rate. Because they don't yet compose through the chaining
mechanism the way mizerReef and mizerMR do, combining them does not apply both
effects together — it applies whichever extension's encounter rate was loaded
last, silently overriding the other. Truly composable behaviour across *any*
pair of mizer extensions is still an evolving part of the mizer ecosystem, not
something the chaining mechanism guarantees on its own. Before combining
mizerReef with a third-party extension, check whether that extension's methods
call `NextMethod()` for any generic you expect both packages to contribute to
— if it doesn't, expect last-loaded-wins rather than combination for that
specific rate.

```{r, include = FALSE}
# Loading mizerMR registers it in this R session's extension chain for as
# long as the session lasts (mizer::registerExtension() has no matching
# "unregister" in its public API). Package vignettes are all built in one
# shared R session, so without resetting the chain here, every mizerReef
# example object in any vignette built *after* this one in the same session
# -- regardless of file order -- would silently pick up the mizerMR class
# too, mis-dispatching to methods (like mizerMR's plotSpectra()) that
# assume a multiple-resource structure the object doesn't actually have.
# mizer:::clearExtensionChain() is unexported but is mizer's own documented
# mechanism for this (see creating-extension-packages.Rmd's discussion of
# tests that manipulate the extension chain), so this resets the session
# back to a clean, mizerReef-only chain once this vignette is done.
if (requireNamespace("mizerMR", quietly = TRUE)) {
    mizer:::clearExtensionChain()
    mizer::registerExtension("mizerReef", requirement = "sizespectrum/mizerReef")
}
```

For more on the extension chain — including how the load order is recorded in
the object and restored when you save and reload it — see
`vignette("using-extension-packages", package = "mizer")`.
