---
title: "Simple Trait-Based Example"
subtitle: "Bonaire Steady State Model Description"
bibliography: references_sorted.bib
csl: ecology.csl
format:
    html:
        toc: true
        toc-depth: 4
        code-fold: true
        code-link: true
        pandoc-args: ["--syntax-highlighting=pygments"]
    pdf:
        toc: true
        toc-depth: 3
        number-sections: true
        number-depth: 2
editor:
    markdown:
        wrap: 72
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(dpi=400)
```

```{r, include = FALSE}
library(ggplot2)
library(dplyr)
library(knitr)
library(knitcitations)
library(mizer)
library(mizerExperimental)
library(mizerReef)

params <- bonaire_model
gear_params(params)["inverts, knife_edge_gear", "catchability"] <- 0
params@other_params$include_sen_mort <- TRUE

```


\newpage

# Model Description

The model includes `r nrow(params@species_params)-1` functional groups of fish as well as a general group of benthic invertebrates. Estimates of observed abundances for each group were to used select reproduction and resource consumption parameters so that the steady state abundances in the model agree with these observations. The process of estbilishing a steady state that agrees with empirical observations is nontrivial. Steady state parameters were tuned following the procedure described on the 
[mizer blog](https://blog.mizer.sizespectrum.org/posts/2021-08-20-a-5-step-recipe-for-tuning-the-model-steady-state/). The R script used to tune the steady state parameters is included at the end of this file. 

Biomass estimates were based on data collected from relatively un-fished reefs within a marine reserve in Bonaire. This data was part of the FORCE dataset that categorized fish community structure and habitat complexity across Caribbean coral reefs [@Dryden2016; @Rogers2014; @Newman2015; @Williams2015; @Williams2016]. Visual surveys recorded fish up to 5 cm in length. The cutoff weight for observed biomass estimates was calculated using published length-weight relationships [@FishBase]. The cutoff size is given in @tbl-biomass along with the observed biomass per 
square meter in grams.

```{r, echo = FALSE}
#| label: tbl-biomass
#| tbl-cap: Observed biomasses
params@species_params %>%
    filter(!is.na(biomass_observed)) %>%
    select(species,biomass_observed, biomass_cutoff) %>%
    kable(row.names = FALSE,
          col.names = c("Species Group", 
                        "Observed Biomass (g/$m^2$)", "Cutoff Size (g)"))
```

# Size Spectrum Dynamics

These dynamics expand on those used in standard \texttt{Mizer} models, see @mizer. 

## Growth

The growth of organisms is dependent on the energy they are able to obtain from consumed food resources.

### Predator-prey encounter rate  
  
The rate $E_{i}(w)$ at which a predator of species $i$ and weight $w$ encounters food is dictated by the predation power $\gamma_i(w)$, interaction matrix $\theta_{ij}$, the vulnerability to predation $V_{ij}(w_p)$, and the size selectivity of predator, given by the predation kernel $\phi_i(w,w_p)$.

#### Interaction Matrix  
  
The $\theta_{ij}$ matrix sets the interaction strength between predator group $i$ prey group $j$. The predator/prey interaction matrix has entries between 0 (if the groups can not interact) and 1, see @fig-interaction. 

```{r, echo = FALSE}
#| label: fig-interaction
#| fig-cap: "group interaction matrix"
p <- ggplot(melt(params@interaction),
       aes(x = prey, y = reorder(predator, desc(predator)), fill = value)) + 
    geom_tile(colour = "black", linewidth = 0.25) +
    scale_fill_gradient(low = "white", high = "blue", limit = c(0, 1),
                        guide = "legend") +
    labs(y = "Predator", x = "Prey", fill = "Interaction") + 
    scale_x_discrete(labels = c("Predators","Herbivores", "Inverts")) +
    scale_y_discrete(labels = c("Inverts","Herbivores", "Predators")) + 
    theme_minimal() + 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

#### Predation Kernel   
  
The parameters for the predation kernels were based on the global average predator prey mass ratio for marine organisms and home range size estimates from @Nash2014. All estimates fall within observed ranges. All groups use a lognormal predation kernel with the same parameters as in @Rogers2014. The parameters are given in @tbl-lognormal. 

```{r, echo = FALSE, results='asis'}
#| label: tbl-lognormal
#| tbl-cap: Parameters for the lognormal predation kernels
params@species_params %>%
    filter(pred_kernel_type == "lognormal") %>%
    select(beta, sigma, gamma) %>%
    kable()
```

### Vulnerability to Predation   
```{r, include = FALSE}
refuge_params <- params@other_params$refuge_params
```

The refuge function, $R_j(w_p)$, describes the proportion of fish of size $w_p$ in prey group $j$ that are hidden from predators. $1 - R_j(w_p)$ is then the proportion of fish of weight $w_p$ and prey group $j$ that are vulnerable to consumption by predators.  This model uses the `r print(params@other_params$refuge_params$method)` method to determine the **refuge profile**, the set of proportions that describe refuge availability across the entire size range of model fish (@Rogers2018). The proportion of prey of weight $w_p$ and group $j$ with access to refuge $R_j(w_p)$ is given by

$$
R_j(w_p) = min\left \{R_{max} \, , \frac{\tau\cdot\eta_k}{\sum_{i}\int_{w_{k-1}}^{w_k} N_i(w)~dw}
            ~~~~~~~ w_p \in (~w_{k-1}, w_k~] \right \}
$$ {#eq-refuge_data}

The parameter $\tau$ is the proportion of fish with access to refuge that are expected to utilize it, $\eta_k$ is the density ($no./m^{2}$) of refuges in size range $(w_{k-1}, w_k]$ and $\sum_{i} \int_{w_{k-1}}^{w_k} N_i(w)~dw$ gives the total density ($no./m^{2}$) of fish from any group in size range $(w_{k-1}, w_k]$. This represents the density of competitors for refuges in size class $k$. With this method, the refuge profile is density dependent. 

All fish are assumed to utilize refuge ($\tau = 1$). Fish smaller than `r paste(refuge_params$w_settle,"g")` are assume to be larval reef fish that have not yet settled to the reef. A maximum proportion of `r paste(100*refuge_params$max_protect,"%")` of fish are protected at any given time. 

The following table gives the fish length bins and refuge densities that define the refuge profile used in the steady state. Length bins and refuge densities were garnered from the FORCE data in the same location as biomass estimates.

```{r, echo = FALSE}
#| label: tbl-Refuge
#| tbl-cap: Parameters for defining the refuge profile.
#| 
params@other_params$method_params %>%
    mutate(refuge_density = round(refuge_density,3)) %>%
    kable(row.names = FALSE,
          col.names = c("Start of Bin (cm)",
                        "End of Bin (cm)",
                        "Refuge Density (no./m^2)"))
```

Each species group can utilise benthic structures differently depending on their specific traits. @tbl-refugeUse indicates how each species group interacts with structures that provide predation refuge. 

```{r, echo = FALSE, results='asis'}
#| label: tbl-refugeUse
#| tbl-cap: Parameters for defining how species interact with structure.
#| 
params@species_params %>%
    select(refuge_user, bad_pred) %>%
    mutate(refuge_user  = case_when(refuge_user == TRUE ~ 'Yes',
                                    refuge_user == FALSE ~ '×')) %>%
    mutate(bad_pred = case_when(bad_pred == TRUE ~ '×',
                                bad_pred == FALSE ~ 'Yes')) %>%
    kable(col.names = c("Group", "Uses refuge?", "Accesses prey in refuge?"))
```

@fig-refuge shows the density-dependent refuge profile produced by the competitive method for the simple trait based Bonaire model at steady state.

```{r, echo = FALSE, fig.height = 3, fig.width = 7}
#| label: fig-refuge
#| fig-cap: "Refuge profile used to find steady state abundances."
q <- plotRefuge(params)
    
```

\newpage

### Consumption  
  
Consumption of the detrital resource is subject to a Holling functional response type II to represent satiation. This relationship is defined by the maximum intake rate, which increases allometrically with body size at rate $n$. The parameter $h$ is the max consumption rate for an invertebrate consumer of size 1 gram. Values for $h$ were chosen so that invertebrates are neither too starved nor totally satiated.

Only a proportion $\alpha_i$ of consumed biomass is retained, while a proportion $1-\alpha_i$ is expelled in the form of feces, which contribute to the detritus. See @tbl-h for the the max consumption rate for invertebrates.

No maximum consumption rate is imposed for herbivores or predators.

```{r, echo = FALSE}
#| label: tbl-h
#| tbl-cap: Consumption parameters
params@species_params %>%
    filter(interaction_detritus > 0) %>%
    select(h, alpha, n) %>%
    kable()
```

#### Metabolic Losses  
  
The energy losses to metabolic needs are comprised of two components. Standard metabolism occurs at rate $k_{s.i}$, scaling allometrically with body size at rate $p$. The units of the coefficients $k_{s.i}$ are $\text{grams}^{1-p}$ per year. Losses due to activity and movement occur at rate $k_i$ in grams per year, scaling with body size at rate $1$.

```{r, results='asis', echo = FALSE}
#| label: tbl-metab
#| tbl-cap: Metabolism parameters
params@species_params %>%
    select(ks, p, k) %>%
    kable()
```

#### Energy Invested into Reproduction  
  
A proportion $\psi_i(w)$ of the energy available for growth andreproduction is used for reproduction. This proportion changes from zero below the weight $w_{mat.i}$ of maturation to one at the maximum weight $w_{max.i}$, where all available energy is used for reproduction.

Maturation length and age data were based on the most commonly observed species from each functional group in the FORCE data set [@Dryden2016; @Rogers2014; @Newman2015; @Williams2015; @Williams2016]. For predators, this was the Graysby grouper \textit{Cephalopholis cruentata}. The most commonly observed herbivore was the stoplight parrotfish, \textit{Sparisoma viridae}. Relevant maturity parameters were pulled from FishBase [@FishBase].

```{r, echo = FALSE}
#| label: tbl-repro
#| tbl-cap: Parameters determining the investment into reproduction.
params@species_params %>%
    select(w_mat, w_max, age_mat) %>%
    kable()
```

#### Somatic growth  
  
What is left over after metabolism and reproduction is taken into
account is invested in somatic growth. Thus the growth rate of an
individual of functional group $i$ and weight $w$ is $$
  g_i(w) = E_{r.i}(w)\left(1-\psi_i(w)\right).
$$ {#eq-growth} When food supply does not cover the requirements of
metabolism and activity, growth and reproduction stops, i.e. there is no
negative growth. There is no starvation mortality in mizerReef.

The values for the model parameters were chosen so that the resulting
growth curves would be close to von Bertalanffy growth curves. The
parameters in @tbl-kvb were taken from the literature.

```{r, echo = FALSE}
#| label: tbl-kvb
#| tbl-cap: Parameters for observed vonBertalanffy growth curves and length-weight relationships
params@species_params %>%
    select(k_vb, w_max, a, b) %>%
    kable()
```

Here the parameters $a$ and $b$ are parameters for the allometric
weight-length relationship $w = a l^b$ where $w$ is measured in grams
and $l$ is measured in centimetres.


## Mortality {#mortality}  
  
The mortality rate $\mu_i(w)$ of an individual of group $i$ and weight
$w$ has three sources: predation mortality $\mu_{p.i}(w)$, external
mortality $\mu_{ext.i}(w)$, fishing mortality $\mu_{f.i}(w)$. External
mortality is composed of residual natural mortality $\mu_{nat.i}(w)$,
and senescence $\mu_{sen.i}(w)$. The mortality rate for group $i$ is
then given by: $$
\mu_i(w)=\mu_{p.i}(w)+\mu_{ext.i}(w)+\mu_{f.i}(w)
$$ {#eq-mort}

### Predation mortality  
  
All consumption by fish translates into corresponding predation
mortalities on the ingested prey individuals. the rate at which all
predators from $j$ consume prey of size $w_p$ is $$
  \mathtt{pred\_rate}_j(w_p) = \int \phi_j(w,w_p) \gamma_j(w) N_j(w) \, dw.
$$ {#eq-pred_rate}

The mortality rate due to predation is then obtained as $$
  \mu_{p.i}(w_p) = \sum_j \mathtt{pred\_rate}_j(w_p)\, V_{ji}(w_p)\, \theta_{ji}.
$$ {#eq-mup}

### Fishing Mortality  
  
Like in `mizer`, fishing mortality in `mizerReef` is imposed  by fishing gears. The total per-capita fishing mortality (1/year) is obtained by summing over the mortality from all gears,

$$\mu_{f.i}(w) = \sum_g F_{g,i}(w)$$

where the fishing mortality $F_{g,i}(w)$ imposed by gear $g$ on group $i$ at size $w$ is calculated as:

$$F_{g,i}(w) = S_{g,i}(w) Q_{g,i} E_{g}$$

The constant $S$ is the selectivity by group, gear and size, $Q$ is the catchability by group and gear and $E$ is the fishing effort by gear.

Fishing parameters for mizerReef models can be set up with `setFishing()`, which contains the details of how to set up gears with different selectivities and
catchabilities for each group.

Fishing mortality $\mu_{f.i}(w)$ is calculated with the function `getFMort()`.
The vignettes were run with following fishing parameters (@tbl-fishing).

```{r, echo = FALSE}
#| label: tbl-fishing
#| tbl-cap: Selectivity parameters and catchability
params@gear_params %>%
    filter(catchability > 0) %>%
    select(species, knife_edge_size, catchability) %>%
    kable(row.names = FALSE,
          col.names = c("Group", "Minimum fishing size [g]", 
                        "Catchability [1/year]"))
```

### External Mortality   
  
```{r, results = 'asis', echo = FALSE}
ext_mort <- params@other_params[['ext_mort_params']]
```

#### Residual natural mortality  
  
Mortality caused by illness, fishing, or predators not explicitly
included in the model is captured by $\mu_{nat.i}(w)$, which is
independent of the functional groups and group abundances. It is assumed
to decrease allometrically with body size:

$$
\mu_{nat.i}(w) = \mu_{nat} w^{1 - n}
$$ {#eq-z0} where $\mu_{nat}$ is the residual natural mortality rate and
$n$ is the allometric scaling exponent. We use a residual mortality rate of 
$\mu_{nat} =$ `r ext_mort$nat_mort` per year at size 1 gram.

#### Senescence mortality   
  
Senescence mortality $\mu_{sen.i}(w)$ is intended to capture mortality
due to illness or old age. It is independent of group abundances.
Senescence mortality is assumed to increase allometrically with body
size. The rate of senescence mortality is given by:

$$
\mu_{sen.i}(w) =  k_{sen} \left(\frac{log_{10}(w)}
                                     {log_{10}(w_{max.i})}\right)^{p_{sen}}
$$ {#eq-extmort} where $k_{sen}$ is the rate of senescence mortality,
$p_{sen}$ defines the slope of the senescence curve, and $w_{max.i}$ is
the maximum body size of functional group $i$ in grams.

The senescence mortality rate for mortalities due to illness and old age is 
set to $k_{sen} =$ `r ext_mort$sen_prop` per year at size 1 gram and increases
allometrically with body size with slope $p_{sen} =$ `r ext_mort$sen_curve`.

## Reproduction {#reproduction}
The reproduction parameters $\epsilon_i$ and $R_{max.i}$ are not directly observable. The values were instead chosen so as to produce steady-state abundances of the groups that are in line with observations and to give reasonable values for the reproduction level.

@tbl-reproduction gives the steady-state reproduction level $e\_repro$ which is defined as the ratio between the actual reproduction rate $R_i$ and the maximal possible reproduction rate $R_{\max.i}$.

```{r, echo = FALSE}
#| label: tbl-reproduction
#| tbl-cap: Parameters determining reproduction
df <- params@species_params
#df$reproduction_level <- getReproductionLevel(params)
df %>%
    select(w_min, erepro, R_max) %>% # , reproduction_level) %>%
    kable()
```

# Resources {#resources}

The fish spectrum is fed by three background resources: plankton , algae, and detritus. Small individuals of all species will feed on plankton while only herbivorous groups and invertebrates feed on algae or detritus. The equations governing the flux of the algae and detritus resources were adapted from @mizerShelf [@deJuan2023].

### Plankton {#plankton}
```{r, echo = FALSE}
wcutoff <- params@resource_params$w_pp_cutoff
lambda <- params@resource_params$lambda
kappa <- params@resource_params$kappa
```

The plankton spectrum $N_P(w)$ tracks the abundance all planktonic food sources. This spectrum starts at a smaller size than the fish spectrum in order to provide food for the smallest individuals (larvae) of the fish spectrum. 

The plankton spectrum ranges from $w_0=`r signif(min(params@w_full), digits = 1)`$ to $w_{cutoff}=`r wcutoff`$ grams. The steady state abundance of plankton at size 1 gram is $\kappa=`r signif(kappa, digits = 3)` (g/m^{2})$. The slope of the plankton spectrum is set to $\lambda=`r signif(lambda, digits = 3)`$.

### Algae
The algal resource is described only by its total biomass $B_A$. Feeding on algae is not is not size-based. Herbivores can feed on algae of any size. In the steady state the total 
algal biomass per square meter is $B_A = `r signif(algae_biomass(params), digits = 3)`$ grams.

#### Algal consumption  
  
For each consumer species $i$, a parameter $\rho_{A.i}$ determines the rate at which individuals of that species encounter algal biomass. The parameters $\rho_{i.A}$ have units of $g^{-n}$ per year. They are non-zero only for species that consume at least some algae. The preference $\theta_{i.A}$ for algae is a value between 0 and 1 specifying the proportion of consumer diets that are comprised of algal matter.

```{r, echo = FALSE}
#| label: tbl-rho-a
#| tbl-cap: Parameters determining rates of algal consumption.
params@species_params %>%
    select(rho_algae, interaction_algae) %>%
    rename(rho = rho_algae) %>%
    filter(rho > 0) %>%
    kable()
```

#### Algal production  
  
```{r, echo = FALSE, include=FALSE}
pA <- params@other_params$algae$growth
```

The rate at which algal biomass is produced by the ecosystem is given by a constant growth rate with units of grams per unit area per unit time. At steady state algal production is `r signif(pA,digits = 4)` grams per square meter per year.


### Detritus
Detritus is consumed by herbivores and benthic invertebrates. Feeding on detritus is not size-based as fish can feed on detritus particles of any size. The detritus resource is described only by its total biomass $B_D$. In the steady state the total detrital biomass per square meter is $B_D = `r signif(detritus_biomass(params), digits = 4)`$ grams.

#### Detrital consumption  
  
For each consumer species $i$, a parameter $\rho_{D.i}$ determines the rate at which individuals of that species encounter detritus. The parameters $\rho_{i.D}$ have units of $g^{-n}$ per year. They are non-zero only for species that consume at least some detritus. The preference $\theta_{i.D}$ for detritus is a value between 0 and 1 specifying the proportion of consumer diets comprised of detritus.

```{r, echo = FALSE}
#| label: tbl-rho-d
#| tbl-cap: Parameters determining rates of detritus consumption.
params@species_params %>%
    select(rho_detritus,interaction_detritus) %>%
    rename(rho = rho_detritus) %>%
    filter(rho > 0) %>%
    kable()
```

#### Detrital production  
  
```{r, echo = FALSE}
#| include: false
e_d <- params@other_params$ext_decomp
s_d <- params@other_params$sen_decomp
r <- getDetritusProduction(params)
pf <- r[1]
pd <- r[2]
dext <- params@other_params$detritus$external
```
Detritus comes from defecation and decomposing dead organisms that die as a result of sources of external mortality. At steady state detritus from decomposing dead organisms is `r paste(round(pd,2))` g/year and defecation produces `r paste(round(pf,2))` g/year. A proportion `ext_decomp` of the external mortality and a proportion `sen_decomp` of senescence mortality decomposes to detritus.  The proportion of detritus that sinks from external mortality is `r paste(100*e_d,"%")` and the proportion that decomposes from senescence mortality is `r paste(100*s_d,"%")`. These values are based on estimates from @Hatcher1988.

External detritus is generated by unmodelled sources or sinks in from the pelagic zone. At steady state external detritus production is `r paste(round(dext,2))` g/year. This value is set so that steady state abundances match empirical observations. Where it is negative, feces and mortality are producing more detritus than can be consumed. Detrital biomass is assumed to be washed away by wave action in this case.

# Tuning the Steady State 
The following R script was used to tune the steady state parameters for this model.
```{r, echo = TRUE, eval = FALSE}
# Setting up a Caribbean coral reef mizer model with multiple resources
# Three groups: Predators, Herbivores, Invertebrates
# Model steady state calibration
# last tuned 21st February 2024

## Setup - load packages -------------------------------------------------------
library(mizer)
library(mizerExperimental)
library(mizerReef)
library(assertthat)
library(here)

## Load parameters -------------------------------------------------------------

# Load species parameter data
bonaire_species <- read.csv(here("inst/bonaire_species.csv"))
bonaire_int     <- read.csv(here("inst/bonaire_int.csv"),  row.names = 1)
bonaire_refuge  <- bonaire_refuge # Average refuge density per square meter 
tuning_profile  <- tuning_profile # 60% refuge for all size classes

# With these parameters, herbivores consume plankton at small sizes and 
#   transition fully to algae by maturity 
# With these parameters, invertebrates consume plankton and detritus,
#   with the proportion of detritus increasing with size

## Set model -------------------------------------------------------------------
    params <- newReefParams(group_params = bonaire_species,
                            interaction = bonaire_int,
                            method = "binned",
                            method_params = tuning_profile)

## Project to first steady state -----------------------------------------------
    params <- reefSteady(params)

## Calibrate biomasses and growth ----------------------------------------------

    # Match observed species group biomasses
    params <- calibrateReefBiomass(params)
    params <- matchBiomasses(params)
    params <- reefSteady(params)
    
    # Match observed growth rates
    params <- matchReefGrowth(params)
    params <- reefSteady(params)
    
    # Check for match with age at maturity
    age_mat_observed = bonaire_species$age_mat
    age_mat_model = age_mat(params)
    data.frame(age_mat_model, age_mat_observed)
    # Not bad
    
    # Check biomass match
    plotBiomassVsSpecies(params)
    # Biomasses way off from observations
    
    # Iterate to refine biomass - run this three times
    params <- params |>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()|>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()|>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()|>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()|>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()|>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()|>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()|>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()|>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()
    
    plotBiomassVsSpecies(params) # spot on

    # Check match with observed age at maturity
    age_mat_observed = bonaire_species$age_mat
    age_mat_model = age_mat(params)
    data.frame(age_mat_model, age_mat_observed)
    # Closer than needed
    
## Now switch to competitive method --------------------------------------------
    params <- newRefuge(params,
                        new_method = "competitive",
                        new_method_params = bonaire_refuge)
    
    # Match biomasses again - run twice
    params <- params |>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()|>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()|>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()|>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()|>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()|>
        calibrateReefBiomass() |> matchBiomasses()|> matchReefGrowth()|> 
        reefSteady()
    
    # Make sure new refuge is in place
    plotVulnerable(params)
    plotRefuge(params)
    # looks good
    
    plotBiomassVsSpecies(params) # spot on
    
    # Check match with observed age at maturity
    age_mat_observed = bonaire_species$age_mat
    age_mat_model = age_mat(params)
    data.frame(age_mat_model, age_mat_observed)
    # Also spot on

## Check resulting spectra and tune resources ----------------------------------
    
    # Spectra should be reasonably straight to match predictions of Sheldon's
    # spectrum but also have nonlinearities at refuge sizes
    plotSpectra(params, total = TRUE, power = 1) # looks straight, some bumps
    plotSpectra(params, total = TRUE, power = 2) # resource looks low
    
    # plot feeding level to check if resource is too low
    plotFeedingLevel(params, species = "inverts")
    # Invertebrate feeding level is stable throughout life - there is enough
    # resource to support fish, not too little or too much
    
# Tune reproduction ------------------------------------------------------------
    # We do not have yield or catch data - can't tune size distribution
    # First attempt to set very low to see what the minimum values are
    params <- setBevertonHolt(params, erepro = 0.0001)
    # Now set setting erepro same for all species, as low as possible
    params <- setBevertonHolt(params, erepro = 0.104)
    params <- reefSteady(params)
    # Check reproduction level (value between 0 and 1) - should be higher for
    # larger, slow growing species and low for small, fast growing ones
    rep <- getReproductionLevel(params)
    # These are very low for predators, should be higher and too high
    # for invertebrates. A reproduction level closer to one means reproduction 
    # rate is almost totally independent of the investment into reproduction
    # Reproduction should be somewhat density independent on reefs
    
    # Check comparison of density dependent & independent reduction
    getRDI(params) / getRDD(params)
    # preds, 1:1 - maybe too density dependent 
    # inverts, herbs 20+:1 more density independent - maybe too much
    
    # increase reproduction level to 0.5 for all
    rep_level <- c(0.5, 0.5, 0.5)
    names(rep_level) <- c("predators","herbivores","inverts")
    params <- setBevertonHolt(params,
                              reproduction_level = rep_level)
    
    # Iterate to get back to steady state
    params <- params |>
        reefSteady()|>
        reefSteady()|>
        reefSteady()
    
    # Check new reproduction - these look better
    rep <- getReproductionLevel(params)
    getRDI(params) / getRDD(params)
    # Now density independent reproduction is double for predators
    # And reproduction is somewhat density dependent for herbs and inverts
    
    # Check new spectra
    plotSpectra(params, total = TRUE, power = 1)
    plotSpectra(params, total = TRUE, power = 2)
    # These still look good

# Plots ------------------------------------------------------------------------
    plotBiomassVsSpecies(params)
    plotRefuge(params)
    plotSpectra(params, power = 1, total = TRUE)
    plotDiet(params)  
    plotGrowthCurves(params)
    plotPredMort(params)
    # Everything looks good here! I am happy with my results.
    
    # Save!
    bonaire_model <- reefSteady(params)

# Save in package --------------------------------------------------------------
    # Params object
    save(bonaire_model,   file = "data/bonaire_model.rda")
    
    # CSV Files
    save(bonaire_species,    file = "data/bonaire_species.rda")
    save(bonaire_int,        file = "data/bonaire_int.rda")
    
    # Things that dont change
    save(bonaire_refuge,  file = "data/bonaire_refuge.rda")
    save(tuning_profile,  file = "data/tuning_profile.rda")
```



