---
title: "Getting started with MizerReef"
author: ""
bibliography: references.bib
format:
  html:
    toc: true
    code-fold: true
    toc-depth: 3
vignette: >
  %\VignetteIndexEntry{Getting started with MizerReef}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

Setting up a `MizerReef` model is very similar to `mizer`. If you need to
brush up on the general `mizer` workflow, see
https://sizespectrum.org/mizer/articles/mizer.html.

This short guide shows how to set up a simple `MizerReef` model from the
beginning, highlighting the package-specific parameters you must supply and
showing three ways to define a predation refuge profile.

Note: this vignette is intended to be runnable in RStudio after the package
has been installed. The examples use the built-in `karpata_species` and
`karpata_int` example data included with the package.

## 1. Setting up species parameters

`MizerReef` builds on `mizer` species parameters. In addition to the
standard `mizer` columns, `MizerReef` requires (or expects) three
additional columns that control whether a group uses refuge and how it
interacts with prey inside refuge:

- `refuge_user` (logical): TRUE if members of the group can use predation
  refuge (i.e. hide inside habitat structure). 
  <span style="color: red; font-weight: bold;">Note: Should we make this size based?</span>
- `bad_pred` (logical): TRUE for groups that are *bad predators* with
  restricted ability to access prey inside refuge (for example eels that can
  extract prey from crevices are *good predators*). In `MizerReef` a value of `TRUE` for
  `bad_pred` means the group cannot access prey inside refuge.
  <span style="color: red; font-weight: bold;">Note: Should I change the name of this variable?</span>
- `satiation` (logical): TRUE if the group should be subject to a Holling
  type II satiation term when feeding on unstructured resources (algae or
  detritus). This mirrors the `MizerReef` design for invertebrate/herbivore
  feeding. Default is TRUE for herbivores and FALSE for carnivores.

In this vignette we'll use the `karpata_species` example dataset which
represents functional groups on a typical Indo-Pacific reef.

```{r setup, message=FALSE}
library(mizerReef)
data("karpata_species")
data("karpata_int")

# Show the species parameter rows and the three reef-specific columns
if(!all(c('refuge_user','bad_pred','satiation') %in% names(karpata_species))){
  karpata_species$refuge_user <- rep(TRUE, nrow(karpata_species))
  karpata_species$bad_pred   <- rep(FALSE, nrow(karpata_species))
  karpata_species$satiation  <- rep(TRUE, nrow(karpata_species))
}

knitr::kable(karpata_species[ , c('species','refuge_user','bad_pred','satiation')])
```

<span style="color:red"><strong>Note for supervisor:</strong> idea, should
we make `refuge_user` a size-based variable? It wouldn't be too
complicated — right now it is just a Boolean that gets multiplied; it could
be a size-based vector and do the same thing to capture ontogenetic change
in refuge use. Not necessarily required for current studies, just an option. Also, does bad_pred need a better name? </span>

## 2. Setting the refuge profile

`MizerReef` provides three main methods to define how refuge availability
varies with prey size:

- Sigmoidal: a smooth declining function controlled by a threshold length
  (L_refuge) and a maximum proportion protected. Good for data-poor reefs or
  when you want a simple, smooth profile.
- Binned: user-specified length bins with a constant protection proportion
  inside each bin. Good for theoretical experiments or when you have coarse
  bin information.
- Competitive (density-dependent): uses refuge density (no./m^2) for each
  length bin and divides refuges among competitors, so protection depends on
  fish density. Use this when you have empirical refuge density data.

Below we reproduce the Chapter 3 refuge-profile figure using the full
example code. This creates three example parameter sets (sigmoidal, binned,
competitive) for a small subset of species and plots the resulting refuge
profiles. The panels are labelled with the method that created them.

```{r refuge-method-bodyshape-examples, fig.height=4, fig.width=10, warning=FALSE, eval=TRUE, fig.alt='Three-panel plot showing refuge protection profiles by method (sigmoidal, binned, competitive) for example species: each panel plots protection proportion against length for multiple species, with legend identifying species.', fig.cap='Refuge protection profiles for sigmoidal, binned and competitive methods.'}

utils::data("body_shape_example_species_params", package = "mizerReef")
utils::data("body_shape_example_int", package = "mizerReef")

# Assign to the local variable names used in the example
sp <- body_shape_example_species_params
int <- body_shape_example_int

## Methods and method parameters used in Chapter 3
methods <- c("sigmoidal", "binned", "competitive")

## Sigmoidal example (one-line specification)
sig <- data.frame(L_refuge = 15, prop_protect = 0.2)

## Binned example: start_L, end_L (cm) with prop_protect per bin
bin <- data.frame(start_L = seq(0, 45, 5),
                  end_L   = seq(5, 50, 5),
                  prop_protect = c(1.0, 0.8, 0.6, 0.4, 0.4,
                                   0.4, 0.3, 0.3, 0, 0))

## Competitive example: refuge density (no./m^2) per bin
comp <- data.frame(start_L = seq(0, 45, 5),
                   end_L   = seq(5, 50, 5),
                   refuge_density = c(0, 10^5, 10^5, 0, 0,
                                      0, 10^6, 10^3, 0, 0))

## Minimal unstructured-resource interaction for the small example
UR_int <- data.frame(algae = rep(0, nrow(sp4)), detritus = rep(0, nrow(sp4)))

## Create params objects for each method
params_sig <- newReefParams(group_params = sp,
                            interaction = int,
                            UR_interaction = UR_int,
                            method = methods[1],
                            method_params = sig,
                            satiation = rep(TRUE, nrow(sp4)),
                            refuge_user = sp$refuge_user,
                            bad_pred = sp$bad_pred)
params_sig@other_params$degrade <- FALSE

params_bin <- newReefParams(group_params = sp4,
                            interaction = int4,
                            UR_interaction = UR_int,
                            method = methods[2],
                            method_params = bin,
                            satiation = rep(TRUE, nrow(sp4)),
                            refuge_user = sp4$refuge_user,
                            bad_pred = sp4$bad_pred)
params_bin@other_params$degrade <- FALSE

params_comp <- newReefParams(group_params = sp4,
                             interaction = int4,
                             UR_interaction = UR_int,
                             method = methods[3],
                             method_params = comp,
                             satiation = rep(TRUE, nrow(sp4)),
                             refuge_user = sp4$refuge_user,
                             bad_pred = sp4$bad_pred)
params_comp@other_params$degrade <- FALSE

## Plot with method labels (use plotRefuge and add titles)
g_sig <- plotRefuge(params_sig) + ggplot2::ggtitle('Method: sigmoidal')
g_bin <- plotRefuge(params_bin) + ggplot2::ggtitle('Method: binned')
g_comp <- plotRefuge(params_comp) + ggplot2::ggtitle('Method: competitive')

## Arrange and print
library(ggpubr)
ggpubr::ggarrange(g_sig, g_bin, g_comp, ncol = 3, common.legend = TRUE, legend = 'top')

## Optionally save the figure (example from Chapter 3)
# ggsave('bodyshape_methods_refuge_examples.png', plot = last_plot(), width = 27, height = 12, units = 'cm', dpi = 600)
```

The three panels show how the same set of functional groups can receive
different protection patterns depending on the chosen method and
method-specific parameters. The binned and competitive methods allow fine
control across size ranges; the sigmoidal method provides a smooth,
parsimonious shape controlled by the threshold length and max protection.

### Changing the refuge profile (examples)

Chapter 3 explores how altering the refuge profile changes model
outcomes. Here we show short code snippets that demonstrate the two common
ways you might change a refuge profile on an existing `params` object
without re-creating everything from scratch.

1) Change the sigmoidal threshold length `L_refuge` or `prop_protect`:

```{r change-sigmoidal, eval=FALSE}
# Update L_refuge and proportion protected for a sigmoidal params object
newRefuge(params_sig, new_refuge = TRUE, new_L_refuge = 20, new_prop_protect = 0.25)
# Recalculate refuge and (if needed) retune steady state: reefSteady(), matchReefGrowth(), etc.
```

2) Scale the binned or competitive bins by a constant (useful for scenario
testing, e.g. habitat loss reducing refuge by 50%):

```{r change-binned, eval=FALSE}
# Scale all bins by 0.5 (50% reduction in protection values)
newRefuge(params_bin, new_refuge = TRUE, scale_bin = 0.5)

# For competitive method, scaling refuge_density will change the availability
# of refuges and therefore the density-dependent allocation.
newRefuge(params_comp, new_refuge = TRUE, scale_bin = 0.5)
```

Notes:
- After changing refuge parameters, the model is typically no longer at the
  previous steady state. To restore steady-state agreement with observed
  biomasses you will usually re-run the tuning sequence (next section).
- Use `plotRefuge()` after `newRefuge()` to visualise the modified profile.

## 3. Unstructured resources

`MizerReef` adds two unstructured resources: `algae` and `detritus`. These
are tracked by total biomass (not as a size spectrum). Parameters and
important settings include:

- `interaction_algae`, `interaction_detritus` in `species_params`: how much
  each group feeds on these resources (0–1).
- `initial_algae_growth`: algal production (g/m^2/year). Set via
  `newReefParams(..., initial_algae_growth = value)` or left as default.
- `algae_capacity`, `detritus_capacity`: carrying capacities when using the
  carrying-capacity dynamics.

If you need to set or change these after creating a `params` object, use
`setURParams()` or pass them into `newReefParams()` when you create the
model. See `?setURParams` for details.

## 4. Creating a model with `newReefParams()`

Putting everything together: below is a compact example that creates a
`MizerParams` object for the `karpata` example using the sigmoidal refuge
method and then runs a short simulation.

```{r create-run, eval = FALSE}
# Example (eval = FALSE to avoid running a long simulation in CRAN checks)
params <- newReefParams(group_params = karpata_species,
                        interaction = karpata_int,
                        UR_interaction = data.frame(algae = karpata_species$interaction_algae,
                                                    detritus = karpata_species$interaction_detritus),
                        method = 'sigmoidal',
                        method_params = data.frame(L_refuge = 15, prop_protect = 0.2),
                        refuge_user = karpata_species$refuge_user,
                        bad_pred = karpata_species$bad_pred,
                        satiation = karpata_species$satiation)

# Run a short projection (adjust t_max and effort as needed)
sim <- project(params, t_max = 5, effort = 1)

# Explore results
plot(sim)
plotRefuge(params)
```

Tip: when switching between refuge methods or changing refuge parameters on
an existing `params` object, use `newRefuge()` to update the refuge
profile and then re-tune steady state components with functions such as
`matchReefGrowth()` and `reefSteady()` as described in the package
documentation.

## 5. Exploring results

Most plotting and summary functions from `mizer` also work with
`MizerReef`. The package adds helper plotting functions for refuge,
algae, and detritus (for example `plotRefuge()`, `plotAlgaeConsumption()`,
`plotDetritusProduction()`). Use `plot(sim)` for an overview and the
specialised helpers for more detail.

## Links and further reading

- The official `mizer` getting started guide: https://sizespectrum.org/mizer/articles/mizer.html
- `MizerReef` reference and articles: https://cmbeese.github.io/mizerReef/
- Functions referenced in this vignette: `newReefParams()`, `setRefuge()`,
  `getRefuge()`, `newRefuge()`, `setURParams()`.

## Tuning the steady state (overview and references)

Reaching a steady state that matches observed biomasses and growth rates is
an iterative process. The tuning approach used for MizerReef follows the
5-step recipe described in the mizer blog: 
https://blog.mizer.sizespectrum.org/posts/2021-08-20-a-5-step-recipe-for-tuning-the-model-steady-state/

The recipe in brief (see the blog for full details):

1. Start with plausible species parameters and a resource spectrum. Create
  an initial `params` object with `newReefParams()`.
2. Use `matchBiomasses()` (from mizer) to adjust reproduction parameters so
  steady-state biomasses match observed values up to the survey cutoff size.
3. Use `matchReefGrowth()` (package helper) to adjust growth-related
  parameters (e.g. `h`, `k_vb`) so growth rates resemble observations.
4. Tune unstructured-resource parameters (algae, detritus) so that resource
  production and consumer feeding levels are consistent; helper functions
  such as `tuneUR_cc()` or `setURcapacity()` can assist.
5. Run `reefSteady()` to find the steady state; check diagnostics and
  iterate steps 2–4 until the steady state is acceptable.

Important notes:
- Tuning is typically done with a noncomplex or a simple tuning refuge
  profile (for example `tuning_profile` or `step_tune` included in the
  package) before switching to the full competitive method.
- After switching to a density-dependent competitive refuge, remember to
  iterate matching and steady-state steps because refuge density changes
  feeding and competition.

## Model description context

For context, here is a short excerpt summarising the model purpose and the
role of habitat structure (extracted from the package model description):

"Habitat structure mediates system dynamics by providing predation refuge.
The most effective refuges fit prey while excluding predators, so refuge
use is primarily governed by body size. In systems with high structural
complexity (for example, coral reefs), refuge use patterns are reflected in
the size structure of fish assemblages. Because the protection offered by
refuges is size-dependent, size-spectrum models provide a natural
framework for exploring how benthic structure influences community
dynamics. MizerReef modifies predator–prey encounter rates to represent the
effects of habitat structure, allowing users to explicitly account for
changes in refuge availability caused by habitat degradation or
modification." 

See the full package model description vignette (`reef-model_description`)
for a complete explanation and references.

## Session info

```{r, echo = FALSE}
sessioninfo::session_info()
```
